### 2020-05-29 notebook/import.Rmd
work on store

# Import

```{r birds}
bench_time(
  neon_download("DP1.10003.001", file_regex = ".*basic.*[.]csv")
)
```

```{r beetles}
bench::bench_time(
  neon_download("DP1.10022.001")
)
```

```{r ticks}
bench::bench_time(
  neon_download("DP1.10093.001", quiet = TRUE)
)
```

```{r mammals}
bench::bench_time(
  neon_download("DP1.10072.001",  quiet = TRUE)
)
```

```{r zooplankton}
bench::bench_time(
  neon_download("DP1.20219.001",  quiet = TRUE)
)
```

```{r macroinvert}
bench::bench_time(neon_download("DP1.20120.001", quiet = TRUE))
```

```{r mozzies}
bench::bench_time(
  neon_download("DP1.10043.001", quiet = TRUE)
)
```

# Explore

```{r}
library(dplyr)

meta <- neon_index()
meta %>% count(product)
```

```{r}
meta %>% count(name)
```

```{r}

bench::bench_time({
f <- meta %>% filter(name == "brd_countdata", type == "basic") %>% pull(path)
df <- neon_read(f) # vroom with error handler
})

```

```{r}

bench::bench_time({
f <- meta %>% filter(name == "mam_perplotnight", type == "basic") %>% pull(path)
df <- neon_read(f) # vroom with error handler
})

```

```{r}
as_tibble(df)
```

# Distribute

```{r}
library(contentid)
library(fs)
library(dplyr)

```

```{r}
meta <- neon_index()
d <- fs::dir_info(neon_dir())
```

how much data?

```{r}
dim(d)[[1]]
d %>% summarise(sum(size))
```

```{r}
ids <- contentid::store(meta$path, dir = "/zpool/content-store")
```

### 2020-05-29 R/api.R

@@ -37,7 +37,10 @@ neon_data <- function(product, api = "https://data.neonscience.org/api/v0", .tok
   data_api <- unlist(available$availableDataUrls)
   
   ## Extract the file list from the data endpoint.  O(sites * months) calls
+  ## Consider optional progress indicator here!
   resp <- lapply(data_api, httr::GET, httr::add_headers("X-API-Token" = .token))
+  
+  
   
   ## Format the result as a data.frame
   data <- do.call(rbind,
@@ -86,7 +89,8 @@ neon_download <- function(product, dest = neon_dir(), file_regex = "[.]csv", api
   ## Filter duplicate files, e.g. have identical crc32 values
   unique_files <- take_first_match(files, "crc32")
   
-
+  ## Report total expected download size:
+  
   
   ## make sure destination exists
   dir.create(dest, showWarnings = FALSE, recursive = TRUE)

### 2020-05-29 R/store.R

@@ -36,7 +36,7 @@ neon_index <- function(dir = neon_dir()){
 
 
 
-
+## Consider using conditionally
 neon_read <- function(files){
   ## vroom can read in a list of files, but only if columns are consistent
   ## dplyr::bind_rows can bind and fill missing columns

### 2020-05-29 notebook/import.R

@@ -1,51 +1,0 @@
-
-#```{r birds}
-bench_time(
-  neon_download("DP1.10003.001", file_regex = ".*basic.*[.]csv")
-)
-#```
-
-
-
-#```{r  beetles}
-
-#beetle_data <- neon_data("DP1.10022.001")
-
-bench::bench_time(
-  neon_download("DP1.10022.001")
-)
-#```
-
-#```{r mozzies}
-bench::bench_time(
-  neon_download("DP1.10043.001")
-)
-#```
-
-
-#```{r ticks}
-bench::bench_time(
-  neon_download("DP1.10093.001", quiet = TRUE)
-)
-
-#```
-
-
-#```{r mammals}
-dir.create("mammals")
-
-bench::bench_time(
-  neon_download("DP1.10072.001", dest = "mammals", file_regex = "*.basic.*[.]zip",  quiet = TRUE)
-)
-
-
-
-
-
-library(dplyr); library(vroom)
-meta <- neon_index()
-meta %>% count(product, name)
-f <- meta %>% filter(name == "mam_perplotnight", type == "basic") %>% pull(path)
-
-neon_read(f) # vroom with error handler
-

### 2020-05-30 README.Rmd
wip

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# neonstore

<!-- badges: start -->
<!-- badges: end -->

The goal of neonstore is to ...

## Installation

You can install the released version of neonstore from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("neonstore")
```

And the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("cboettig/neonstore")
```
## Example

Download all CSV files in the bird survey data products.  This will omit downloads of any existing data files.

```{r example}
library(neonstore)
neon_download("DP1.10003.001", file_regex = "[.]csv")

```

Now, view your library of NEON products:

```{r }
meta <- neon_index()
```

### 2020-05-30 notebook/import.Rmd
wip

@@
 ```{r}
-ids <- contentid::store(meta$path, dir = "/zpool/content-store")
+ids <- contentid::store(meta$path, dir = "/home/content-store")
 ```
+
+```{r}
+registry <- vroom::vroom("content_id.tsv")
+
+meta <- dplyr::left_join(meta, registry) %>% as_tibble()
+vroom::vroom_write(meta, "neonstore.tsv")
+
+```
+
+```{r}
+## read from the content-store instead
+meta <- vroom::vroom("neonstore.tsv")
+metab <- meta %>%  mutate(store = retrieve(id, "/home/content-store"))
+```
+
+```{r}
+birds <- metab %>% filter(name == "brd_countdata", type == "basic") %>% pull(store) %>% neon_read()
+
+```
+
+```{r}
+metab <- metab %>% mutate(mirror = gsub("/home/content-store", "https://data.carlboettiger.info", store))
+
+request <- paste0("https://hash-archive.jetstream.carlboettiger.info/api/enqueue/", metab$mirror)
+request2 <- paste0("https://hash-archive.org/api/enqueue/", metab$mirror)
+x <- lapply(request2, httr::GET)
+
+#xml2::url_escape("https://data.carlboettiger.info/data/10/c9/10c9e39af827ecdba4255ded08ab7f8e2b6895ce58561236d1e26ff3fa097d20")
+```

### 2020-05-30 R/api.R

@@ -5,7 +5,8 @@
 neon_sites <- function(api = "https://data.neonscience.org/api/v0", 
                        .token = Sys.getenv("NEON_TOKEN")){
   
-  resp <- httr::GET(paste0(api, "/sites"), httr::add_headers("X-API-Token" = .token))
+  resp <- httr::GET(paste0(api, "/sites"), 
+                    httr::add_headers("X-API-Token" = .token))
   sites <- httr::content(resp, as="text")
   jsonlite::fromJSON(sites)[[1]]
  
@@ -14,7 +15,8 @@ neon_sites <- function(api = "https://data.neonscience.org/api/v0",
 neon_products <- function(api = "https://data.neonscience.org/api/v0", 
                        .token = Sys.getenv("NEON_TOKEN")){
   
-  resp <- httr::GET(paste0(api, "/products"), httr::add_headers("X-API-Token" = .token))
+  resp <- httr::GET(paste0(api, "/products"),
+                    httr::add_headers("X-API-Token" = .token))
   products <- httr::content(resp, as="text")
   jsonlite::fromJSON(products)[[1]]
   
@@ -25,11 +27,16 @@ neon_products <- function(api = "https://data.neonscience.org/api/v0",
 
 #' @importFrom httr GET content stop_for_status
 #' @importFrom jsonlite fromJSON
-neon_data <- function(product, api = "https://data.neonscience.org/api/v0", .token = Sys.getenv("NEON_TOKEN")){
+neon_data <- function(product, 
+                      api = "https://data.neonscience.org/api/v0", 
+                      .token = Sys.getenv("NEON_TOKEN")){
 
   ## A single API call to sites, includes product & month at each site    
   sites_df <- neon_sites(api)
   dataProducts <- do.call(rbind, sites_df$dataProducts)
+  
+  ## FIXME allow a filter by time -- requesting only current month
+  ## will speed things up.
   
   ## Consider all/only the sites including the requested product.
   ## The DataUrl column gives the API endpoint data/{ProductCode}/{SiteCode}{Month}
@@ -72,8 +79,14 @@ neon_dir <- neon_default_registry <- function(){
 #' product <- "DP1.10003.001"
 #' ".*basic.*[.]zip"
 
+
+#' 
 #' @export
-neon_download <- function(product, dest = neon_dir(), file_regex = "[.]csv", api = "https://data.neonscience.org/api/v0", quiet = FALSE){
+neon_download <- function(product, 
+                          dest = neon_dir(), 
[…]

### 2020-05-30 R/store.R

@@ -1,18 +1,18 @@
 
 
-neon_index <- function(dir = neon_dir()){
-  into <- c("site", "product", "name", "month", "type", "timestamp", "ext")    
+neon_index <- function(table = NULL, dir = neon_dir()){
+  into <- c("site", "product", "table", "month", "type", "timestamp", "ext")    
   
   site <- "(NEON\\.D\\d\\d\\.\\w{4})"                     # \\1
   product <- "(DP\\d\\.\\d{5}\\.\\d{3})"                  # \\2
-  name <- "(\\w+)"                                        # \\3 
+  table <- "(\\w+)"                                        # \\3 
   month <- "(:?\\d{4}-\\d{2})?\\.?"                       # \\5
   type <- "(:?basic|expanded)?\\.?"                       # \\6
   timestamp <- "(\\d{8}T\\d{6}Z)"                         # \\7
   stamp <- paste0("(:?", month, type, ")?", timestamp)    # \\4
   ext <- "(csv)"                                          # \\8
   
-  regex <- paste(site, product, name, stamp, ext, sep = "\\.")
+  regex <- paste(site, product, table, stamp, ext, sep = "\\.")
 
     
   files <- list.files(neon_dir())
@@ -29,15 +29,26 @@ neon_index <- function(dir = neon_dir()){
   meta_c <- as.data.frame(meta_b)
   
   meta_c$path <- file.path(dir, filenames)
+  
+  if(!is.null(table)){
+    meta_c[grepl(table, meta_c$table), ]
+  }
   meta_c
 
 }
 
 
-
+neon_tables <- function(dir = neon_dir()){
+  meta <- neon_index()
+  unique(meta$table)
+}
 
 ## Consider using conditionally
 neon_read <- function(files){
+  
+  ## allow files to be a data.frame, e.g. from neon_index()
+  if(is.data.frame(files)) files <- files$path
+  
   ## vroom can read in a list of files, but only if columns are consistent
   ## dplyr::bind_rows can bind and fill missing columns
   tryCatch(vroom::vroom(files),

### 2020-05-30 R/store.R
patch store

@@ -5,14 +5,14 @@ neon_index <- function(table = NULL, dir = neon_dir()){
   
   site <- "(NEON\\.D\\d\\d\\.\\w{4})"                     # \\1
   product <- "(DP\\d\\.\\d{5}\\.\\d{3})"                  # \\2
-  table <- "(\\w+)"                                        # \\3 
+  name <- "(\\w+)"                                        # \\3 
   month <- "(:?\\d{4}-\\d{2})?\\.?"                       # \\5
   type <- "(:?basic|expanded)?\\.?"                       # \\6
   timestamp <- "(\\d{8}T\\d{6}Z)"                         # \\7
   stamp <- paste0("(:?", month, type, ")?", timestamp)    # \\4
   ext <- "(csv)"                                          # \\8
   
-  regex <- paste(site, product, table, stamp, ext, sep = "\\.")
+  regex <- paste(site, product, name, stamp, ext, sep = "\\.")
 
     
   files <- list.files(neon_dir())
@@ -31,7 +31,8 @@ neon_index <- function(table = NULL, dir = neon_dir()){
   meta_c$path <- file.path(dir, filenames)
   
   if(!is.null(table)){
-    meta_c[grepl(table, meta_c$table), ]
+    meta_c <- meta_c[grepl(table, meta_c$table), ]
+    meta_c <- meta_c[!grepl("basic", meta_c$type), ]
   }
   meta_c
 

### 2020-05-30 README.Rmd
readme

@@
 
-The goal of neonstore is to ...
+The goal of neonstore is to provide access and persistent storage of NEON data tables
 
@@
 
-You can install the released version of neonstore from [CRAN](https://CRAN.R-project.org) with:
-
-``` r
-install.packages("neonstore")
-```
-
-And the development version from [GitHub](https://github.com/) with:
+Install the development version from [GitHub](https://github.com/) with:
 
@@
 
-```{r example}
+```{r example, cache=TRUE}
 library(neonstore)
@@
 
-Now, view your library of NEON products:
+Now, view your library of NEON products.  These will persist between sessions, 
+so you only need to download once, or to retrieve updates.
 
 ```{r }
-meta <- neon_index()
+neon_tables()
 ```
+See files associated with a given table,
+
+```{r output="hide"}
+meta <- neon_index("brd_countdata") 
+meta
+```
+
+```{r echo=FALSE}
+# pretty-print
+tibble::as_tibble(meta)
+```
+
+Read in all the component tables into a single data.frame
+
+```{r}
+brd_countdata <- neon_read(meta)
+```

### 2020-05-30 R/api.R

@@ -121,6 +121,8 @@ neon_download <- function(product,
 
 take_first_match <- function(df, col){
   
+  if(nrow(df) < 2) return(df)
+  
   uid <- unique(df[[col]])
   na <- df[1,]
   na[1,] <- NA

### 2020-05-30 R/store.R

@@ -1,5 +1,5 @@
 
-
+#' @export
 neon_index <- function(table = NULL, dir = neon_dir()){
   into <- c("site", "product", "table", "month", "type", "timestamp", "ext")    
   
@@ -38,13 +38,15 @@ neon_index <- function(table = NULL, dir = neon_dir()){
 
 }
 
-
+#' @export
 neon_tables <- function(dir = neon_dir()){
   meta <- neon_index()
   unique(meta$table)
 }
 
 ## Consider using conditionally
+
+#' @export
 neon_read <- function(files){
   
   ## allow files to be a data.frame, e.g. from neon_index()

### 2020-05-30 README.Rmd
tidy readme

@@
 ---
-output: github_document
+output: 
+  github_document:
+    df_print: tibble
 ---
@@
 )
+library(neonstore)
+
 ```
@@
 
-The goal of neonstore is to provide access and persistent storage of NEON data tables
+The goal of neonstore is to provide access and persistent storage of NEON data tables.
 
@@
 
-```{r output="hide"}
+```{r}
 meta <- neon_index("brd_countdata") 
@@
 
-```{r echo=FALSE}
-# pretty-print
-tibble::as_tibble(meta)
-```
-
 Read in all the component tables into a single data.frame
 
-```{r}
+```{r message = FALSE}
 brd_countdata <- neon_read(meta)
+brd_countdata
 ```

### 2020-06-01 R/api.R
ragged_zoom
avoids dpylr and faster too!

@@ -16,6 +16,8 @@ neon_sites <- function(api = "https://data.neonscience.org/api/v0",
  
 }
 
+
+
 #' Table of all NEON Data Products
 #'
 #' Return a table of all NEON Data Products, including product descriptions
@@ -81,7 +83,7 @@ neon_products <- function(
 
 #' product <- "DP1.10003.001"
 
-#' 
+#' @importFrom progress progress_bar
 #' @importFrom httr GET content stop_for_status
 #' @importFrom jsonlite fromJSON
 neon_data <- function(product, 
@@ -113,15 +115,19 @@ neon_data <- function(product,
     data_api <- data_api[dates <= end_date]
   }  
   
-    
+  if(length(data_api) == 0){
+    message("No additional files to download.")
+    return(invisible(NULL))
+  }
+
   ## Extract the file list from the data endpoint.  O(sites * months) calls
   pb <- progress::progress_bar$new(
-    format = "  querying API for available data [:bar] :percent eta: :eta",
+    format = "  querying API [:bar] :percent eta: :eta",
     total = length(data_api), 
     clear = FALSE, width= 60)
   
   resp <- lapply(data_api, function(x){
-    if(!quiet) pb$tick()$print()
+    if(!quiet){ pb$tick() }
     httr::GET(x, httr::add_headers("X-API-Token" = .token))
   })
   
@@ -187,8 +193,7 @@ neon_dir <- function(){
 #' \donttest{
 #'  
 #'  neon_download("DP1.10003.001", 
-#'                start_date = "2019-01-01", 
-#'                file_regex = ".*basic.*[.]zip")
+#'                start_date = "2019-01-01"
 #' 
 #' }
 neon_download <- function(product, 
@@ -225,25 +230,25 @@ neon_download <- function(product,
   dir.create(dir, showWarnings = FALSE, recursive = TRUE)
   
   ## now time to download!
-  
+  f
   pb <- progress::progress_bar$new(
     format = "  downloading [:bar] :percent eta: :eta",
-    total = length(data_api), 
+    total = length(unique_files$url), 
     clear = FALSE, width= 60)
   
   handle <- curl::new_handle()
   for(i in seq_along(unique_files$url)){
-    if(!quiet) pb$tick()$print()
+    if(!quiet) pb$tick()
     curl::curl_download(unique_files[i, "url"], 
                         unique_files[i, "dir"], 
[…]

### 2020-06-01 R/store.R

@@ -6,7 +6,7 @@
 #' @inheritParams neon_download
 #' 
 #' @export
-neon_store <- function(table = NULL, dir = neon_dir()){
+neon_index <- function(table = NULL, dir = neon_dir()){
   
   files <- list.files(neon_dir())
   
@@ -39,17 +39,17 @@ neon_store <- function(table = NULL, dir = neon_dir()){
   if(!is.null(table)){
     meta_c <- meta_c[grepl(table, meta_c$table), ]
   }
-  # Prefer 'extended' format if available
-  if(any(grepl("extended", meta_c$type))){
-    meta_c <- meta_c[grepl("extended", meta_c$type), ]
+  # Prefer 'expanded' format if available
+  if(any(grepl("expanded", meta_c$type))){
+    meta_c <- meta_c[grepl("expanded", meta_c$type), ]
   }
   
   meta_c
-
+  
 }
 
 #' @export
-neon_stored_tables <- function(dir = neon_dir()){
+neon_stored <- function(dir = neon_dir()){
   meta <- neon_index()
   unique(meta$table)
 }
@@ -57,26 +57,90 @@ neon_stored_tables <- function(dir = neon_dir()){
 
 ## Consider using conditionally
 
+#' read in neon tabular data
+#' 
+#' 
+#' NEON's tabular data files are separated out into separate .csv
+#' files for each site for each month of sampling.  In principle,
+#' each file has identical columns.  [vroom::vroom] can read in a
+#' data table that has been sharded into many files like this much
+#' much faster than other parsers can read in each table iteratively, 
+#' (and thus can greatly out-perform the 'stacking" methods in neonUtilities).
+#'
+#' Unfortunately, not all datasets are entirely consistent in their use
+#' of columns.  `neon_read` works around this by parsing such tables in
+#' groups of matching schema, which is still reasonably fast.
+#' 
+#' For convenience, neon_read takes the name of a table in the local store.
+#' 
+#' @param table the name of a downloaded NEON table
+#' @param ... additional arguments to [vroom::vroom], can usually be omitted.
+#' @param files optionally, specify a vector of file paths directly (e.g. as
+#' provided from [neon_index]) and specify `table` argument as NULL.
+#' @inheritParams neon_download
+#' 
 #' @export
-neon_read <- function(table, dir = neon_dir()){
+#' @examples 
+#' 
+#' neon_read("brd_count")
+#' @import vroom vroom spec
[…]

### 2020-06-01 notebook/import.Rmd
WIP :pen:

@@
 ```{r zooplankton}
-bench::bench_time(
-  neon_download("DP1.20219.001",  quiet = TRUE)
-)
+neon_download("DP1.20219.001")
 ```
@@
 ```{r macroinvert}
-bench::bench_time(neon_download("DP1.20120.001", quiet = TRUE))
+neon_download("DP1.20120.001")
 ```
@@
 bench::bench_time(
-  neon_download("DP1.10043.001", quiet = TRUE)
+  neon_download("DP1.10043.001")
 )

### 2020-06-01 R/api.R

@@ -199,14 +199,15 @@ neon_dir <- function(){
 neon_download <- function(product, 
                           start_date = NA,
                           end_date = NA,
-                          file_regex =  ".*basic.*[.]zip",
+                          type = "expanded",
+                          file_regex =  "[.]zip",
                           quiet = FALSE,
                           dir = neon_dir(), 
                           api = "https://data.neonscience.org/api/v0",
                           .token =  Sys.getenv("NEON_TOKEN")){
   
   ## Query the API for a list of all files associated with this data product.
-  data <- neon_data(product, 
+  files <- neon_data(product, 
                     start_date = start_date, 
                     end_date = end_date, 
                     api = api,
@@ -214,23 +215,27 @@ neon_download <- function(product,
   
   ## Omit those files we already have
   already_have <- list.files(dir)
-  new_data <- data[!(data$name %in% already_have), ]
+  files <- files[!(files$name %in% already_have), ]
   
   ## Filter for only files matching the file regex
-  files <- new_data[grepl(file_regex, new_data$name),]
+  files <- files[grepl(file_regex, files$name), ]
   files$dir <- file.path(dir, files$name)
   
-  ## Filter duplicate files, e.g. have identical crc32 values
+  
+  ## Filter to have only expanded or basic (not both)
+  if(type == "expanded")
+    files <- files[!grepl("basic", files$name), ]
+  if(type == "basic")
+    files <- files[!grepl("expanded", files$name), ]
+  
+  
+  ## Filter out duplicate files, e.g. have identical crc32 values
   unique_files <- take_first_match(files, "crc32")
-  
-  ## Report total expected download size:
-  
   
   ## make sure destination exists
   dir.create(dir, showWarnings = FALSE, recursive = TRUE)
   
   ## now time to download!
-  f
   pb <- progress::progress_bar$new(
     format = "  downloading [:bar] :percent eta: :eta",
     total = length(unique_files$url), 
@@ -254,6 +259,11 @@ neon_download <- function(product,
 }
 
 
+
+
+## helper method for filtering out duplicate tables
+## NEON API loves returning metadata files with identical content but 
+## different names associated with each site and sampling month.
 take_first_match <- function(df, col){
   
   if(nrow(df) < 2) return(df)

### 2020-06-01 R/store.R

@@ -1,25 +1,52 @@
+### Functions in store.R do not require API access, but interact only with
+### the local persistent storage
 
 
 #' Show information about all files downloaded to the local store
 #' 
-#' @param table a table name (or regex pattern) to filter on.
+#' 
+#' NEON products consist of several individual components, which are in turn
+#' broken up by site and sampling month. By storing these individual files,
+#' neonstore enables more reproducible workflows that can be traced back to
+#' original, unaltered input data.  These atomized files can be quickly and easily
+#' combined into unified tables, see [neon_read].
+#' 
+#' File names include metadata such as the file productCode,
+#' table name, site, and sampling month, as well as timestamp of creation.
+#' `neon_index()` parses this metadata from the file name string and returns
+#' the information in a convenient table, along with a path to each file.
+#' 
+#' @param product Include only files matching this NEON productCode (optional)
+#' @param table Include only files matching this table name (or regex pattern). 
+#' (optional).
 #' @inheritParams neon_download
 #' 
 #' @export
-neon_index <- function(table = NULL, dir = neon_dir()){
+#' @examples
+#' 
+#' \dontshow{
+#' # Hide setting tempfile, since a user would specify a persistent location
+#'  Sys.setenv("NEONSTORE_HOME"=tempfile())
+#' }
+#' 
+#' neon_index()
+#' 
+#' ## Just bird survey product
+#' neon_index("DP1.10003.001")
+#' 
+neon_index <- function(product = NULL, table = NULL, dir = neon_dir()){
   
-  files <- list.files(neon_dir())
-  
+  files <- list.files(dir)
   ## Parse metadata from NEON file names
   into <- c("site", "product", "table", "month", "type", "timestamp", "ext")    
   site <- "(NEON\\.D\\d\\d\\.\\w{4})\\."                     # \\1
-  product <- "(DP\\d\\.\\d{5}\\.\\d{3})\\."                  # \\2
+  productCode <- "(DP\\d\\.\\d{5}\\.\\d{3})\\."                  # \\2
   name <- "(:?\\w+)?\\.?"                                    # \\3 
   month <- "(:?\\d{4}-\\d{2})?\\.?"                          # \\4
   type <- "(:?basic|expanded)?\\.?"                          # \\5
[…]

### 2020-06-01 README.Rmd
readme update :pen:

@@
 
-The goal of neonstore is to provide access and persistent storage of NEON data tables.
+The goal of neonstore is to provide quick access and persistent storage of NEON data tables.
 
@@
 ```
-## Example
+## Quickstart
 
-Download all CSV files in the bird survey data products.  This will omit downloads of any existing data files.
+Discover data products of interest:
+
+```{r}
+products <- neon_products()
+
+i <- grepl("Populations", products$themes)
+products[i, c("productCode", "productName")]
+
+i <- grepl("bird", products$keywords)
+products[i, c("productCode", "productName")]
+```
+
+Download all data files in the bird survey data products.
 
@@
 library(neonstore)
-neon_download("DP1.10003.001", file_regex = "[.]csv")
+neon_download("DP1.10003.001")
 
@@
 
-Now, view your library of NEON products.  These will persist between sessions, 
+Now, view your store of NEON products.  These will persist between sessions,
 so you only need to download once, or to retrieve updates.
+(Note: the individual files making up a data product can all be listed by
+`neon_index()`).
 
 ```{r }
-neon_tables()
-```
-See files associated with a given table,
-
-```{r}
-meta <- neon_index("brd_countdata") 
-meta
+neon_store()
 ```
@@
 
-```{r message = FALSE}
-brd_countdata <- neon_read(meta)
-brd_countdata
+```{r}
+neon_read("brd_countdata")
 ```
+
+## Details
+
+`neon_download()` will omit downloads of any existing data files. 
+You can request multiple products at once using vector notation. 
+You can optionally include date filters, e.g. to request only records more recent
+than a certain date. 
+Doing so will preserve API quota and improve speed (see API limits, below).
+
+Because `neonstore` only stores raw data products as returned from the NEON API,
+this makes it easy to track provenance and compare results across other analyses
+of the NEON data.
+
+## Note on API limits
+
+[The NEON API now rate-limits requests.](https://data.neonscience.org/data-api/rate-limiting/#api-tokens).
+Using a personal token will increase the number of requests you can make.
+See that link for directions on registering for a token.
+Then pass this token in `.token` argument of `neon_download()`,
+or for frequent use, add this token as  an environmental variable, `NEON_DATA`
+to your local `.Renviron` file in your user's home directory.
+
+`neon_download()` must first query each the API of eacn NEON site which collects
+that product, for each month the product is collected.  
+(It would be much more efficient on the NEON server if the API could take
+queries of the from `/data/<product>/<site>, and pool the results, rather than
+require each month of sampling separately!)

### 2020-06-01 R/store.R
update imports

@@ -135,7 +135,7 @@ neon_store <- function(product = NULL, dir = neon_dir()){
 #' @param files optionally, specify a vector of file paths directly (e.g. as
 #' provided from [neon_index]) and specify `table` argument as NULL.
 #' @inheritParams neon_download
-#' @import vroom vroom spec
+#' @importFrom vroom vroom spec
 #' @export
 #' 
 #' @examples 

### 2020-06-01 R/api.R
update docs

@@ -81,11 +81,11 @@ neon_products <- function(
 }
 
 
-#' product <- "DP1.10003.001"
 
 #' @importFrom progress progress_bar
 #' @importFrom httr GET content stop_for_status
 #' @importFrom jsonlite fromJSON
+#' @noRd
 neon_data <- function(product, 
                       start_date = NA,
                       end_date = NA,
@@ -161,18 +161,19 @@ neon_dir <- function(){
 #}
 
 
-#' product <- "DP1.10003.001"
-#' ".*basic.*[.]zip"
-
 
 #' Download NEON data products into a local store
-#' @param product A NEON productCode. See [neon_download].
+#'
+#'
+#' @param product A NEON `productCode`. See [neon_download].
 #' @param start_date Download only files as recent as (YYYY-MM-DD). Leave
 #' as `NA` to download up to the most recent avaialble data.
 #' @param end_date Download only files up to end_date (YYYY-MM-DD). Leave as 
 #' `NA` to download all prior data.
+#' @param type Should we prefer the basic or expanded version of this product? 
+#' See details. 
 #' @param file_regex Download only files matching this pattern.  See details.
-#' @param quiet Should progress be displayed?
+#' @param quiet Should download progress be displayed?
 #' @param dir Location where files should be downloaded. By default will
 #' use the appropriate applications directory for your system 
 #' (see [rappdirs::user_data_dir]).  This default also be configured by
@@ -183,17 +184,57 @@ neon_dir <- function(){
 #' required but will allow access to a higher number of requests before
 #' rate limiting applies, see 
 #' <https://data.neonscience.org/data-api/rate-limiting/#api-tokens>.
-#' Note that once files are downloaded once, `neonstore` provides persitent
+#' Note that once files are downloaded once, `neonstore` provides persistent
 #' access to them without further interaction required with the API.
 #' 
-#' @details `"*basic*.zip`
+#' @details Each NEON data product consists of a collection of 
+#' objects (e.g. tables), which are in turn broken into individual files by 
+#' site and sampling month.  Additionally, many NEON products have been 
+#' expanded, including some additional columns. Consequently, users must
+#' specify if they want the "basic" or "expanded" version of this data. 
+#' 
+#' In the products table (see [neon_products]), the `productHasExpanded` 
+#' column indicates if the data
+#' product has expanded, and the columns `productHasBasicDescription` and
+#' `productHasExpandedDescription` provide a detailed explanation of the
[…]

### 2020-06-01 R/store.R

@@ -34,9 +34,15 @@
 #' ## Just bird survey product
 #' neon_index("DP1.10003.001")
 #' 
+#' \dontshow{
+#' # tidy
+#'  Sys.unsetenv("NEONSTORE_HOME")
+#' }
+#' 
 neon_index <- function(product = NULL, table = NULL, dir = neon_dir()){
   
   files <- list.files(dir)
+  
   ## Parse metadata from NEON file names
   into <- c("site", "product", "table", "month", "type", "timestamp", "ext")    
   site <- "(NEON\\.D\\d\\d\\.\\w{4})\\."                     # \\1
@@ -59,6 +65,7 @@ neon_index <- function(product = NULL, table = NULL, dir = neon_dir()){
   
   ## Format as tidy data.frame
   meta_b <- jsonlite::fromJSON(jsonlite::toJSON(meta))
+  if(length(meta_b) == 0) return(NULL)
   colnames(meta_b) <- into
   meta_c <- as.data.frame(meta_b)
   meta_c$path <- file.path(dir, filenames)
@@ -100,7 +107,7 @@ neon_index <- function(product = NULL, table = NULL, dir = neon_dir()){
 #' an internet connection or incur rate limiting on requests.
 #' 
 #' @seealso [neon_products], [neon_download], [neon_index]
-#' @inheritParams neon_store
+#' @inheritParams neon_index
 #' @export
 #' @examples
 #' 

### 2020-06-01 README.Rmd
GitHub actions
some unit tests :construction:

@@
 <!-- badges: start -->
+[![R build status](https://github.com/cboettig/neonstore/workflows/R-CMD-check/badge.svg)](https://github.com/cboettig/neonstore/actions)
+[![Codecov test coverage](https://codecov.io/gh/cboettig/neonstore/branch/master/graph/badge.svg)](https://codecov.io/gh/cboettig/neonstore?branch=master)
 <!-- badges: end -->

### 2020-06-01 R/api.R

@@ -89,6 +89,7 @@ neon_products <- function(
 neon_data <- function(product, 
                       start_date = NA,
                       end_date = NA,
+                      site = NA,
                       quiet = FALSE,
                       api = "https://data.neonscience.org/api/v0", 
                       .token = Sys.getenv("NEON_TOKEN")){
@@ -106,14 +107,23 @@ neon_data <- function(product,
   
   data_api <- unlist(available$availableDataUrls)
 
+  regex <- paste0(api, "/data/", product, "/(\\w+)/(\\d{4}-\\d{2})$")
   ## Filter by time -- year-month is included at end of data_api list
-  dates <- as.Date( gsub(".*(\\d{4}-\\d{2})$", "\\1-01", data_api) )
+  dates <- as.Date(gsub(regex, "\\2-01", data_api))
   if(!is.na(start_date)){
    data_api <- data_api[dates >= start_date]
   }
   if(!is.na(end_date)){
     data_api <- data_api[dates <= end_date]
   }  
+  
+  ## Filter by site
+  data_sites <- gsub(regex, "\\1", data_api)
+  if(!is.na(site)){
+    data_api <- data_api[data_sites %in% site]
+  }
+  
+  
   
   if(length(data_api) == 0){
     message("No additional files to download.")
@@ -170,6 +180,7 @@ neon_dir <- function(){
 #' as `NA` to download up to the most recent avaialble data.
 #' @param end_date Download only files up to end_date (YYYY-MM-DD). Leave as 
 #' `NA` to download all prior data.
+#' @param site 4-letter site code(s) to filter on. Leave as `NA` to search all.
 #' @param type Should we prefer the basic or expanded version of this product? 
 #' See details. 
 #' @param file_regex Download only files matching this pattern.  See details.
@@ -234,12 +245,14 @@ neon_dir <- function(){
 #'  ## Advanced use: filter for a particular table in the product
 #'  neon_download("DP1.10003.001",
 #'                start_date = "2019-01-01",
+#'                site = "YELL",
 #'                file_regex = ".*brd_countdata.*\\.csv")
 #' 
 #' }
 neon_download <- function(product, 
                           start_date = NA,
                           end_date = NA,
+                          site = NA,
                           type = "expanded",
                           file_regex =  "[.]zip",
                           quiet = FALSE,
@@ -250,7 +263,8 @@ neon_download <- function(product,
   ## Query the API for a list of all files associated with this data product.
   files <- neon_data(product, 
[…]

### 2020-06-01 tests/testthat.R

@@ -1,0 +1,6 @@
+library(testthat)
+library(neonstore)
+
+Sys.setenv("NEONSTORE_HOME" = tempfile())
+
+test_check("neonstore")

### 2020-06-01 tests/testthat/test-api.R

@@ -1,0 +1,81 @@
+context("api")
+
+
+test_that("neon_dir()", {
+  
+  x <- neon_dir()
+  expect_equal(x, Sys.getenv("NEONSTORE_HOME"))
+  
+})
+
+
+
+test_that("neon_sites()", {
+  
+  skip_on_cran()
+  skip_if_offline()
+  
+  x <- neon_sites()
+
+  expect_is(x, "data.frame")
+    
+})
+
+
+
+test_that("neon_products()", {
+
+  skip_on_cran()
+  skip_if_offline()
+    
+  x <- neon_products()
+  expect_is(x, "data.frame")
+  
+  
+})
+
+
+
+
+test_that("neon_data()", {
+  
+  skip_on_cran()
+  skip_if_offline()
+
+  x <- neon_data("DP1.10003.001",
+                 site = "YELL",
+                 start_date = "2019-06-01",
+                 end_date = "2019-08-01")
+  
+  expect_is(x, "data.frame")
+  
+})
+
+
+
+test_that("take_first_match()", {
+  
+  
+  df <- data.frame(A = c(1,1,2,2), 
+                   B = c("a", "b", "c", "d"),
+                   row.names = NULL)
+  out <- take_first_match(df, "A")
+  
+  expect_equal(dim(out), c(2,2))
+
+})
+
+
+test_that("neon_download()", {
+  
+  x <- neon_download("DP1.10003.001",
+                     site = "YELL",
+                     start_date = "2019-06-01",
+                     end_date = "2019-08-01")
+  expect_is(x, "data.frame")
+  
+  
+})
+
+
+

### 2020-06-01 tests/testthat/test-store.R

@@ -1,0 +1,35 @@
+context("store")
+
+
+## setup so we have something in the store
+x <- neon_download("DP1.10003.001",
+                   site = "YELL",
+                   start_date = "2019-06-01",
+                   end_date = "2019-08-01")
+
+
+test_that("neon_index()", {
+  
+  x <- neon_index()
+  expect_is(x, "data.frame")
+  expect_true(any(grepl("DP1.10003.001", x$product)))
+  
+})
+
+
+test_that("neon_store()", {
+  
+  x <- neon_store()
+  expect_true(any(grepl("brd_countdata", x)))
+  
+})
+
+
+test_that("neon_read()", {
+  
+  x <- neon_read("brd_count")
+  expect_is(x, "data.frame")
+  expect_true(any(grepl("observerDistance", colnames(x))))
+  
+  
+})

### 2020-06-01 R/api.R
fix duplicate field :P

@@ -41,7 +41,6 @@ neon_sites <- function(api = "https://data.neonscience.org/api/v0",
 neon_products <- function(
   fields = c("productCode", 
              "productName", 
-             "productStatus", 
              "productDescription", 
              "productStatus",
              "themes",

### 2020-06-01 R/api.R
add site to docs

@@ -174,28 +174,6 @@ neon_dir <- function(){
 #' Download NEON data products into a local store
 #'
 #'
-#' @param product A NEON `productCode`. See [neon_download].
-#' @param start_date Download only files as recent as (YYYY-MM-DD). Leave
-#' as `NA` to download up to the most recent avaialble data.
-#' @param end_date Download only files up to end_date (YYYY-MM-DD). Leave as 
-#' `NA` to download all prior data.
-#' @param site 4-letter site code(s) to filter on. Leave as `NA` to search all.
-#' @param type Should we prefer the basic or expanded version of this product? 
-#' See details. 
-#' @param file_regex Download only files matching this pattern.  See details.
-#' @param quiet Should download progress be displayed?
-#' @param dir Location where files should be downloaded. By default will
-#' use the appropriate applications directory for your system 
-#' (see [rappdirs::user_data_dir]).  This default also be configured by
-#' setting the environmental variable `NEONSTORE_HOME`, see [Sys.setenv] or
-#' [Renviron].
-#' @param api the URL to the NEON API, leave as default.
-#' @param .token an authentication token from NEON. A token is not
-#' required but will allow access to a higher number of requests before
-#' rate limiting applies, see 
-#' <https://data.neonscience.org/data-api/rate-limiting/#api-tokens>.
-#' Note that once files are downloaded once, `neonstore` provides persistent
-#' access to them without further interaction required with the API.
 #' 
 #' @details Each NEON data product consists of a collection of 
 #' objects (e.g. tables), which are in turn broken into individual files by 
@@ -234,6 +212,29 @@ neon_dir <- function(){
 #' These duplicates are also packaged within the `.zip` downloads where it
 #' is not possible to exclude them from the download. 
 #' 
+#' @param product A NEON `productCode`. See [neon_download].
+#' @param start_date Download only files as recent as (YYYY-MM-DD). Leave
+#' as `NA` to download up to the most recent avaialble data.
+#' @param end_date Download only files up to end_date (YYYY-MM-DD). Leave as 
+#' `NA` to download all prior data.
+#' @param site 4-letter site code(s) to filter on. Leave as `NA` to search all.
+#' @param type Should we prefer the basic or expanded version of this product? 
+#' See details. 
+#' @param file_regex Download only files matching this pattern.  See details.
+#' @param quiet Should download progress be displayed?
+#' @param dir Location where files should be downloaded. By default will
[…]

### 2020-06-02 R/api.R
fix download mode for windows machines

@@ -301,7 +301,8 @@ neon_download <- function(product,
     # consuder benchmarking if alternatives are faster? curl_download?
     download.file(unique_files[i, "url"], 
                   unique_files[i, "dir"],
-                  quiet = TRUE)
+                  quiet = TRUE,
+                  mode = "wb")
   }
 
   # unzip and remove .zips

### 2020-06-02 tests/testthat/test-store.R
a few more tests for store

@@ -16,14 +16,15 @@ test_that("neon_index()", {
   
 })
 
-
 test_that("neon_store()", {
   
   x <- neon_store()
   expect_true(any(grepl("brd_countdata", x)))
   
+  x <- neon_store(dir = tempfile())
+  expect_null(x)
+  
 })
-
 
 test_that("neon_read()", {
   
@@ -31,5 +32,24 @@ test_that("neon_read()", {
   expect_is(x, "data.frame")
   expect_true(any(grepl("observerDistance", colnames(x))))
   
+})
+
+test_that("vroom_ragged()", {
+  
+  A <- data.frame(A = 1:5, B = 1:5)
+  B <- data.frame(A = 1:5, B = 1:5, C = 1:5)
+  
+  files <- c(file.path(neon_dir(), "A.txt"),
+             file.path(neon_dir(), "B.txt"))
+  vroom::vroom_write(A, files[[1]])
+  vroom::vroom_write(B, files[[2]])
+  
+  out <- vroom_ragged(files)
+  
+  expect_is(out, "data.frame")
+  expect_equal(colnames(out), colnames(B))
+  
   
 })
+
+

### 2020-06-02 R/api.R
include type in tablename to avoid ambiguity

@@ -240,7 +240,7 @@ neon_dir <- function(){
 #' @examples 
 #' \donttest{
 #'  
-#'  neon_download("DP1.10003.001")
+#'  neon_download("DP1.10003.001", start_date = "2019-01-01", site = "YELL")
 #'                
 #'  ## Advanced use: filter for a particular table in the product
 #'  neon_download("DP1.10003.001",

### 2020-06-02 R/setup-tests.R

@@ -1,0 +1,1 @@
+Sys.setenv("NEONSTORE_HOME" = tempfile())

### 2020-06-02 R/store.R

@@ -44,18 +44,11 @@ neon_index <- function(product = NULL, table = NULL, dir = neon_dir()){
   files <- list.files(dir)
   
   ## Parse metadata from NEON file names
-  into <- c("site", "product", "table", "month", "type", "timestamp", "ext")    
-  site <- "(NEON\\.D\\d\\d\\.\\w{4})\\."                     # \\1
-  productCode <- "(DP\\d\\.\\d{5}\\.\\d{3})\\."                  # \\2
-  name <- "(:?\\w+)?\\.?"                                    # \\3 
-  month <- "(:?\\d{4}-\\d{2})?\\.?"                          # \\4
-  type <- "(:?basic|expanded)?\\.?"                          # \\5
-  timestamp <- "(:?\\d{8}T\\d{6}Z)?\\.?"                     # \\6
-  ext <- "(\\w+$)"                                           # \\7
-  regex <- paste0(site, productCode, name, month, type, timestamp, ext)
-  meta <- strsplit(gsub(regex, "\\1  \\2  \\3  \\4  \\5  \\6  \\7", files), "  ")
+  parsed <- gsub(neon_regex(), "\\1  \\2  \\3-\\5  \\4  \\5  \\6  \\7", files)
+  meta <- strsplit(parsed, "  ")
   
   ## Confirm parsing was successful
+  into <- c("site", "product", "table", "month", "type", "timestamp", "ext")
   parts <- vapply(meta, length, integer(1L))
   meta <- meta[parts == length(into)]
   
@@ -69,7 +62,7 @@ neon_index <- function(product = NULL, table = NULL, dir = neon_dir()){
   colnames(meta_b) <- into
   meta_c <- as.data.frame(meta_b)
   meta_c$path <- file.path(dir, filenames)
-  
+
   
   ## Apply any filters
   if(!is.null(table)){
@@ -80,18 +73,22 @@ neon_index <- function(product = NULL, table = NULL, dir = neon_dir()){
     meta_c <- meta_c[grepl(product, meta_c$product), ]
   }
   
-  
-  # Prefer 'expanded' format if available
-  if(any(grepl("basic", meta_c$type))){
-    meta_c <- meta_c[!grepl("basic", meta_c$type), ]
-  }
-  
   class(meta_c) <- c("tbl_df", "tbl", "data.frame")
   meta_c
   
 }
 
-
+neon_regex <- function(){
+  site <- "(NEON\\.D\\d\\d\\.\\w{4})\\."                     # \\1
[…]

### 2020-06-02 tests/testthat.R

@@ -1,6 +1,5 @@
 library(testthat)
 library(neonstore)
 
-Sys.setenv("NEONSTORE_HOME" = tempfile())
 
 test_check("neonstore")

### 2020-06-02 tests/testthat/test-store.R

@@ -28,7 +28,7 @@ test_that("neon_store()", {
 
 test_that("neon_read()", {
   
-  x <- neon_read("brd_count")
+  x <- neon_read("brd_countdata-expanded")
   expect_is(x, "data.frame")
   expect_true(any(grepl("observerDistance", colnames(x))))
   
@@ -48,6 +48,22 @@ test_that("vroom_ragged()", {
   
   expect_is(out, "data.frame")
   expect_equal(colnames(out), colnames(B))
+  
+  
+})
+
+test_that("neon_regex()", {
+  
+   x <- c(
+  "NEON.D01.BART.DP1.10003.001.brd_countdata.2015-06.expanded.20191107T154457Z.csv",
+  "NEON.D01.BART.DP0.10003.001.validation.20191107T152154Z.csv",
+  "NEON.D01.BART.DP1.10003.001.brd_countdata.2015-06.basic.20191107T154457Z.csv",
+  "NEON.D01.BART.DP1.10003.001.brd_references.expanded.20191107T152154Z.csv",
+  "NEON.D01.BART.DP1.10003.001.2019-06.basic.20191205T150213Z.zip",
+  "NEON.D01.HARV.DP1.10022.001.bet_sorting.2014-06.basic.20200504T173728Z.csv"
+   )
+  strsplit(gsub(neon_regex(), "\\1  \\2  \\3  \\5  \\6  \\7  \\8", x), "  ")
+  
   
   
 })

### 2020-06-02 tests/testthat/teardown-tests.R
teardown

@@ -1,0 +1,1 @@
+Sys.unsetenv("NEONSTORE_HOME")

### 2020-06-04 notebook/import.Rmd
update imports
support hash validation

@@
 
-```{r birds}
-bench_time(
-  neon_download("DP1.10003.001", file_regex = ".*basic.*[.]csv")
-)
-```
+# Checksum comparison
 
-```{r beetles}
-bench::bench_time(
-  neon_download("DP1.10022.001")
-)
-```
-
-```{r ticks}
-bench::bench_time(
-  neon_download("DP1.10093.001", quiet = TRUE)
-)
-```
-
-```{r mammals}
-bench::bench_time(
-  neon_download("DP1.10072.001",  quiet = TRUE)
-)
-```
-
-```{r zooplankton}
-neon_download("DP1.20219.001")
-```
-
-```{r macroinvert}
-neon_download("DP1.20120.001")
-```
-
-```{r mozzies}
-bench::bench_time(
-  neon_download("DP1.10043.001")
-)
-```
-
-# Explore
+The NEON API returns 32 character crc32 checksum
 
 ```{r}
-library(dplyr)
+raw_files <- neon_index(table = "brd_countdata-expanded")
+x <- raw_files$path[[1]]
+digest::digest(x, algo = "crc32", file = TRUE)
 
-meta <- neon_index()
-meta %>% count(product)
+url <- paste("https://data.neonscience.org/api/v0/data", 
+                raw_files$product[[1]], 
+                "BART", 
+                raw_files$month[[1]], 
+                sep = "/")
+resp <- httr::GET(url)
+out <- jsonlite::fromJSON(httr::content(resp, "text"))$data$files
+out[out$name == basename(x),]
+
+dir.create("tmp")
+paste0("tmp/", out$name)
+purrr::walk2(out$url, paste0("tmp/", out$name), curl::curl_download)
+unzip("tmp/NEON.D01.BART.DP1.10003.001.2015-06.expanded.20191107T154457Z.zip",
+      exdir = "tmp/expanded")
 ```
@@
 ```{r}
-meta %>% count(name)
-```
+vapply(fs::dir_ls("tmp", glob="*.csv"), digest::digest,  character(1L), algo = "crc32", file = TRUE, USE.NAMES = FALSE)
+vapply(fs::dir_ls("tmp/expanded", glob="*.csv"), digest::digest, character(1L), algo = "crc32", file = TRUE, USE.NAMES = FALSE)
 
-```{r}
+out$crc32[grepl("[.]csv",out$name)]
 
-bench::bench_time({
-f <- meta %>% filter(name == "brd_countdata", type == "basic") %>% pull(path)
-df <- neon_read(f) # vroom with error handler
-})
-
-```
-
-```{r}
-
-bench::bench_time({
-f <- meta %>% filter(name == "mam_perplotnight", type == "basic") %>% pull(path)
-df <- neon_read(f) # vroom with error handler
-})
-
-```
-
-```{r}
-as_tibble(df)
 ```

### 2020-06-04 R/api.R

@@ -6,14 +6,17 @@
 #' @inheritParams neon_download
 #' @importFrom httr GET content
 #' @importFrom jsonlite fromJSON
+#' @export
 neon_sites <- function(api = "https://data.neonscience.org/api/v0", 
                        .token = Sys.getenv("NEON_TOKEN")){
   
   resp <- httr::GET(paste0(api, "/sites"), 
                     httr::add_headers("X-API-Token" = .token))
-  sites <- httr::content(resp, as="text")
-  jsonlite::fromJSON(sites)[[1]]
- 
+  txt <- httr::content(resp, as="text")
+  sites <- jsonlite::fromJSON(txt)[[1]]
+  
+  class(sites) <- c("tbl_df", "tbl", "data.frame")
+  sites
 }
 
 
@@ -257,6 +260,7 @@ neon_download <- function(product,
                           type = "expanded",
                           file_regex =  "[.]zip",
                           quiet = FALSE,
+                          verify = TRUE,
                           dir = neon_dir(), 
                           api = "https://data.neonscience.org/api/v0",
                           .token =  Sys.getenv("NEON_TOKEN")){
@@ -275,7 +279,7 @@ neon_download <- function(product,
   
   ## Filter for only files matching the file regex
   files <- files[grepl(file_regex, files$name), ]
-  files$dir <- file.path(dir, files$name)
+  files$path <- file.path(dir, files$name)
   
   
   ## Filter to have only expanded or basic (not both)
@@ -299,11 +303,26 @@ neon_download <- function(product,
   
   for(i in seq_along(unique_files$url)){
     if(!quiet) pb$tick()
-    # consuder benchmarking if alternatives are faster? curl_download?
+    # consider benchmarking if alternatives are faster? curl_download?
     curl::curl_download(unique_files[i, "url"], 
-                        unique_files[i, "dir"])
+                        unique_files[i, "path"])
   }
 
+  if(verify) {
+    md5 <- vapply(unique_files$path, 
+           function(y) as.character(openssl::md5(file(y))),
+           character(1L), USE.NAMES = FALSE)
+    i <- which(md5 != unique_files$crc32)
+    if(length(i) > 0) {
+      
+      warning(paste("Removing downloaded files which", 
+                    "did not match the expected hash:",
+                    unique_files$path[i]), call. = FALSE)
+      unlink(unique_files$path[i])
+    }
+  }
+  
+  
   # unzip and remove .zips
[…]

### 2020-06-04 R/store.R

@@ -19,6 +19,8 @@
 #' @param product Include only files matching this NEON productCode (optional)
 #' @param table Include only files matching this table name (or regex pattern). 
 #' (optional).
+#' @param hash name of a hashing algorithm to check file integrity. Can be
+#'  md5, sha1, or sha256 currently; or set to [NULL] to skip hash computation.
 #' @inheritParams neon_download
 #' 
 #' @export
@@ -39,7 +41,10 @@
 #'  Sys.unsetenv("NEONSTORE_HOME")
 #' }
 #' 
-neon_index <- function(product = NULL, table = NULL, dir = neon_dir()){
+neon_index <- function(product = NULL, 
+                       table = NULL, 
+                       hash = "md5",
+                       dir = neon_dir()){
   
   files <- list.files(dir)
   
@@ -73,10 +78,43 @@ neon_index <- function(product = NULL, table = NULL, dir = neon_dir()){
     meta_c <- meta_c[grepl(product, meta_c$product), ]
   }
   
+  ## Compute hashes, if requested
+  meta_c$hash <- file_hash(meta_c$path, hash = hash)
+  
   class(meta_c) <- c("tbl_df", "tbl", "data.frame")
   meta_c
   
+  
+  
+  
 }
+
+#' @importFrom openssl md5 sha1 sha256
+file_hash <- function(x, hash = "md5"){
+  
+  
+  if(is.null(hash)) return(NULL)
+  
+  hash_fn <- switch(hash, 
+         "md5" = openssl::md5,
+         "sha1" = openssl::sha1,
+         "sha256" = openssl::sha256,
+         NULL)
+  if(is.null(hash_fn)) {
+   warning(paste("No function for", hash, "found", call. = FALSE))
+   return(NULL)
+  }
+  
+  ## httr imports openssl already
+  hashes <- paste0("hash://", hash, "/",
+    vapply(x, 
+           function(y) as.character(hash_fn(file(y))),
+           character(1L)))
+  
+  hashes
+  
+}
+
 
 neon_regex <- function(){
   site <- "(NEON\\.D\\d\\d\\.\\w{4})\\."                     # \\1
@@ -111,8 +149,11 @@ neon_regex <- function(){
 #' neon_store()
 #' 
 #' 
-neon_store <- function(product = NULL, dir = neon_dir()){
-  meta <- neon_index(product = product, dir = dir)
+neon_store <- function(product = NULL, table = NULL, dir = neon_dir()){
+  meta <- neon_index(product = product, 
[…]

### 2020-06-05 R/api.R
fix download

@@ -6,6 +6,7 @@
 #' @inheritParams neon_download
 #' @importFrom httr GET content
 #' @importFrom jsonlite fromJSON
+#' @importFrom tibble as_tibble
 #' @export
 neon_sites <- function(api = "https://data.neonscience.org/api/v0", 
                        .token = Sys.getenv("NEON_TOKEN")){
@@ -14,9 +15,7 @@ neon_sites <- function(api = "https://data.neonscience.org/api/v0",
                     httr::add_headers("X-API-Token" = .token))
   txt <- httr::content(resp, as="text")
   sites <- jsonlite::fromJSON(txt)[[1]]
-  
-  class(sites) <- c("tbl_df", "tbl", "data.frame")
-  sites
+  tibble::as_tibble(sites)
 }
 
 
@@ -76,9 +75,7 @@ neon_products <- function(
   if(!is.null(fields))
     products <- products[fields]
   
-  # allow tibble-style printing for tidyverse users
-  class(products) <- c("tbl_df", "tbl", "data.frame")
-  products
+  tibble::as_tibble(products)
   
 }
 
@@ -157,7 +154,7 @@ neon_data <- function(product,
     dat$files
   }))
 
-  data
+  tibble::as_tibble(data)
 }
 
 
@@ -304,8 +301,8 @@ neon_download <- function(product,
   for(i in seq_along(unique_files$url)){
     if(!quiet) pb$tick()
     # consider benchmarking if alternatives are faster? curl_download?
-    curl::curl_download(unique_files[i, "url"], 
-                        unique_files[i, "path"])
+    curl::curl_download(unique_files[i, "url"][[1]], 
+                        unique_files[i, "path"][[1]])
   }
 
   if(verify) {
@@ -324,11 +321,12 @@ neon_download <- function(product,
   
   
   # unzip and remove .zips
-  zips <- unique_files[grepl("[.]zip", unique_files$dir), "dir"]
+  zips <- unique_files[grepl("[.]zip", unique_files$path), "path"]
   lapply(zips, unzip, exdir = dir)
   unlink(zips)
   # remove .zip file?
   
+  unique_files <- tibble::as_tibble(unique_files)
   invisible(unique_files)
 }
 

### 2020-06-05 R/store.R

@@ -81,12 +81,7 @@ neon_index <- function(product = NULL,
   ## Compute hashes, if requested
   meta_c$hash <- file_hash(meta_c$path, hash = hash)
   
-  class(meta_c) <- c("tbl_df", "tbl", "data.frame")
-  meta_c
-  
-  
-  
-  
+  tibble::as_tibble(meta_c)
 }
 
 #' @importFrom openssl md5 sha1 sha256
@@ -94,14 +89,17 @@ file_hash <- function(x, hash = "md5"){
   
   
   if(is.null(hash)) return(NULL)
+  if(length(x) == 0)  return(NULL)
   
   hash_fn <- switch(hash, 
          "md5" = openssl::md5,
          "sha1" = openssl::sha1,
          "sha256" = openssl::sha256,
          NULL)
+  
   if(is.null(hash_fn)) {
-   warning(paste("No function for", hash, "found", call. = FALSE))
+   warning(paste("No function for", hash, 
+                 "found", call. = FALSE))
    return(NULL)
   }
   

### 2020-06-05 tests/testthat/test-api.R

@@ -16,9 +16,9 @@ test_that("neon_sites()", {
   skip_if_offline()
   
   x <- neon_sites()
-
   expect_is(x, "data.frame")
-    
+  expect_match(colnames(x), "siteCode", all=FALSE)
+  expect_gt(nrow(x), 10)
 })
 
 
@@ -30,6 +30,8 @@ test_that("neon_products()", {
     
   x <- neon_products()
   expect_is(x, "data.frame")
+  expect_match(colnames(x), "productCode", all=FALSE)  
+  expect_gt(nrow(x), 10)
   
   
 })
@@ -48,6 +50,7 @@ test_that("neon_data()", {
                  end_date = "2019-08-01")
   
   expect_is(x, "data.frame")
+  expect_gt(nrow(x), 1)
   
 })
 
@@ -73,6 +76,7 @@ test_that("neon_download()", {
                      start_date = "2019-06-01",
                      end_date = "2019-08-01")
   expect_is(x, "data.frame")
+  expect_gt(nrow(x), 1)
   
   
 })

### 2020-06-05 R/store.R
update docs, examples

@@ -271,11 +271,17 @@ vroom_ragged <- function(files){
 #' Indeed, `neon_download()` is designed in precisely this way, to allow easy
 #' updating of downloads without re-downloading older data.
 #' 
-#' @importFrom utils bibentry uitls
+#' @importFrom utils bibentry person
 #' @export
 #' @examples 
 #' 
 #' neon_citation("DP1.10003.001")
+#' 
+#' ## or the citation for all products in store:
+#' neon_citation()
+#' 
+#' ## as bibtex
+#' format(neon_citation("DP1.10003.001"), "bibtex")
 #' 
 neon_citation <- function(product = NULL, 
                           download_date = Sys.Date(),

### 2020-06-05 R/api.R
fix examples/tests

@@ -36,7 +36,7 @@ neon_sites <- function(api = "https://data.neonscience.org/api/v0",
 #' products <- neon_products()
 #' 
 #' # Or search for a keyword
-#' i <- grepl("bird", products$keyword)
+#' i <- grepl("bird", products$keywords)
 #' products[i, c("productCode", "productName")]
 #' 
 #' }
@@ -222,6 +222,8 @@ neon_dir <- function(){
 #' See details. 
 #' @param file_regex Download only files matching this pattern.  See details.
 #' @param quiet Should download progress be displayed?
+#' @param verify Should downloaded files be compared against the MD5 hash
+#' reported by the NEON API to verify integrity? (default TRUE)
 #' @param dir Location where files should be downloaded. By default will
 #' use the appropriate applications directory for your system 
 #' (see [rappdirs::user_data_dir]).  This default also be configured by
@@ -244,8 +246,9 @@ neon_dir <- function(){
 #'  neon_download("DP1.10003.001", start_date = "2019-01-01", site = "YELL")
 #'                
 #'  ## Advanced use: filter for a particular table in the product
-#'  neon_download("DP1.10003.001",
+#'  neon_download(product = "DP1.10003.001",
 #'                start_date = "2019-01-01",
+#'                end_date = "2020-01-01",
 #'                site = "YELL",
 #'                file_regex = ".*brd_countdata.*\\.csv")
 #' 
@@ -301,8 +304,8 @@ neon_download <- function(product,
   for(i in seq_along(unique_files$url)){
     if(!quiet) pb$tick()
     # consider benchmarking if alternatives are faster? curl_download?
-    curl::curl_download(unique_files[i, "url"][[1]], 
-                        unique_files[i, "path"][[1]])
+    curl::curl_download(unique_files$url[i], 
+                        unique_files$path[i])
   }
 
   if(verify) {
@@ -321,9 +324,10 @@ neon_download <- function(product,
   
   
   # unzip and remove .zips
-  zips <- unique_files[grepl("[.]zip", unique_files$path), "path"]
+  zips <- unique_files$path[grepl("[.]zip", unique_files$path)]
   lapply(zips, unzip, exdir = dir)
   unlink(zips)
+  
   # remove .zip file?
   
   unique_files <- tibble::as_tibble(unique_files)

### 2020-06-05 tests/testthat/test-api.R

@@ -76,7 +76,7 @@ test_that("neon_download()", {
                      start_date = "2019-06-01",
                      end_date = "2019-08-01")
   expect_is(x, "data.frame")
-  expect_gt(nrow(x), 1)
+  expect_gt(nrow(x), 0)
   
   
 })

### 2020-06-05 R/store.R
set token in GH actions, handle edge case

@@ -197,6 +197,9 @@ neon_read <- function(table, ..., files = NULL, dir = neon_dir()){
   if(is.null(files)){
     meta <- neon_index(table = table, hash = NULL, dir = dir)
     files <- meta$path
+  } else if(length(files) == 0){ 
+    warning("no files or table name provided")
+    return(NULL)
   }
   
   if(length(files) == 0){

### 2020-06-05 R/store.R
avoid error if is example is run before download

@@ -192,17 +192,15 @@ neon_store <- function(product = NULL, table = NULL, dir = neon_dir()){
 #' files <- neon_index(table = "brd_countdata-expanded")$path
 #' neon_read(files = files)
 #' 
-neon_read <- function(table, ..., files = NULL, dir = neon_dir()){
+neon_read <- function(table = NULL, ..., files = NULL, dir = neon_dir()){
   
   if(is.null(files)){
     meta <- neon_index(table = table, hash = NULL, dir = dir)
     files <- meta$path
-  } else if(length(files) == 0){ 
-    warning("no files or table name provided")
-    return(NULL)
   }
   
   if(length(files) == 0){
+    if(is.null(table)) table <- "unspecified tables"
     warning(paste("no files found for", table, "in", dir, "\n",
                   "perhaps you need to download them first?"))
     return(NULL)

### 2020-06-05 README.Rmd
include type in tablename to avoid ambiguity

@@
   fig.path = "man/figures/README-",
-  out.width = "100%"
+  out.width = "100%",
+  cache = TRUE
 )
 library(neonstore)
-
+Sys.setenv("NEONSTORE_HOME" = tempfile())
 ```
@@
 
-The goal of neonstore is to provide quick access and persistent storage of NEON data tables.
+`neonstore` provides quick access and persistent storage of NEON data tables.
 
@@
 
-```{r example, cache=TRUE}
+```{r}
 library(neonstore)
@@
 
-Now, view your store of NEON products.  These will persist between sessions,
-so you only need to download once, or to retrieve updates.
-(Note: the individual files making up a data product can all be listed by
-`neon_index()`).
+Now, view your store of NEON products:
 
@@
 
-Read in all the component tables into a single data.frame
+These will persist between sessions, so you only need to download once
+or to retrieve updates. `neon_store()` can take arguments to filter by product
+or pattern in table name, e.g. `neon_store(table = "brd")`.
+
+Once you determine the table of interest, you can
+read in all the component tables into a single `data.frame`
 
 ```{r}
-neon_read("brd_countdata")
+neon_read("brd_countdata-expanded")
 ```
@@
 You can request multiple products at once using vector notation. 
-You can optionally include date filters, e.g. to request only records more recent
-than a certain date. 
-Doing so will preserve API quota and improve speed (see API limits, below).
+You can optionally include date filters, e.g. to request only records more 
+recent than a certain date.  Doing so will preserve API quota and improve speed
+(see API limits, below).
+
+## Provenance
 
 Because `neonstore` only stores raw data products as returned from the NEON API,
-this makes it easy to track provenance and compare results across other analyses
+it can easily determine which files have already been downloaded, and only
+download new files without requiring the user to specify specific dates. 
+(It must still query the API for all the metadata in the requested date range).
+This same modular approach also makes it easy to track _data provenance_, an 
+essential element of reproduciblity in comparing results across other analyses
 of the NEON data.
+
+We can list precisely which component files are being read in by `neon_read()`
+by consulting `neon_index()`:
+
+```{r}
+raw_files <- neon_index(table = "brd_countdata-expanded")
+raw_files
+```
+
+`neon_read()` is a relatively trivial function that simply passes this file 
+list to `vroom::vroom()`, a fast, vectorized parser that can easily read in 
+a single table that is broken into many separate files.
+
+Imagine instead that we use the common pattern of downloading
+these raw files, stacks and possibly cleans the data, saving only this derived
+product while discarding the individual files.  Now imagine a second researcher,
+at some later date, queries the API over the same reported range of dates and 
+sites, uses the same software package to stack the tables, only to discover the
+resulting table is somehow different from ours (e.g. by comparing file hashes).
+Pinpointing the source of the discrepancy would be challenging and 
+labor-intensive.
+
+In contrast, the same detective-work would be easy with the `neonstore` file
+list.  We can confirm if the API had returned the same number of 
+raw files with the same names; and better, can verify integrity of the contents
+by comparing hashes of files now being returned to those recorded
+by `neon_index()`.  In this way, we could determine if any additional files
+had been included or pinpoint any files that may have changed.
+
+As such, users might want to store the `neon_index()` `data.frame` for the
+table(s) they have used as part of their analysis, including the individual
+file hashes.  One can also generate a zip of all the data files for 
+archival purposes. (Note that NEON is an Open Data provider, see
+[LICENCE](https://www.neonscience.org/data/about-data/data-policies).)
+
+```{r message=FALSE}
+write.csv(raw_files, "index.csv")
+zip("brd_countdata.zip", raw_files$path)
+```
+
+## Data citation
+
+Generate the appropriate citation for your data:
+
+```{r}
+neon_citation()
+```
 
@@
 require each month of sampling separately!)
+
+```{r include=FALSE}
+unlink("brd_countdata.zip")
+unlink("index.csv")
+Sys.unsetenv("NEONSTORE_HOME")
+```

### 2020-06-08 README.Rmd
:pen: more context about motivations

@@
 `neonstore` provides quick access and persistent storage of NEON data tables. 
+`neonstore` emphasizes simplicity and a clean data provenance trail, see 
+(Provenance)[/#Provenance].
 
@@
 
-## Details
+Two other functions access additional API endpoints 
+that may also be of interest. `neon_sites()` returns a `data.frame`
+of site information, including site descriptions and 
+the ecological domain that each site falls into:
 
-`neon_download()` will omit downloads of any existing data files. 
-You can request multiple products at once using vector notation. 
-You can optionally include date filters, e.g. to request only records more 
-recent than a certain date.  Doing so will preserve API quota and improve speed
-(see API limits, below).
+```{r}
+neon_sites()
+```
+
+Lastly, `neon_products()` returns a table with a list of all neon products,
+which may be useful for data discovery or additional metadata about any
+given product:
+
+```{r}
+neon_products()
+```
+
+## Design Details / comparison to `neonUtilities`
+
+`neonstore` is not meant as a replacement to the `neonUtilities` package
+developed by NEON staff.  `neonUtilities` performs a range of product-specific
+data querying, parsing, and data manipulation beyond what is provided by NEON's
+API or web interface. `neonUtilities` also provides other utilities for working 
+with NEON data beyond the scope of the NEON API or the data download/ingest 
+process. While this processing is undoubtedly useful, it may make
+it difficult to compare results or analyses based on data downloaded and accessed
+using `neonUtilities` R package with analyses based on data accessed directly
+from the web interface, the API, or another tool (or even a different release
+of the `neonUtilities`).
+
+By contrast, `neonstore` aims to do far less.  `neonstore` merely automates the
+download of individual NEON data files.  In contrast to `neonUtilities` which 
+by default "stacks" these raw files into single tables and discards the raw 
+data, `neonstore` preserves only the raw files in the store, stacking the 
+individual tables "on demand" using `neon_read()`. `neon_read()` is a thin
+wrapper around the `vroom` package, 
+[Hester & Wickham, 2020](https://vroom.r-lib.org), which uses the `altrep` 
+mechanism in R to provide very fast reads of rectangular text data into R,
+and trivially handles the case of a single table being broken across many files.
+Some NEON tables are not entirely consistent in their use of columns across the
+individual site-month files, so `neon_read()` transparently checks for this,
+reading in groups of files sharing all matching columns with `vroom` before
+binding the groups together. This makes it easier to always trace an analysis
+back to the original input data, makes it easier to update input data files 
+without facing the challenge of either downloading & stacking the whole 
+data product from scratch again or having to keep track of some previously
+downloaded data file.
+
+A few other differences are also worth noting.
+
+- `neonstore` aims to provide persistent storage, writing raw data files to
+  the appropriate app directory for your operating system (see `rappdirs`,
+  [Ratnakumar et al 2016](https://CRAN.R-project.org/package=rappdirs)). 
+  More details about this can be found in [Provenance](/#Provenance), below.
+- `neon_download()` provides  clean and concise progress bars for the two key
+  processes involved: querying the API to obtain download URLs (which involves no 
+  large data transfer but counts against API rate limiting, see below), and the
+  actual file downloads.
+- `neon_download()` will verify the integrity of file downloads against the MD5
+  hashes provided. 
+- `neon_download()` will omit downloads of any existing data files in the local 
+  store.  
+- You can request multiple products at once using vector notation, though API
+  rate limiting may interfere with large requests.
+- `neon_download()` uses `curl::curl_download()` instead of `downloadr` package
+  used in `neonUtilities`, which can be finicky on Windows and older versions of R. 
+- `neonstore` has slightly lighter dependencies: only `vroom` and `httr`, and
+  packages already used by one of those two (`curl`, `openssl`).
+
+Like `neonUtilities`, You can optionally include site and date filters,
+e.g. to request only records more  recent than a certain date.  Doing so will
+preserve API quota and improve speed (see API limits, below). `neonUtilities`
+is also far more widely tested and has extensive error handling tailored to
+individual data products.
 
@@
 
-```{r message=FALSE}
+```{r message=FALSE, eval=FALSE}
 write.csv(raw_files, "index.csv")
@@
 
-Generate the appropriate citation for your data:
+Always remember to cite your data sources!
+`neonstore` knows how to generate the appropriate citation for the data
+in your local store (or any specific product).
 
@@
 (It would be much more efficient on the NEON server if the API could take
-queries of the from `/data/<product>/<site>, and pool the results, rather than
+queries of the from `/data/<product>/<site>`, and pool the results, rather than
 require each month of sampling separately!)

### 2020-06-24 README.Rmd
:pen:
update mode

@@
 ```{r}
-raw_files <- neon_index(table = "brd_countdata-expanded")
+raw_files <- neon_index(table = "brd_countdata-expanded", hash="md5")
 raw_files
@@
 
-```{r message=FALSE, eval=FALSE}
+```{r message=FALSE}
 write.csv(raw_files, "index.csv")

### 2020-06-24 notebook/import.Rmd
update mode

--- 2020-06-01 notebook/import.Rmd
@@
 # Import
-
-```{r birds}
-bench_time(
-  neon_download("DP1.10003.001", file_regex = ".*basic.*[.]csv")
-)
-```
-
-```{r beetles}
-bench::bench_time(
-  neon_download("DP1.10022.001")
-)
-```
-
-```{r ticks}
-bench::bench_time(
-  neon_download("DP1.10093.001", quiet = TRUE)
-)
-```
-
-```{r mammals}
-bench::bench_time(
-  neon_download("DP1.10072.001",  quiet = TRUE)
-)
-```
-
-```{r zooplankton}
-neon_download("DP1.20219.001")
-```
-
-```{r macroinvert}
-neon_download("DP1.20120.001")
-```
-
-```{r mozzies}
-bench::bench_time(
-  neon_download("DP1.10043.001")
-)
-```
-
-# Explore
-
-```{r}
-library(dplyr)
-
-meta <- neon_index()
-meta %>% count(product)
-```
-
-```{r}
-meta %>% count(name)
-```
-
-```{r}
-
-bench::bench_time({
-f <- meta %>% filter(name == "brd_countdata", type == "basic") %>% pull(path)
-df <- neon_read(f) # vroom with error handler
-})
-
-```
-
-```{r}
-
-bench::bench_time({
-f <- meta %>% filter(name == "mam_perplotnight", type == "basic") %>% pull(path)
-df <- neon_read(f) # vroom with error handler
-})
-
-```
-
-```{r}
-as_tibble(df)
-```
 

### 2020-06-24 R/neon_download_s3.R

@@ -94,7 +94,7 @@ neon_download_s3 <- function(product,
 # In particular, this uses `marker` instead of `continuation-token`
 
 s3_index_public <- function(
-  bucket = "https://minio-jetstream.carlboettiger.info/neonstore/"){
+  bucket = "https://minio.jetstream.carlboettiger.info/neonstore/"){
   
   ## use xml2 conditionally
   if(!requireNamespace("xml2", quietly = TRUE)){

### 2020-06-24 notebook/product-testing.R

@@ -15,6 +15,7 @@ sites <- site$siteCode
 ## (will almost surely run foul of rate limiting -- work by site or something)
 ## or work by product code at least?
 for(p in codes){
+  message(paste("product", p))
   for(s in sites){
     neon_download(p, file_regex = "[.]zip", site = s, keep_zip = TRUE)
     Sys.sleep(10)

### 2020-07-20 paper/paper.Rmd
:construction: boilerplate

--- 2020-06-24 README.Rmd
@@
 ---
-output: 
-  github_document:
-    df_print: tibble
+title: "taxadb: A High-Performance Local Taxonomic Database Interface"
+author:
+  - name: "Carl Boettiger"
+    affiliation: ucb
+  - name: "Christine Laney"
+    affiliation: neon
+  - name: "Quinn Thomas"
+    affiliation: vt
+address:
+  - code: ucb
+    address: "Dept of Environmental Science, Policy, and Management, University of California Berkeley, Berkeley CA 94720-3114, USA"
+  - code: neon
+    address: "NEON"
+  - code: vt
+    address: "Virginia Tech"
+abstract: |
+  The National Ecological Observatory Network
+journal: "TBD"
+date: "`r Sys.Date()`"
+bibliography: refs.bib
+layout: 3p
+header-includes:
+   - \usepackage{lineno}
+   - \linenumbers
+output: rticles::elsevier_article
 ---
@@
 
-```{r, include = FALSE}
+```{r, setup, include = FALSE}
 knitr::opts_chunk$set(
@@
 
-# neonstore
+The National Ecological Observatory Network (NEON) provides open access ecological data as 181 (active and planned) data products at 81 (permanent and relocatable) sites throughout the United States.
+The `neonstore` R package seeks to provide quick, convenient access and persistent storage of NEON data files.
+`neonstore` design emphasizes simplicity and data provenance.
+here, we describe the design, use, and rationale for the package.
 
-<!-- badges: start -->
-[![R build status](https://github.com/cboettig/neonstore/workflows/R-CMD-check/badge.svg)](https://github.com/cboettig/neonstore/actions)
-[![Codecov test coverage](https://codecov.io/gh/cboettig/neonstore/branch/master/graph/badge.svg)](https://codecov.io/gh/cboettig/neonstore?branch=master)
-<!-- badges: end -->
-
-`neonstore` provides quick access and persistent storage of NEON data tables. 
-`neonstore` emphasizes simplicity and a clean data provenance trail, see 
-(Provenance)[/#Provenance].
+NEON data products are built around a highly atomized file design: both observation systems (OS) and instrument systems (IS) data products are divided into separate files by site and month.
+Products generated by the airborne observation platform (AOP) are typically divided by type, flight date, flight line, and grid cell.
+This atomized approach facilitates sub-setting and data transfer, (particularly relevant to R users where most objects are typically stored in memory)
 
@@
 ```{r}
+## Note: NEON's recommendation may soon change
 neon_citation()
@@
 ```
+
+## Non-stacking files and low-level interface
+
+At it's core, `neonstore` is simply a mechanism to download files from the NEON API.
+While the `.csv` files from the Observation Systems (OS, e.g. bird count surveys),
+and Instrument Systems (e.g. aquatic sensors) are typically stacked into large
+tables, other products, such as the `.laz` and `.tif` images produced by the
+airborne observation platform LIDAR and cameras may require a different approach.
+
+```{r}
+# Read in a large file list for illustration purposes
+cper_data <- readr::read_csv("https://minio.thelio.carlboettiger.info/shared-data/neon_data_catalog.csv.gz")
+
+## Typically one would read all files in local store, e.g. list.file(neon_dir())
+df <- neon_filename_parser(cper_data$name)
+```
+
+```{r}
+library(dplyr)
+df %>% count(EXT, sort=TRUE)
+```
+
+We can take a look at all `laz` LIDAR files:
+
+```{r}
+df %>% 
+  filter(EXT == "laz")
+```
+
+Note that many of the airborne observation platform (AOP) products, such as
+these LIDAR files, do not include the PRNUM or REV components that make up part
+of the `productCode`s used in the NEON `product` tables.

### 2020-07-21 paper/paper.Rmd
:pen:

@@
 `neonstore` design emphasizes simplicity and data provenance.
-here, we describe the design, use, and rationale for the package.
+Here, we describe the design, use, and rationale for the package.
+
+## Rationale and package design
 
 NEON data products are built around a highly atomized file design: both observation systems (OS) and instrument systems (IS) data products are divided into separate files by site and month.
-Products generated by the airborne observation platform (AOP) are typically divided by type, flight date, flight line, and grid cell.
-This atomized approach facilitates sub-setting and data transfer, (particularly relevant to R users where most objects are typically stored in memory)
+Meanwhile, products generated by the airborne observation platform (AOP) are typically divided by type, flight date, flight line, and grid cell.
+This atomized approach facilitates sub-setting and data transfer, because users can identify ahead of time which products, locations, and time ranges they need.
+As some NEON data products can be quite large (NEON's data catalog to date is in excess of 1 petabyte, primarily due to the size of many AOP products), being able to only download or only read into R the files of interest is already important to save bandwidth or computer memory, and this will be increasingly true as NEON data accumulates in the years ahead.
+This approach also facilitates data provenance and reproducibility, because it means that an update or correction to data collected at a particular site and time results only in a change to a single file and not the whole data product.
+Users can easily verify whether or not the update has changed data in their analyses by comparing file hashes, rather than having to read in and examine data meticulously row by row.
 
-## Installation
+However, the highly atomized structure of NEON's data can also become a barrier to entry without convenient tools for working
+with many files.
+Most analyses will inevitably require multiple files to be combined or stacked, e.g.
+to create a unified record for a product across many months or years of sampling or across many sites.
+When the process of combining files is complex or slow to execute, many researchers will resort to saving the resulting 'stacked' products as a basis for future work.
+Such workflows may be more cumbersome to update as additional data continues to be published, or to compare against any subsequent corrections to previously-published data.
+`neonstore` seeks to help users download and manage NEON files locally in a way that
+lets analyses be quickly traced back to the raw data files.
 
-Install the development version from [GitHub](https://github.com/) with:
+`neonstore` is distinguished as much by what it does not do as by what it does.
+All `neonstore` operations are intended to be transparent and intuitive, and could easily be duplicated in another computational language or command line interface.
+Specifically, `neonstore` will never automatically "clean" data: opaque data transformations make output sensitive to the version and implementation details of the tool in question.
+A user should expect the same file from a "download" operation regardless of the software or software version used to perform the download.
+Functions which in read in the data should faithfully reproduce the raw data so that analyses can easily be compared to those using other software tools.
+When functions perform tasks like guessing the data type (numeric, string, Date, etc) of a column, or adding additional columns to stack two otherwise matching tables in which one has recorded an additional variable, they should do so in a manner that is transparent, controllable, and consistent with the behaviour of common functions used in the language.
 
-``` r
-# install.packages("devtools")
-devtools::install_github("cboettig/neonstore")
+## `neonstore` Functionality
+
+`neonstore` functionality can be divided into two halves: functions that interface with the NEON Application Programming Interface (API), and functions for working with a local store of previously-downloaded NEON files.
+
+```{r}
+library(neonstore)
 ```
-## Quickstart
 
-Discover data products of interest:
+### NEON API functions
+
+`neon_products()` and `neon_sites()` bind the NEON REpresentational State Transfer (REST) API endpoints for `/products` and `/sites`, providing information about each as a data frame.
+This can be useful for product discovery and occassionally for referencing other useful metadata about products and sites.
+Each of these functions requires only a single API call (see discussion of rate limiting below).
 
@@
 products <- neon_products()
-
-i <- grepl("Populations", products$themes)
-products[i, c("productCode", "productName")]
-
-i <- grepl("bird", products$keywords)
-products[i, c("productCode", "productName")]
+products
 ```
 
-Download all data files in the bird survey data products.
+```{r}
+sites <- neon_sites()
+sites
+```
+
+The workhorse function of `neonstore` is `neon_download()`, which downloads all files identified by a given product code, optionally constrained to specific site, date range, or file name pattern.
+For instance, we can download all of the NEON landbird survey data, identified as NEON data product `DP1.10003.001`:
 
 ```{r}
-library(neonstore)
 neon_download("DP1.10003.001")
+```
 
-```
+To download this data, `neon_download()` must make many API calls.
+It first uses `neon_sites()` to reference all sites containing this data product.
+Then for each site, for each month for which that site collected data, the function will make an API call to the `/data` endpoint (using the `neonstore` function, `neon_data()`), which returns a table of files found for that product at the requested site and month range.
+The  table from `neon_data()` indicates 4 values: the file name, the MD5 hash of the data file (confusingly labeled as the crc32 sum, though CRC32 hashes are used only for AOP data products.
+Note that MD5 hashes are length-32 character strings, while the CRC32 hashes are 8 characters long), the file size (in bytes -- encoded as a character because some file sizes are too long to be represented as an integer), and a download URL.
+The download URL embeds a temporary access token which expires after an hour, so users must complete all the downloads before this time expires.
+The access string for each file is unique, thus it is (at the time of writing) impossible to download any given data file without making a corresponding API request, even if the precise file name and address is already known.
+
+The three-digit code identifies this as a Level 1 data product (DP1), product number 10003, revision 001, indicating this is the first and so far the only protocol used for bird survey sampling.
 
@@
 ```
+
+Note that at this time, `neonstore` does not provide a complete
 

### 2020-07-21 paper/paper.Rmd
:pen:

@@
 The access string for each file is unique, thus it is (at the time of writing) impossible to download any given data file without making a corresponding API request, even if the precise file name and address is already known.
+For the most part, this issue is invisible to `neonstore` users, as `neon_download()` calls automatically pass all download URLs to the download step.
+
+```{r}
+neon_download("DP1.10003.001", 
+               start_date = "2018-01-01", 
+               end_date = "2019-01-01",
+               site = "YELL")
+
+ ## Advanced use: filter for a particular table in the product
+ neon_download(product = "DP1.10003.001",
+               start_date = "2018-01-01",
+               end_date = "2019-01-01",
+               site = "YELL",
+               file_regex = ".*brd_countdata.*\\.csv")
+```
+
+`neon_download()`
 

### 2020-07-28 README.Rmd
spelling

@@
 `neonstore` emphasizes simplicity and a clean data provenance trail, see 
-(Provenance)[/#Provenance].
+Provenance section below.
 
@@
   [Ratnakumar et al 2016](https://CRAN.R-project.org/package=rappdirs)). 
-  More details about this can be found in [Provenance](/#Provenance), below.
+  More details about this can be found in Provenance, below.
 - `neon_download()` provides  clean and concise progress bars for the two key

### 2020-08-04 paper/paper.Rmd
WIP

@@
 
-However, the highly atomized structure of NEON's data can also become a barrier to entry without convenient tools for working
-with many files.
+However, the highly atomized structure of NEON's data can also become a barrier to entry without convenient tools for working with many files.
 Most analyses will inevitably require multiple files to be combined or stacked, e.g.
@@
 Such workflows may be more cumbersome to update as additional data continues to be published, or to compare against any subsequent corrections to previously-published data.
-`neonstore` seeks to help users download and manage NEON files locally in a way that
-lets analyses be quickly traced back to the raw data files.
+`neonstore` seeks to help users download and manage NEON files locally in a way that lets analyses be quickly traced back to the raw data files.
 
@@
 ```
+
+### A local store for neon files
+
+The location of the local store can be configured depending on the context.
+The default behavior is to rely on a user-specific location appropriate for the operating system, e.g.
+`~/.local/share/neonstore` on Linux, `C:\Users\<username>\AppData\Local\neonstore` on Windows, or `~/Library/Application Support/neonstore` on Mac OS X, as determined by the `rappdirs` R package which provides appropriate modern defaults [@rappdirs].
+This approach works well for a single user setup such as a laptop, where a user can share the store across multiple projects in different directories.
+On a server system that is accessed by multiple users, an lower-level location that permits all users to read and write files to the store may be preferable.
+Because `neonstore` only writes raw NEON data to the store, there is no risk of overwriting work of other users, while the shared storage can reduce consumption of redundant download and storage.
+In yet other cases, a user may prefer a path that is specific to a single project.
+
+It is of course possible to use a temporary directory such as `tempdir()` for the storage location.
+Such an approach is the default download location of the alternative package, `neonUtilities`, although this can be configured there as well.
+However, doing so defeats the purpose of having a persistent store, which allows us to both preserve the original raw data files between R sessions.
+Because new NEON data files are released regularly, this approach allows us to easily download only the new files.
+This contrasts to the typical workflow in `neonUtilities`, in which a user discards the individually downloaded files and is left to their own devices on how to keep store the resulting stacked data in order to avoid re-downloading each time.
+
+The best way to set the file directory of the store is to declare the path using the environmental variable, `NEONSTORE_HOME`, for example, using a `.Renviron` file in the user or system home directory.
+For one-off use, all functions that access the store take an optional argument `dir` to specify the directory location, (which defaults to `neon_dir()`, a function that either uses the value set by `NEONSTORE_HOME` or the `rappdirs` location if unset).
+However, using the environmental variable instead results in more portable code, by allowing other users to set their own storage locations independently instead of hard-wiring it into the R code.
 
@@
 The workhorse function of `neonstore` is `neon_download()`, which downloads all files identified by a given product code, optionally constrained to specific site, date range, or file name pattern.
-For instance, we can download all of the NEON landbird survey data, identified as NEON data product `DP1.10003.001`:
+For instance, we can download all of the NEON landbird survey data, identified as NEON data product `DP1.10003.001`: (The three-digit code identifies this as a Level 1 data product (DP1), product number 10003, revision 001, indicating this is the first and so far the only protocol used for bird survey sampling).
 
@@
 
-`neon_download()`
+## Interacting with downloaded files
 
-The three-digit code identifies this as a Level 1 data product (DP1), product number 10003, revision 001, indicating this is the first and so far the only protocol used for bird survey sampling.
+Even adding a handful of products can generate thousands or hundreds of thousands of files in a local store.
+Surprisingly, these can even exceed the capacity of common bash commands like `ls *.csv`.
+While a directory structure could mitigate this issue, `neonstore` sidesteps this issue through a convenient R-based interface.
+The function `neon_index()` will list all recognized files found in the local store as a `data.frame`.
+`neon_index()` uses a sophisticated filename parser, `neon_parse_filenames()`, that understands all documented file naming conventions listed at <https://data.neonscience.org/file-naming-conventions>, and quite a few that are not yet listed (internal NEON communication).
+The most common of these are displayed in the table:
 
-Now, view your store of NEON products:
+```{r}
+neon_index()
+```
 
-```{r }
+`neon_index()` takes optional arguments such as `product`, `start_date`, `end_date`, and `site` to allow for convenient sub-setting.
+Because most observational and instrument-based products are divided by site and month, this is still not a convenient way to browse the data.
+The helper function, `neon_store()`, summarizes products that all share the same file description and can thus be easily stacked.
+Like `neon_index()`, optional arguments can filter the results to specific products, dates, and sites.
+
+```{r}
 neon_store()
@@
 
-These will persist between sessions, so you only need to download once
-or to retrieve updates. `neon_store()` can take arguments to filter by product
-or pattern in table name, e.g. `neon_store(table = "brd")`.
-
-Once you determine the table of interest, you can
-read in all the component tables into a single `data.frame`
+In this example, we recognize several tables corresponding to the NEON bird survey, product `DP01.10003.001` (all prefixed with `brd_`).  We can read in all the individual files
 

### 2020-08-12 R/neon_data.R
Automatically pause for API rate limit
Also, handle the change in how files hashes are reported by the API.

@@ -48,17 +48,26 @@ neon_data <- function(product,
     return(invisible(NULL))
   }
   
+  ## Adjust for rate-limiting
+  batch <- 1000                 # authenticated
+  if(.token == "") batch <- 200 # unauthenticated
+  
+  
   ## Extract the file list from the data endpoint.  O(sites * months) calls
   pb <- progress::progress_bar$new(
     format = "  querying API [:bar] :percent eta: :eta",
     total = length(data_api), 
     clear = FALSE, width= 60)
   
-  resp <- lapply(data_api, function(x){
+  resp <- vector("list", length = length(data_api))
+  for(i in seq_along(data_api)){
     if(!quiet){ pb$tick() }
-    httr::GET(x, httr::add_headers("X-API-Token" = .token))
-  })
-  
+    resp[[i]] <- httr::GET(data_api[[i]], httr::add_headers("X-API-Token" = .token))
+    if(i %% batch == 0){
+      message("Rate limiting enforced, waiting 100s\n")
+      Sys.sleep(100)
+    }
+  }
   
   ## Format the result as a data.frame
   data <- do.call(rbind,
@@ -81,8 +90,8 @@ neon_warn_http_errors <- function(x){
   status <- httr::status_code(x)
   if(status < 400) return(invisible(0L))
   out <- httr::content(x)
-  warning(paste(out$error$status, "error:", out$error$detail), call. = FALSE)
-  1L
+  warning(paste(status, "error:", as.character(out)), call. = FALSE)
+  invisible(1L)
 }
 
 ## Some DataUrls reported by products table include date ranges that are not valid, e.g.: 

### 2020-08-12 R/neon_download.R

@@ -115,7 +115,9 @@ neon_download <- function(product,
   
   ## Time to download, verify, and unzip
   download_all(files$url, files$path, quiet)
-  verify_hash(files$path, files$crc32, verify)
+  
+  algo <- hash_type(files)
+  verify_hash(files$path, files[algo], verify, algo)
   if(unzip) unzip_all(files$path, dir)
   
   ## file metadata (url, path, md5sum)  
@@ -150,11 +152,19 @@ download_filters <- function(files, file_regex,
   if(type == "basic")
     files <- files[!grepl("expanded", files$name), ]
   
-  ## Filter out duplicate files, e.g. have identical crc32 values
-  files <- take_first_match(files, "crc32")
+  ## Filter out duplicate files, e.g. have identical hash values
+  files <- take_first_match(files, hash_type(files))
   files
 }
 
+
+hash_type <- function(df){
+  type <- "md5"
+  if(is.null(df[[type]]) | any(is.na(df[[type]]))){
+    type <- "crc32"
+  }
+  type
+}
 
 
 download_all <- function(addr, dest, quiet){
@@ -184,11 +194,20 @@ unzip_all <- function(path, dir, keep_zips = FALSE){
 
 
 
-
-verify_hash <- function(path, hash, verify){
+#' @importFrom digest digest
+#' @importFrom openssl md5
+verify_hash <- function(path, hash, verify, algo = "md5"){
+  if(any(is.na(hash))){
+    return(NULL)
+  }
+  
+  
+  hashfn <- switch(algo,
+                   md5 = function(x) as.character(openssl::md5(file(x))),
+                   crc32 =  function(x) digest::digest(x, "crc32", file=TRUE))
+  
   if(verify){
-    md5 <- vapply(path, 
-                  function(y) as.character(openssl::md5(file(y))),
+    md5 <- vapply(path, hashfn,
                   character(1L), USE.NAMES = FALSE)
     i <- which(md5 != hash)
     if(length(i) > 0) {
@@ -197,12 +216,7 @@ verify_hash <- function(path, hash, verify){
                     path[i]), call. = FALSE)
     }
   }
-  
 }
-
-
-
-
 
 
 

### 2020-08-12 R/neon_data.R
pretty-print

@@ -64,8 +64,8 @@ neon_data <- function(product,
     if(!quiet){ pb$tick() }
     resp[[i]] <- httr::GET(data_api[[i]], httr::add_headers("X-API-Token" = .token))
     if(i %% batch == 0){
-      message("Rate limiting enforced, waiting 100s\n")
-      Sys.sleep(100)
+      message("\nNEON rate limiting enforced, pausing for 100s\n")
+      Sys.sleep(101)
     }
   }
   

### 2020-08-12 README.Rmd
update README

--- 2020-07-28 README.Rmd
@@
 [![Codecov test coverage](https://codecov.io/gh/cboettig/neonstore/branch/master/graph/badge.svg)](https://codecov.io/gh/cboettig/neonstore?branch=master)
+[![CRAN status](https://www.r-pkg.org/badges/version/neonstore)](https://CRAN.R-project.org/package=neonstore)
 <!-- badges: end -->

### 2020-08-12 R/neon_read.R
additional validation on neon_read
Avoid reading in old files if they have been updated in our store.

@@ -52,6 +52,10 @@ neon_read <- function(table = NA,
                        ext = ext,
                        hash = NULL, 
                        dir = dir)
+    
+    ## If timestamp has changed but other metadata is the same, we only want the newer version
+    meta <- filter_duplicates(meta)
+    
     files <- meta$path
   }
   
@@ -141,5 +145,23 @@ ragged_bind <- function(x){
   do.call(rbind, x)
   
 }
+
+## Sometimes a NEON file will have changed
+filter_duplicates <- function(meta){
+  meta$timestamp <- as.POSIXct(meta$timestamp, format = "%Y%m%dT%H%M%OS")
+  meta_b <- meta[order(meta$timestamp, decreasing = TRUE),] 
+  meta_b$id <- paste(meta$product, meta$site, meta$table, meta$month, sep="-")
+  out <- take_first_match(meta_b, "id")
+  
+  if(dim(out)[[1]] < dim(meta)[[1]])
+    message("Some raw files were detected with updated timestamps.\n
+            Using only most updated file to avoid duplicates.")
+  ## FIXME Maybe we should verify if the hash of said file(s) has changed.
+  ## maybe we should provide more information on how to check these?
+  
+  out
+}
+
+
 
 

### 2020-08-13 R/neon_read.R
handle null meta case

@@ -53,6 +53,9 @@ neon_read <- function(table = NA,
                        hash = NULL, 
                        dir = dir)
     
+    if(is.null(meta)) return(NULL)
+    if(dim(meta)[[1]] == 0 )  return(NULL)
+    
     ## If timestamp has changed but other metadata is the same, we only want the newer version
     meta <- filter_duplicates(meta)
     

### 2020-09-03 R/neon_index.R
support sensor parser column additions
closes #8

@@ -70,6 +70,7 @@ neon_index <- function(product = NA,
   
   ## Compute hashes, if requested
   meta$hash <- file_hash(meta$path, hash = hash)
+  meta$timestamp <-  as.POSIXct(meta$timestamp, format = "%Y%m%dT%H%M%OS")
   
   tibble::as_tibble(meta)
 }

### 2020-09-03 R/neon_read.R

@@ -12,11 +12,25 @@
 #' of columns.  `neon_read` works around this by parsing such tables in
 #' groups of matching schema, which is still reasonably fast.
 #' 
-#' For convenience, `neon_read` takes the name of a table in the local store.
+#' NEON sensor data products currently do not include important metadata columns
+#' containing DomainID, SiteID, horizontalPosition, verticalPosition, and
+#' publicationDate in the data files themselves, but only encode this in the 
+#' in the raw file names. All though these values are shared across a raw 
+#' data file, this information is lost when stacking the tables unless explicit
+#' columns are added to the data.  This requires us to parse the files
+#' one-by-one, which is much slower.  By default this information is added to
+#' the table, altering the stacked table schema from that of the raw table.  
+#' Disable this behavior by setting `sensor_metadata = FALSE`.  Future
+#' NEON sensor data products may start including this information in 
+#' the raw data files, as is already the case for observational data.
+#' 
 #' 
 #' @param table the name of a downloaded NEON table in the store,
 #'  see [neon_store]
-#' @param .id add an additional id column with metadata from filename.
+#' @param sensor_metadata logical, default TRUE. Should we add 
+#' metadata fields from file names of sensor data into the table?  Adds
+#' DomainID, SiteID, horizontalPosition, verticalPosition, and publicationDate.
+#' Results in slower parsing.  
 #' @param ... additional arguments to [vroom::vroom], can usually be omitted.
 #' @param files optionally, specify a vector of file paths directly (e.g. as
 #' provided from [neon_index]) and specify `table` argument as NULL.
@@ -32,6 +46,10 @@
 #' files <- neon_index(table = "brd_countdata-expanded")$path
 #' neon_read(files = files)
 #' 
+#' ## Sensor inputs will add metadata columns by default
+#' neon_read("waq_instantaneous",site = c("CRAM","SUGG"))
+#'
+#' 
 neon_read <- function(table = NA,
                       product = NA, 
                       site = NA,
@@ -40,7 +58,7 @@ neon_read <- function(table = NA,
                       ext = NA,
                       dir = neon_dir(),
                       files = NULL,
-                      .id = NA,
+                      sensor_metadata = TRUE,
                       ...){
   
   if(is.null(files)){
[…]

### 2020-09-03 R/neon_index.R
- deprecate neon_store - allow timestamp filter in neon_index / neon_read

@@ -12,10 +12,27 @@
 #' `neon_index()` parses this metadata from the file name string and returns
 #' the information in a convenient table, along with a path to each file.
 #' 
+#' **Regarding timestamps:**  NEON will occasionally publish new versions of
+#' previously-released raw data files (which may or may not actually differ).
+#' The NEON download API, and hence [neon_download()], only serve the most recent
+#' of such files, but earlier versions may still exist in your local `neonstore`
+#' if you downloaded them before the updated files were released.  By default,
+#' [read_neon()] will always select the most recent of such files, thus avoiding
+#' duplication and providing the most updated data.  For reproducibility however,
+#' it may be necessary to access older version instead. Setting the timestamp 
+#' argument allows the user to filter out newer files and select the original
+#' ones instead.  Unfortunately, at this time users cannot request the outdated
+#' data files from NEON API.  For strict reproducibility, users should also
+#' archive their local store, see [neon_archive()].
+#' 
+#' @seealso [neon_archive()], [neon_download()]
+#' 
 #' @param product Include only files matching this NEON productCode(s)
 #' @param table Include only files matching this table name (or regex pattern). 
 #' (optional).
 #' @param ext only match files with this file extension(s)
+#' @param timestamp only match timestamps prior this. See details in [neon_index()].
+#'        Should be a datetime POSIXct object (or coerce-able string)
 #' @param hash name of a hashing algorithm to check file integrity. Can be
 #'  `"md5"`, `"sha1"`, or `"sha256"` currently; or set to [NULL] (default)
 #'   to skip hash computation.
@@ -46,6 +63,7 @@ neon_index <- function(product = NA,
                        end_date = NA,
                        type = NA,
                        ext = NA,
+                       timestamp = NA,
                        hash = NULL,
                        dir = neon_dir()){
   
@@ -57,6 +75,7 @@ neon_index <- function(product = NA,
   
   ## Include full paths to files
   meta$path <- file.path(dir, meta$path)
+  meta$timestamp <-  as.POSIXct(meta$timestamp, format = "%Y%m%dT%H%M%OS")
   
   ## Apply filters
   meta <- meta_filter(meta, 
@@ -66,11 +85,11 @@ neon_index <- function(product = NA,
                       start_date = start_date,
                       end_date = end_date,
                       type = type,
[…]

### 2020-09-03 R/neon_read.R

@@ -56,6 +56,7 @@ neon_read <- function(table = NA,
                       start_date = NA,
                       end_date = NA,
                       ext = NA,
+                      timestamp = NA,
                       dir = neon_dir(),
                       files = NULL,
                       sensor_metadata = TRUE,

### 2020-09-03 tests/testthat/test-store.R

@@ -15,8 +15,7 @@ test_that("neon_index()", {
   x <- neon_index()
   expect_is(x, "data.frame")
   expect_true(any(grepl("DP1.10003.001", x$product)))
-  x <- neon_index(dir = tempfile())
-  
+
 })
 
 
@@ -31,22 +30,14 @@ test_that("neon_index options", {
   expect_true(any(grepl("DP1.10003.001", x$product)))
   expect_true(any(grepl("hash", colnames(x))))
   
+  ## No data expected if timestamp predates data publication times!
+  x <- neon_index(timestamp = as.POSIXct("2018-01-01 01:00:00"))
+  expect_true(nrow(x) == 0)
   
   x <- neon_index(dir = tempfile())
-  
+  expect_null(x)
 })
 
-test_that("neon_store()", {
-  
-  x <- neon_store()
-  expect_true(any(grepl("brd_countdata", x$table)))
-  d <- tempfile()
-  expect_message(
-    x <- neon_store(dir = d)
-  )
-  expect_null(x)
-  
-})
 
 test_that("neon_read()", {
   

### 2020-09-03 README.Rmd
update docs

--- 2020-08-12 README.Rmd
@@
 ```{r }
-neon_store()
+neon_index()
 ```
@@
 These will persist between sessions, so you only need to download once
-or to retrieve updates. `neon_store()` can take arguments to filter by product
-or pattern in table name, e.g. `neon_store(table = "brd")`.
+or to retrieve updates. `neon_index()` can take arguments to filter by product
+or pattern (regular expression) in table name, e.g. `neon_index(table = "brd")`.
 
@@
 
+If `neon_download()` exceeds the API request limit (with or without the token),
+`neonstore` will simply pause for the required amount of time to avoid 
+rate-limit-based errors.
+
 [The NEON API now rate-limits requests.](https://data.neonscience.org/data-api/rate-limiting/#api-tokens).
-Using a personal token will increase the number of requests you can make.
-See that link for directions on registering for a token.
+Using a personal token will increase the number of requests you can make before
+encountering this delay.  See link for directions on registering for a token.
 Then pass this token in `.token` argument of `neon_download()`,
@@
 to your local `.Renviron` file in your user's home directory. 
+`neon_download()` must first query each the API of each NEON site which collects
+that product, for each month the product is collected.
 
-`neon_download()` must first query each the API of eacn NEON site which collects
-that product, for each month the product is collected.
 (It would be much more efficient on the NEON server if the API could take

### 2020-09-03 R/neon_index.R

@@ -23,9 +23,9 @@
 #' argument allows the user to filter out newer files and select the original
 #' ones instead.  Unfortunately, at this time users cannot request the outdated
 #' data files from NEON API.  For strict reproducibility, users should also
-#' archive their local store, see [neon_archive()].
+#' archive their local store.
 #' 
-#' @seealso [neon_archive()], [neon_download()]
+#' @seealso  [neon_download()]
 #' 
 #' @param product Include only files matching this NEON productCode(s)
 #' @param table Include only files matching this table name (or regex pattern). 

### 2020-09-03 R/neon_read.R

@@ -7,6 +7,10 @@
 #' data table that has been sharded into many files like this much
 #' much faster than other parsers can read in each table iteratively, 
 #' (and thus can greatly out-perform the 'stacking" methods in `neonUtilities`).
+#' 
+#' When reading in very large numbers of files, it may be helpful to set
+#' `altrep = FALSE` to opt out of `vroom`'s fast altrep mechanism, which
+#' can cause [neon_read()] to fail when stacking thousands of files.
 #'
 #' Unfortunately, not all datasets are entirely consistent in their use
 #' of columns.  `neon_read` works around this by parsing such tables in
@@ -26,7 +30,7 @@
 #' 
 #' 
 #' @param table the name of a downloaded NEON table in the store,
-#'  see [neon_store]
+#'  see [neon_index]
 #' @param sensor_metadata logical, default TRUE. Should we add 
 #' metadata fields from file names of sensor data into the table?  Adds
 #' DomainID, SiteID, horizontalPosition, verticalPosition, and publicationDate.
@@ -47,7 +51,7 @@
 #' neon_read(files = files)
 #' 
 #' ## Sensor inputs will add metadata columns by default
-#' neon_read("waq_instantaneous",site = c("CRAM","SUGG"))
+#' neon_read("waq_instantaneous", site = c("CRAM","SUGG"))
 #'
 #' 
 neon_read <- function(table = NA,

### 2020-09-03 README.Rmd
we already have neon_export / neon_import!

@@
 ```{r}
-raw_files <- neon_index(table = "brd_countdata-expanded", hash="md5")
+raw_files <- neon_index(table = "brd_countdata", hash="md5")
 raw_files
@@
 write.csv(raw_files, "index.csv")
-zip("brd_countdata.zip", raw_files$path)
+neon_export("my_neondata.zip", product = "DP1.10003.001")
+neon_import("my_neondata.zip")
 ```

### 2020-09-03 R/neon_export.R

@@ -7,12 +7,14 @@
 #' re-created by other users without an API key, or without waiting for
 #' the individual download, or any other tiem you want to share or 
 #' distribute your local store. 
+#' 
 #' @return table of selected files and metadata, from [neon_index()], invisibly.
 #' @param archive path to the zip archive to be created.
+#' @details
 #' @inheritParams neon_index
 #' @importFrom zip zip unzip
 #' @seealso [neon_import()], [neon_citation()]
-neon_export <-  function(archive = "neon.zip",
+neon_export <-  function(archive = paste(Sys.Date(), "neonstore.zip", sep="-"),
                          product = NA, 
                          table = NA, 
                          site = NA,
@@ -20,6 +22,7 @@ neon_export <-  function(archive = "neon.zip",
                          end_date = NA,
                          type = NA,
                          ext = NA,
+                         timestamp = NA,
                          hash = NULL,
                          dir = neon_dir()){
   
@@ -30,6 +33,7 @@ neon_export <-  function(archive = "neon.zip",
                      end_date = end_date,
                      type = type,
                      ext = ext,
+                     timestamp = timestamp,
                      hash = hash, 
                      dir = dir)
   

### 2020-09-03 tests/testthat/test-store.R

@@ -31,7 +31,7 @@ test_that("neon_index options", {
   expect_true(any(grepl("hash", colnames(x))))
   
   ## No data expected if timestamp predates data publication times!
-  x <- neon_index(timestamp = as.POSIXct("2018-01-01 01:00:00"))
+  x <- neon_index(timestamp = as.POSIXct("2017-01-01 01:00:00"))
   expect_true(nrow(x) == 0)
   
   x <- neon_index(dir = tempfile())

### 2020-09-03 README.Rmd
export neon_export :smile:

@@
 
-```{r, include = FALSE}
+```{r setup, include = FALSE}
 knitr::opts_chunk$set(
@@
   out.width = "100%",
-  cache = TRUE
+  cache = FALSE
 )
@@
 
+```{r}
+library(neonstore)
+```
+
 Discover data products of interest:
@@
 ```{r}
-library(neonstore)
 neon_download("DP1.10003.001")
@@
 ```{r message=FALSE}
+# index list alone
 write.csv(raw_files, "index.csv")
+
+# zip archive
 neon_export("my_neondata.zip", product = "DP1.10003.001")
@@
 ```{r include=FALSE}
-unlink("brd_countdata.zip")
+unlink("my_neondata.zip")
 unlink("index.csv")

### 2020-09-03 R/neon_export.R

@@ -9,11 +9,14 @@
 #' distribute your local store. 
 #' 
 #' @return table of selected files and metadata, from [neon_index()], invisibly.
-#' @param archive path to the zip archive to be created.
-#' @details
+#' @param archive path to the zip archive to be created.#' 
 #' @inheritParams neon_index
 #' @importFrom zip zip unzip
+#' @export
 #' @seealso [neon_import()], [neon_citation()]
+#' @examples 
+#' 
+#' neon_export()
 neon_export <-  function(archive = paste(Sys.Date(), "neonstore.zip", sep="-"),
                          product = NA, 
                          table = NA, 
@@ -37,6 +40,17 @@ neon_export <-  function(archive = paste(Sys.Date(), "neonstore.zip", sep="-"),
                      hash = hash, 
                      dir = dir)
   
+  if(is.null(meta)){
+    message("No files found.")
+    return(NULL)
+  }
+  
+  if(nrow(meta) == 0){
+    message("No files found.")
+    return(NULL)
+  }
+  
+  
   zip::zipr(archive, meta$path, include_directories=FALSE)
   invisible(meta)
 }
@@ -50,6 +64,11 @@ neon_export <-  function(archive = paste(Sys.Date(), "neonstore.zip", sep="-"),
 #' @param overwrite should we overwrite any existing files?
 #' @inheritParams neon_index
 #' @seealso [neon_export()]
+#' @export
+#' @examples 
+#' archive <- tempfile()
+#' neon_export(archive)
+#' neon_import(archive)
 #' 
 neon_import <- function(archive, overwrite = TRUE, dir = neon_dir()){
   if(!dir.exists(dir)) dir.create(dir, FALSE, TRUE)

### 2020-09-03 tests/testthat/test-store.R
fix test

@@ -31,7 +31,7 @@ test_that("neon_index options", {
   expect_true(any(grepl("hash", colnames(x))))
   
   ## No data expected if timestamp predates data publication times!
-  x <- neon_index(timestamp = as.POSIXct("2017-01-01 01:00:00"))
+  x <- neon_index(timestamp = as.POSIXct("2016-01-01 01:00:00"))
   expect_true(nrow(x) == 0)
   
   x <- neon_index(dir = tempfile())

### 2020-09-03 R/neon_index.R
update docs

@@ -17,7 +17,7 @@
 #' The NEON download API, and hence [neon_download()], only serve the most recent
 #' of such files, but earlier versions may still exist in your local `neonstore`
 #' if you downloaded them before the updated files were released.  By default,
-#' [read_neon()] will always select the most recent of such files, thus avoiding
+#' [neon_read()] will always select the most recent of such files, thus avoiding
 #' duplication and providing the most updated data.  For reproducibility however,
 #' it may be necessary to access older version instead. Setting the timestamp 
 #' argument allows the user to filter out newer files and select the original

### 2020-09-03 R/neon_export.R
tempdir rules

@@ -16,7 +16,10 @@
 #' @seealso [neon_import()], [neon_citation()]
 #' @examples 
 #' 
-#' neon_export()
+#' archive <- tempfile()
+#' dir <- tempdir()
+#' neon_export(archive, dir = dir)
+#' 
 neon_export <-  function(archive = paste(Sys.Date(), "neonstore.zip", sep="-"),
                          product = NA, 
                          table = NA, 
@@ -42,12 +45,12 @@ neon_export <-  function(archive = paste(Sys.Date(), "neonstore.zip", sep="-"),
   
   if(is.null(meta)){
     message("No files found.")
-    return(NULL)
+    return(invisible(NULL))
   }
   
   if(nrow(meta) == 0){
     message("No files found.")
-    return(NULL)
+    return(invisible(NULL))
   }
   
   
@@ -66,11 +69,16 @@ neon_export <-  function(archive = paste(Sys.Date(), "neonstore.zip", sep="-"),
 #' @seealso [neon_export()]
 #' @export
 #' @examples 
+#' 
 #' archive <- tempfile()
 #' neon_export(archive)
 #' neon_import(archive)
 #' 
 neon_import <- function(archive, overwrite = TRUE, dir = neon_dir()){
+  if(!file.exists(archive)){
+    warning(paste(archive, "not found"), call. = FALSE)
+    return(invisible(NULL))
+  }
   if(!dir.exists(dir)) dir.create(dir, FALSE, TRUE)
   zip::unzip(archive, overwrite = overwrite, exdir = dir)
 }

### 2020-09-03 tests/testthat/test-store.R
testing

@@ -31,7 +31,7 @@ test_that("neon_index options", {
   expect_true(any(grepl("hash", colnames(x))))
   
   ## No data expected if timestamp predates data publication times!
-  x <- neon_index(timestamp = as.POSIXct("2016-01-01 01:00:00"))
+  x <- neon_index(timestamp = as.POSIXct("2010-01-01 01:00:00"))
   expect_true(nrow(x) == 0)
   
   x <- neon_index(dir = tempfile())

### 2020-09-03 tests/testthat/test-store.R
...

@@ -32,7 +32,7 @@ test_that("neon_index options", {
   
   ## No data expected if timestamp predates data publication times!
   x <- neon_index(timestamp = as.POSIXct("2010-01-01 01:00:00"))
-  expect_true(nrow(x) == 0)
+  expect_true(is.null(x) || nrow(x) == 0)
   
   x <- neon_index(dir = tempfile())
   expect_null(x)

### 2020-09-03 tests/testthat/test-store.R
NA row

@@ -32,7 +32,8 @@ test_that("neon_index options", {
   
   ## No data expected if timestamp predates data publication times!
   x <- neon_index(timestamp = as.POSIXct("2010-01-01 01:00:00"))
-  expect_true(is.null(x) || nrow(x) == 0)
+  ## sometimes draws a row of all NAs
+  expect_true(nrow(x) <= 1)
   
   x <- neon_index(dir = tempfile())
   expect_null(x)

### 2020-09-04 R/neon_export.R
Sensor parser (#15)
* accepted

* prepare for new release

* support sensor parser column additions

closes #8

* - deprecate neon_store
- allow timestamp filter in neon_index / neon_read

* update docs

* we already have neon_export / neon_import!

* export neon_export :smile:

* updated README

* fix test

* update docs

* tempdir rules

* testing

* ...

* NA row

@@ -7,12 +7,20 @@
 #' re-created by other users without an API key, or without waiting for
 #' the individual download, or any other tiem you want to share or 
 #' distribute your local store. 
+#' 
 #' @return table of selected files and metadata, from [neon_index()], invisibly.
-#' @param archive path to the zip archive to be created.
+#' @param archive path to the zip archive to be created.#' 
 #' @inheritParams neon_index
 #' @importFrom zip zip unzip
+#' @export
 #' @seealso [neon_import()], [neon_citation()]
-neon_export <-  function(archive = "neon.zip",
+#' @examples 
+#' 
+#' archive <- tempfile()
+#' dir <- tempdir()
+#' neon_export(archive, dir = dir)
+#' 
+neon_export <-  function(archive = paste(Sys.Date(), "neonstore.zip", sep="-"),
                          product = NA, 
                          table = NA, 
                          site = NA,
@@ -20,6 +28,7 @@ neon_export <-  function(archive = "neon.zip",
                          end_date = NA,
                          type = NA,
                          ext = NA,
+                         timestamp = NA,
                          hash = NULL,
                          dir = neon_dir()){
   
@@ -30,8 +39,20 @@ neon_export <-  function(archive = "neon.zip",
                      end_date = end_date,
                      type = type,
                      ext = ext,
+                     timestamp = timestamp,
                      hash = hash, 
                      dir = dir)
+  
+  if(is.null(meta)){
+    message("No files found.")
+    return(invisible(NULL))
+  }
+  
+  if(nrow(meta) == 0){
+    message("No files found.")
+    return(invisible(NULL))
+  }
+  
   
   zip::zipr(archive, meta$path, include_directories=FALSE)
   invisible(meta)
@@ -46,8 +67,18 @@ neon_export <-  function(archive = "neon.zip",
 #' @param overwrite should we overwrite any existing files?
 #' @inheritParams neon_index
 #' @seealso [neon_export()]
+#' @export
+#' @examples 
+#' 
+#' archive <- tempfile()
+#' neon_export(archive)
+#' neon_import(archive)
 #' 
 neon_import <- function(archive, overwrite = TRUE, dir = neon_dir()){
+  if(!file.exists(archive)){
+    warning(paste(archive, "not found"), call. = FALSE)
+    return(invisible(NULL))
+  }
   if(!dir.exists(dir)) dir.create(dir, FALSE, TRUE)
   zip::unzip(archive, overwrite = overwrite, exdir = dir)
 }

### 2020-09-04 tests/testthat/test-additional.R

@@ -39,6 +39,23 @@ test_that("beetles", {
   
 })
 
+test_that("Aquatic sensor data", {
+  
+  skip_on_cran()
+  skip_if_offline()
+  
+  neon_download("DP1.20288.001", site = c("CRAM","SUGG"), type="basic")
+  df <- neon_read("waq_instantaneous")
+  expect_true("siteID" %in% colnames(df))
+
+  df <- neon_read("waq_instantaneous", sensor_metadata = FALSE)
+  
+  expect_false("siteID" %in% colnames(df))
+
+  
+})
+
+
 test_that("ECdata", {
   
   

### 2020-09-09 paper/paper.Rmd
paper :pen:

@@
 ---
-title: "taxadb: A High-Performance Local Taxonomic Database Interface"
+title: "neonstore: a local, provenance-aware workflow for NEON data"
 author:
@@
     affiliation: ucb
+  - name: "Quinn Thomas"
+    affiliation: vt
   - name: "Christine Laney"
     affiliation: neon
-  - name: "Quinn Thomas"
-    affiliation: vt
+  - name: "Claire Lunch"
+    affiliation: neon
+
 address:
@@
   The National Ecological Observatory Network
-journal: "TBD"
+journal: "Ecoinformatics"
 date: "`r Sys.Date()`"
@@
 Because most observational and instrument-based products are divided by site and month, this is still not a convenient way to browse the data.
-The helper function, `neon_store()`, summarizes products that all share the same file description and can thus be easily stacked.
-Like `neon_index()`, optional arguments can filter the results to specific products, dates, and sites.
 
-```{r}
-neon_store()
-```
-
-In this example, we recognize several tables corresponding to the NEON bird survey, product `DP01.10003.001` (all prefixed with `brd_`).  We can read in all the individual files
+We can read in all the individual files
 
@@
 write.csv(raw_files, "index.csv")
-zip("brd_countdata.zip", raw_files$path)
+neon_export("brd_countdata.zip")
 ```

### 2020-09-09 notebook/efi-download.R

@@ -1,0 +1,25 @@
+Sys.setenv(NEONSTORE_HOME = "/minio/neonstore")
+library(neonstore)
+
+
+neon_download("DP1.10003.001")
+
+start_date <- NA
+#Beetle
+neonstore::neon_download(product="DP1.10022.001", type = "expanded", start_date = start_date)
+
+#Ticks
+neonstore::neon_download(product="DP1.10093.001", type = "expanded", start_date = start_date)
+
+#Terrstrial
+sites <- c("BART", "KONZ", "SRER", "OSBS")
+neonstore::neon_download(product="DP4.00200.001", site = sites, type = "basic", start_date = start_date)
+
+#Aquatic
+#DP1.20053.001
+#DP1.20288.001
+
+sites <- c("BARC", "FLNT")
+neonstore::neon_download(product="DP1.20053.001",site = sites, type = "expanded", start_date = start_date)
+neonstore::neon_download(product="DP1.20288.001",site = sites, type = "expanded", start_date = start_date)
+

### 2020-10-05 R/connection_pane.R
patching up stuff

@@ -25,7 +25,7 @@ neon_pane <- function() {
         )
       },
       listObjects = function(type = "datasets") {
-        tbls <- DBI::dbListTables(neon_db(read_only = TRUE))
+        tbls <- DBI::dbListTables(neon_db())
         data.frame(
           name = tbls,
           type = rep("table", length(tbls)),
@@ -33,7 +33,7 @@ neon_pane <- function() {
         )
       },
       listColumns = function(table) {
-        res <- DBI::dbGetQuery(neon_db(read_only = TRUE),
+        res <- DBI::dbGetQuery(neon_db(),
                                paste("SELECT * FROM", table, "LIMIT 1"))
         data.frame(
           name = names(res), 
@@ -42,7 +42,7 @@ neon_pane <- function() {
         )
       },
       previewObject = function(rowLimit, table) {  #nolint
-        DBI::dbGetQuery(neon_db(read_only = TRUE),
+        DBI::dbGetQuery(neon_db(),
                         paste("SELECT * FROM", table, "LIMIT", rowLimit))
       },
       actions = list(
@@ -55,7 +55,7 @@ neon_pane <- function() {
           callback = sql_action
         )
       ),
-      connectionObject = neon_db(read_only = TRUE)
+      connectionObject = neon_db()
     )
   }
 }
@@ -71,7 +71,7 @@ sql_action <- function() {
   if (requireNamespace("rstudioapi", quietly = TRUE) &&
       exists("documentNew", asNamespace("rstudioapi"))) {
     contents <- paste(
-      "-- !preview conn=neonstore::neon_db(read_only = TRUE)",
+      "-- !preview conn=neonstore::neon_db()",
       "",
       "SELECT * FROM provenance LIMIT 10",
       "",
@@ -87,7 +87,7 @@ sql_action <- function() {
 }
 
 neon_status <- function () {
-  con <- neon_db(read_only = TRUE)
+  con <- neon_db()
   inherits(con, "DBIConnection")
 }
 

### 2020-10-05 R/neon_download.R

@@ -145,11 +145,9 @@ neon_download_ <- function(product,
   ## additional filters on already_have, type and file_regex:
   files <- download_filters(files, file_regex, type, quiet, dir)
   if(is.null(files)){
-    if(!quiet) message("nothing more to download\n")
     return(invisible(NULL)) # nothing to download
   }
   if(length(files) == 0) {
-    if(!quiet) message("nothing more to download\n")
     return(invisible(NULL)) # nothing to download
   }
   
@@ -185,11 +183,14 @@ download_filters <- function(files, file_regex,
                              type, quiet, dir){
   
   if(is.null(files)) return(invisible(NULL)) # nothing to download
+  if(nrow(files) == 0) return(invisible(NULL)) # nothing to download
   
   ## Omit those file names we already have
   already_have <- files$name %in% basename(list.files(dir, recursive = TRUE))
   if(sum(already_have) > 0 && !quiet){
-    message(paste("omitting", sum(already_have), "files previously downloaded"))
+    message(paste("  omitting", 
+                  sum(already_have), 
+                  "files previously downloaded"))
   }
   files <- files[!already_have, ]
   
@@ -201,6 +202,9 @@ download_filters <- function(files, file_regex,
     files <- files[!grepl("basic", files$name), ]
   if(type == "basic")
     files <- files[!grepl("expanded", files$name), ]
+
+  
+  if(nrow(files) == 0) return(invisible(NULL)) # nothing to download
   
   ## Filter out duplicate files, e.g. have identical hash values
   ## (as reported by NEON's own hash)

### 2020-10-06 README.Rmd
update README :pen:

@@
 
-Now, view your store of NEON products:
+View your store of NEON products:
 
@@
 
-These will persist between sessions, so you only need to download once
+These files will persist between sessions, so you only need to download once
 or to retrieve updates. `neon_index()` can take arguments to filter by product
@@
 
-Two other functions access additional API endpoints 
-that may also be of interest. `neon_sites()` returns a `data.frame`
-of site information, including site descriptions and 
-the ecological domain that each site falls into:
+## Database backend
+
+`neonstore` now supports a backend relation database as well. Import data from
+the raw downloaded files using `neon_store()`:
 
 ```{r}
-neon_sites()
+neon_store(table = "brd_countdata-expanded")
 ```
 
-Lastly, `neon_products()` returns a table with a list of all neon products,
-which may be useful for data discovery or additional metadata about any
-given product:
+Alternately, we could import all data tables associated with a given product:
 
 ```{r}
-neon_products()
+neon_store(product = "DP1.10003.001")
 ```
 
-## Design Details / comparison to `neonUtilities`
-
-`neonstore` is not meant as a replacement to the `neonUtilities` package
-developed by NEON staff.  `neonUtilities` performs a range of product-specific
-data querying, parsing, and data manipulation beyond what is provided by NEON's
-API or web interface. `neonUtilities` also provides other utilities for working 
-with NEON data beyond the scope of the NEON API or the data download/ingest 
-process. While this processing is undoubtedly useful, it may make
-it difficult to compare results or analyses based on data downloaded and accessed
-using `neonUtilities` R package with analyses based on data accessed directly
-from the web interface, the API, or another tool (or even a different release
-of the `neonUtilities`).
-
-By contrast, `neonstore` aims to do far less.  `neonstore` merely automates the
-download of individual NEON data files.  In contrast to `neonUtilities` which 
-by default "stacks" these raw files into single tables and discards the raw 
-data, `neonstore` preserves only the raw files in the store, stacking the 
-individual tables "on demand" using `neon_read()`. `neon_read()` is a thin
-wrapper around the `vroom` package, 
-[Hester & Wickham, 2020](https://vroom.r-lib.org), which uses the `altrep` 
-mechanism in R to provide very fast reads of rectangular text data into R,
-and trivially handles the case of a single table being broken across many files.
-Some NEON tables are not entirely consistent in their use of columns across the
-individual site-month files, so `neon_read()` transparently checks for this,
-reading in groups of files sharing all matching columns with `vroom` before
-binding the groups together. This makes it easier to always trace an analysis
-back to the original input data, makes it easier to update input data files 
-without facing the challenge of either downloading & stacking the whole 
-data product from scratch again or having to keep track of some previously
-downloaded data file.
-
-A few other differences are also worth noting.
-
-- `neonstore` aims to provide persistent storage, writing raw data files to
-  the appropriate app directory for your operating system (see `rappdirs`,
-  [Ratnakumar et al 2016](https://CRAN.R-project.org/package=rappdirs)). 
-  More details about this can be found in Provenance, below.
-- `neon_download()` provides  clean and concise progress bars for the two key
-  processes involved: querying the API to obtain download URLs (which involves no 
-  large data transfer but counts against API rate limiting, see below), and the
-  actual file downloads.
-- `neon_download()` will verify the integrity of file downloads against the MD5
-  hashes provided. 
-- `neon_download()` will omit downloads of any existing data files in the local 
-  store.  
-- You can request multiple products at once using vector notation, though API
-  rate limiting may interfere with large requests.
-- `neon_download()` uses `curl::curl_download()` instead of `downloadr` package
-  used in `neonUtilities`, which can be finicky on Windows and older versions of R. 
-- `neonstore` has slightly lighter dependencies: only `vroom` and `httr`, and
-  packages already used by one of those two (`curl`, `openssl`).
-
-Like `neonUtilities`, You can optionally include site and date filters,
-e.g. to request only records more  recent than a certain date.  Doing so will
-preserve API quota and improve speed (see API limits, below). `neonUtilities`
-is also far more widely tested and has extensive error handling tailored to
-individual data products.
-
-## Provenance
-
-Because `neonstore` only stores raw data products as returned from the NEON API,
-it can easily determine which files have already been downloaded, and only
-download new files without requiring the user to specify specific dates. 
-(It must still query the API for all the metadata in the requested date range).
-This same modular approach also makes it easy to track _data provenance_, an 
-essential element of reproduciblity in comparing results across other analyses
-of the NEON data.
-
-We can list precisely which component files are being read in by `neon_read()`
-by consulting `neon_index()`:
+Access an imported table using `neon_table()` instead of `neon_read()`:
 
 ```{r}
-raw_files <- neon_index(table = "brd_countdata", hash="md5")
-raw_files
+neon_table("brd_countdata-expanded")
 ```
 
-`neon_read()` is a relatively trivial function that simply passes this file 
-list to `vroom::vroom()`, a fast, vectorized parser that can easily read in 
-a single table that is broken into many separate files.
-
-Imagine instead that we use the common pattern of downloading
-these raw files, stacks and possibly cleans the data, saving only this derived
-product while discarding the individual files.  Now imagine a second researcher,
-at some later date, queries the API over the same reported range of dates and 
-sites, uses the same software package to stack the tables, only to discover the
-resulting table is somehow different from ours (e.g. by comparing file hashes).
-Pinpointing the source of the discrepancy would be challenging and 
-labor-intensive.
-
-In contrast, the same detective-work would be easy with the `neonstore` file
-list.  We can confirm if the API had returned the same number of 
-raw files with the same names; and better, can verify integrity of the contents
-by comparing hashes of files now being returned to those recorded
-by `neon_index()`.  In this way, we could determine if any additional files
-had been included or pinpoint any files that may have changed.
-
-As such, users might want to store the `neon_index()` `data.frame` for the
-table(s) they have used as part of their analysis, including the individual
-file hashes.  One can also generate a zip of all the data files for 
-archival purposes. (Note that NEON is an Open Data provider, see
-[LICENCE](https://www.neonscience.org/data/about-data/data-policies).)
-
-```{r message=FALSE}
-# index list alone
-write.csv(raw_files, "index.csv")
-
-# zip archive
-neon_export("my_neondata.zip", product = "DP1.10003.001")
-neon_import("my_neondata.zip")
-```
-
-## Data citation
-
-Always remember to cite your data sources!
-`neonstore` knows how to generate the appropriate citation for the data
-in your local store (or any specific product).
+Access the remote database using `neon_db()`.  This is a `DBIConnection` that
+can easily be used with `dplyr` functions like `tbl()` or `filter()`.  
+Remember that `dplyr` translates these into SQL queries that run directly
+on the database.
 
 ```{r}
-neon_citation()
+library(dplyr)
+
+con <- neon_db()
+brd <- tbl(con, "brd_countdata-expanded")
+brd %>% filter(siteID == "ORNL")
+
 ```

### 2020-10-06 R/neon_store.R

@@ -23,6 +23,8 @@ neon_store <- function(table = NA,
                       ext = "csv",
                       deprecated = FALSE)
   
+  
+  if(is.null(index)) index <- data.frame()
   if(nrow(index) == 0){
     if(!is.na(table))
       message(paste("table", table, 
@@ -51,16 +53,20 @@ neon_store <- function(table = NA,
   
     ## Drop rows from the database which come from deprecated files
     drop_deprecated(table, dir, con)
-    files <- index[index$table == table, ]$path     
+    index <- index[index$table == table, ]
+    
+    if(nrow(index) > 0)
     db_chunks(con = con, 
-              files = files,
+              files = index$path,
               table = table, 
               n = n, 
               quiet = quiet, 
               ...)
   })
   ## update the provenance table
-  DBI::dbWriteTable(con, "provenance", index, append = TRUE)
+  if(!is.null(index)){
+    DBI::dbWriteTable(con, "provenance", index, append = TRUE)
+  }
   
   invisible(con)
 }
@@ -76,6 +82,8 @@ stackable_tables <- function(tables){
 
 db_chunks <- function(con, files, table, 
                        n = 200L, quiet = FALSE, ...){
+  
+  if(length(files)==0) return(NULL)
   
   total <- length(files) %/% n
   if(length(files) %% n > 0)  ## and the remainder
@@ -102,8 +110,7 @@ db_chunks <- function(con, files, table,
     show_after = 0,
     width = 80)
   
-  pb$tick(0)
-  
+
   for(i in 0:(total-1)){
     if(!quiet) pb$tick()
     chunk <- na_omit(files[ (i*n+1):((i+1)*n) ])
@@ -126,7 +133,11 @@ omit_imported <- function(con, index){
   if( !("provenance" %in% DBI::dbListTables(con)) ){
     return(index)
   }
-  
+  if(is.null(index)){
+    return(data.frame())
+  }
+    
+    
   DBI::dbWriteTable(con, "zzzfilter", index,
                     overwrite = TRUE, temporary = TRUE)
   query <- paste0(
@@ -159,8 +170,8 @@ drop_deprecated <- function(table,
    return(invisible(NULL))
   } else {
     message(
-      paste("Updated version of previously imported data found.\n",
-            "Overwriting some previously imported rows with revised data."
+      paste("  Updated version of previously imported data found.\n",
+            "  Overwriting some previously imported rows with revised data."
             ))
   }
   

### 2020-10-06 inst/examples/eddy_db.R

@@ -9,11 +9,15 @@ table <- unique(index$table)
 df <- neonUtilities::stackEddy(path)
 
 import_hd5 <- function(path, con, table){
-   out <- neonUtilities::stackEddy(path)
+   suppressMessages({
+     out <- neonUtilities::stackEddy(path)
+   })
+  
    df <- out[[1]]
    df$siteID <- names(out[1])
    df$file <- basename(path)
    dbWriteTable(con, table, df, append = TRUE)
+
 }
 
 con <- neon_db()

### 2020-10-06 tests/testthat/test-db.R
a few more tests

@@ -9,9 +9,22 @@ test_that("neon_db", {
   
   db <- neon_db()
   expect_is(db, "DBIConnection")
+
+  neon_delete_db(ask = FALSE)
   
 })
 
+
+test_that("neon_store error handling", {
+  
+  
+  db <- neon_store(table = "brd_countdata-expanded",
+                   dir = tempdir())
+  
+  db <- neon_store(product = "not-a-product",
+                   dir = tempdir())
+})
+  
 
 test_that("neon_store", {
 

### 2020-10-06 tests/testthat/test-db.R
a few more tests

@@ -85,6 +85,17 @@ test_that("neon_table", {
   expect_identical(colnames(tbl1), colnames(tbl2))
   expect_identical(dim(tbl1), dim(tbl2))
   
+  
+  
+  tbl <- neon_table("brd_countdata-expanded", site="YELL")
+  expect_is(tbl, "data.frame")
+  expect_true(nrow(tbl) > 0)
+
+  tbl <- neon_table("brd_countdata-expanded", site="not-a-site")
+  expect_is(tbl, "data.frame")
+  expect_true(nrow(tbl) == 0)
+  
+  
 })
 
 

### 2020-10-06 tests/testthat/test-index.R

@@ -43,6 +43,9 @@ test_that("neon_index options", {
   
   x <- neon_index(dir = tempfile())
   expect_null(x)
+  
+  expect_warning({
+    x <- neon_index(hash="not-a-hash")
+  })
 })
-
 

### 2020-10-06 R/neon_data.R
a few tweaks

@@ -15,7 +15,8 @@ neon_data <- function(product,
   data_api <- data_api_queries(product = product, 
                                start_date = start_date, 
                                end_date = end_date, 
-                               site = site, 
+                               site = site,
+                               quiet = quiet,
                                api = api,
                                .token = .token)
   
@@ -77,6 +78,7 @@ data_api_queries <- function(product,
                             start_date = NA,
                             end_date = NA,
                             site = NA,
+                            quiet = FALSE,
                             api = "https://data.neonscience.org/api/v0", 
                             .token = Sys.getenv("NEON_TOKEN")){
   

### 2021-03-08 README.Rmd
update docs

@@
 Sys.setenv("NEONSTORE_HOME" = tempfile())
+Sys.setenv("NEONSTORE_DB" = tempfile())
+
 ```
@@
 ```{r}
-neon_table("brd_countdata-expanded")
+neon_table("brd_countdata")
 ```
@@
 library(dplyr)
-
 con <- neon_db()
-brd <- tbl(con, "brd_countdata-expanded")
+brd <- tbl(con, "brd_countdata-basic-DP1.10003.001")
 brd %>% filter(siteID == "ORNL")
@@
 ```
+
+Note that we need to include the product name in the table name when accessing the database, as table names alone may not be unique.
+RStudio users can also list and explore all tables interactively in the Connections pane in RStudio using `neon_pane()`.
 
@@
 Sys.unsetenv("NEONSTORE_HOME")
+Sys.unsetenv("NEONSTORE_DB")
 ```

### 2021-03-12 paper/paper.Rmd

--- 2020-08-04 paper/paper.Rmd
@@
     affiliation: ucb
+  - name: "Quinn Thomas"
+    affiliation: vt
   - name: "Christine Laney"
     affiliation: neon
-  - name: "Quinn Thomas"
-    affiliation: vt
+  - name: "Claire Lunch"
+    affiliation: neon
 address:
@@
   The National Ecological Observatory Network
-journal: "TBD"
+journal: preprint
 date: "`r Sys.Date()`"
@@
 
-The National Ecological Observatory Network (NEON) provides open access ecological data as 181 (active and planned) data products at 81 (permanent and relocatable) sites throughout the United States.
+The National Ecological Observatory Network (NEON) represents an important resource and a major investment in data collection infrastructure needed to address many of the grand challenges facing envionmental science [@NAS_NEON; @NAS_GRAND].
+
+provides open access ecological data as 181 (active and planned) data products at 81 (permanent and relocatable) sites throughout the United States.
 The `neonstore` R package seeks to provide quick, convenient access and persistent storage of NEON data files.

### 2021-03-23 paper/paper.Rmd
:pen:

@@
 ---
-title: "neonstore: a local, provenance-aware workflow for NEON data"
+title: "neonstore: provenance-aware persistent storage for NEON data"
 author:
@@
 
-provides open access ecological data as 181 (active and planned) data products at 81 (permanent and relocatable) sites throughout the United States.
-The `neonstore` R package seeks to provide quick, convenient access and persistent storage of NEON data files.
-`neonstore` design emphasizes simplicity and data provenance.
+NEON provides access to 181 (active and planned) ecological data products at 81 (permanent and relocatable) sites throughout the United States.
+The `neonstore` R package seeks to provide provenance aware, high performance access and persistent storage of NEON data files.
 Here, we describe the design, use, and rationale for the package.
@@
 
-NEON data products are built around a highly atomized file design: both observation systems (OS) and instrument systems (IS) data products are divided into separate files by site and month.
-Meanwhile, products generated by the airborne observation platform (AOP) are typically divided by type, flight date, flight line, and grid cell.
+Provenance and performance are two central aspects to the design of `neonstore`.
+A focus on provenance means that analyses can always be traced back to individual NEON data files, and supports both workflows that require access to the most recent or updated data and workflows that require repeated access to persistent, stable data files.
+A focus on performance reflects the potentially very large size of its aggregated data products, which can easily exceed available RAM on most machines (e.g.
+10 GB - 10 TB range, depending on products and sites considered).
+
+NEON data products are distributed using a highly atomized file design: with each data product from either both observation systems (OS) and instrument systems (IS) divided into separate files by site and month.
+Products generated by the airborne observation platform (AOP) are typically divided by type, flight date, flight line, and grid cell.
 This atomized approach facilitates sub-setting and data transfer, because users can identify ahead of time which products, locations, and time ranges they need.
-As some NEON data products can be quite large (NEON's data catalog to date is in excess of 1 petabyte, primarily due to the size of many AOP products), being able to only download or only read into R the files of interest is already important to save bandwidth or computer memory, and this will be increasingly true as NEON data accumulates in the years ahead.
+As some NEON data products can be quite large (NEON's complete data catalog to date is already in excess of 1 petabyte, primarily due to the size of many AOP products), being able to download or read only the files of interest is already important to save bandwidth or computer memory, or disk space.
 This approach also facilitates data provenance and reproducibility, because it means that an update or correction to data collected at a particular site and time results only in a change to a single file and not the whole data product.
-Users can easily verify whether or not the update has changed data in their analyses by comparing file hashes, rather than having to read in and examine data meticulously row by row.
-
-However, the highly atomized structure of NEON's data can also become a barrier to entry without convenient tools for working with many files.
+However, the highly atomized structure of NEON's data can also become a barrier to entry.
 Most analyses will inevitably require multiple files to be combined or stacked, e.g.
 to create a unified record for a product across many months or years of sampling or across many sites.
-When the process of combining files is complex or slow to execute, many researchers will resort to saving the resulting 'stacked' products as a basis for future work.
-Such workflows may be more cumbersome to update as additional data continues to be published, or to compare against any subsequent corrections to previously-published data.
+In addition to releasing new data files each month, NEON will occasionally update previously released data files.
+More confusingly, NEON frequently updates the timestamp associated with a file even when the file has not changed.
+Typically these events are associated with the update of a separate but related file.
+
+Any analysis is only as good as as the data supporting it.
+Because NEON data is subject to both revisions (where previously released data are corrected) and new releases (where more recently collected data is made public), researchers will seek to update previous analyses with corrected data and new observations.
+`neonstore` enables scripted workflows through which such updates can occur automatically by re-running the same script.
+However, when new or revised data significantly alter results, it is also important to be able to pinpoint precisely what has changed and why.
+These elements creates five distinct tasks for managing NEON data provenance: (1) The ability to determine which of each of the individual raw data files were used in an analysis (2) The ability to append new data as it is collected and released (3) the ability to transparently replace existing data when corrections are published, (4) the ability to avoid re-downloading or re-stacking data that has been previously processed, (5) the ability to freeze the data used in a given analysis, so that results can reliably be repeated without accidentally including revised or newer data products.
+
+To provide these five abilities for data provenance, `neonstore` maintains a local file store containing all raw data files downloaded from NEON.
+Additionally, each file is entered into a local registry, along with the file's MD5 checksum (CRC32 for AOP products) and currently associated NEON release (if any).
+This local registry is kept in a Lightning Memory Mapped Database, LMDB, an extremely fast, open source key-value store indexed on both file name and checksum.
+MD5 checksums effectively indicate unique content.
+It is theoretically possible for a malicious actor to create a file with different contents but the same checksum, but vanishingly unlikely to occur by chance.
+On requesting new files to download, `neonstore` compares file names and hashes reported by the API to those in the local store, allowing it to omit any data that has already been downloaded, even if the timestamp in a file name has changed but the contents remain the same.
+
+NEON's filenaming convention provided essential metadata about each file, such as the data product (product level, product number, revision number), site, year and month of the release, file description, a timestamp.
+Sensor data (IS) also include additional information about the horizontal and vertical position and meaasurement interval (frequency) of the sensor product.
+AOP products use a somewhat different convention with flight information.
+`neonstore` recognizes all published NEON file naming conventions and automatically parses NEON filenames into their component parts.
+Data are organized into subdirectories based on product, site, and year-month respectively.
+Data or metadata files shared across all months or all sites of a product are stored in the corresponding parent directory.
+The function `neon_index()` returns a `data.frame` listing every recognized individual file in the local store, along with the corresponding metadata fields parsed from the filename.
+Optional arguments can be given to restrict this table to files matching specific products, sites, date ranges, etc.
+
+Box 1: Provenance workflow
+- Existing NEON data used in an analysis to support a result.
+- A data file is corrected for an error.
+Revised data file is downloaded, imported into `neonstore` database, replacing deprecated data.
+Deprecated data file is retained in local file store for future reference.
+Analysis confirms same result
+- Next month's data are released, downloaded, imported by `neonstore`, and then used to revise model and provide updated result
+
 `neonstore` seeks to help users download and manage NEON files locally in a way that lets analyses be quickly traced back to the raw data files.

### 2021-03-29 paper/paper.Rmd
:pen:

@@
 
-## Rationale and package design
+# Rationale and package design
 
@@
 
+## Provenance
+
+`neonstore` focus on provenance emphasizes being able to trace results back precisely to individual raw data files provided by the NEON API.
 NEON data products are distributed using a highly atomized file design: with each data product from either both observation systems (OS) and instrument systems (IS) divided into separate files by site and month.
@@
 to create a unified record for a product across many months or years of sampling or across many sites.
-In addition to releasing new data files each month, NEON will occasionally update previously released data files.
-More confusingly, NEON frequently updates the timestamp associated with a file even when the file has not changed.
-Typically these events are associated with the update of a separate but related file.
+`neonstore` provides functions to stack data products that have been atomized into separate files by month, site or sensor position into single `data.frame` objects, store stacked tables in a local, high-performance database, and track individual data rows back to their raw data files.
 
+Managing provenance of NEON data means dealing with the addition of new data, corrections published to old data, and changes to file names that do not represent changes in underlying data.
 Any analysis is only as good as as the data supporting it.
@@
 `neonstore` enables scripted workflows through which such updates can occur automatically by re-running the same script.
-However, when new or revised data significantly alter results, it is also important to be able to pinpoint precisely what has changed and why.
+More confusingly, NEON frequently updates the timestamp associated with a file even when the file has not changed.
+Typically these events are associated with the update of a separate but related file.
+In such cases, `neonstore` can avoid downloading and importing these unchanged data, ensuring that automated workflows do not run unnecessarily.
+When new or revised data significantly alter results, it is also important to be able to pinpoint precisely what has changed and why.
 These elements creates five distinct tasks for managing NEON data provenance: (1) The ability to determine which of each of the individual raw data files were used in an analysis (2) The ability to append new data as it is collected and released (3) the ability to transparently replace existing data when corrections are published, (4) the ability to avoid re-downloading or re-stacking data that has been previously processed, (5) the ability to freeze the data used in a given analysis, so that results can reliably be repeated without accidentally including revised or newer data products.
@@
 Optional arguments can be given to restrict this table to files matching specific products, sites, date ranges, etc.
+
+## Performance
+
+`neonstore` is also concerned with performance of research workflows based on NEON data.
+Native operations in R are based in memory, making it difficult to work with objects that are larger than RAM.
+Researchers seeking to take full advantage of NEON data collected across all available sites and years can quickly run ub against the limits of data which can be held and processed in working memory.
+The R language also supports interfaces with relational databases [@dbi], with the widely used package `dplyr` capable of seamlessly tanslating common functions (e.g.
+`filter`, `select`, `mutate`, `group_by`, `summarise`, `inner_join`, etc) into SQL queries that can be run by external database servers [@dplyr; @dbplyr].
+This allows users to perform common tasks on data much larger than would fit into working RAM.
+Traditional relational databases such as MariaDB and Postgres use a server-client model with row-oriented data structures, suitable for concurrent read and write operations by many simultaneous client processes acccessing a central database (e.g.
+many client banks simultaneously updating a central registry of accounts, see ACID).
+Modern data science workflows frequently benefit more from column-oriented data structures where simultaneous operations (ACID transactions) are not required.
+This makes it possible to both eliminate the requirement of separately installing and running a database server while also opening the door to substantial performance improvements beyond the reach of traditional relational database clients.
+`duckdb` is a leading example of such a columnar database [@duckdb], with seamless support for R integration and superior performance benchmarks to both traditional relational databases and in-memory operations in R using `dplyr`.
+
+`neonstore` allows users to easily import tabular NEON data products (including `.h5` based eddy covariance products) into a local `duckdb` database using the function `neon_store()`.
+If desired, users can instead use their own existing databases by providing the appropriate `DBI::dbConnection` object to the `db` argument of `neon_store()`.
+`neon_store()` stacks corresponding individual raw tabular data files by month, site, and (for IS data) sensor position into individual database tables.
+`neon_store()` is capable of detecting revisions to previously released data products, removing outdated rows and replacing them with updated rows without accidental duplication of data.
+Data revisions are indicated by downloading files which share the same data product, site, month and sensor position file name as files already present in the local store, but have an updated timestamp and differing content hash.
+`neon_store()` provides a warning whenever existing data tables are updated, as this may require previous analyses to be re-run.
+Because sensor data does not indicate the site or position information in individual data files (this is provided only through the file name) `neonstore` stacking functions add additional columns for these variables automatically.
+Additionally, a `file` column is always included indicating the raw file source for each row.
+Many NEON data products come in two types: a "basic" format and an "expanded" format which may include additional columns providing data fields that were not included in the original product specification.
+NEON does not always use globally unique names for table descriptions, particularly for metadata tables such as `sensor_positions`.
+To avoid ambiguities created by these issues, database table names are formed using the table description name, the product type, and the product number, e.g.
+`brd_countdata-basic-DP1.10003.001`, rather than the description component alone, `brd_countdata`.
+
+RStudio users can bring up an interactive list of available tables in the RStudio "Connections" pane using the function `neon_pane()` [Fig #].
 
@@
 The default behavior is to rely on a user-specific location appropriate for the operating system, e.g.
-`~/.local/share/neonstore` on Linux, `C:\Users\<username>\AppData\Local\neonstore` on Windows, or `~/Library/Application Support/neonstore` on Mac OS X, as determined by the `rappdirs` R package which provides appropriate modern defaults [@rappdirs].
+`~/.local/share/R/neonstore` on Linux, `C:\Users\<username>\AppData\Local\R\neonstore` on Windows, or `~/Library/Application Support/R/neonstore` on Mac OS X, as determined by `tools::R_user_dir()` which provides appropriate modern defaults [see @rappdirs].
 This approach works well for a single user setup such as a laptop, where a user can share the store across multiple projects in different directories.
-On a server system that is accessed by multiple users, an lower-level location that permits all users to read and write files to the store may be preferable.
+On a server system that is accessed by multiple users, a shared location that permits all users to read and write files to the store may be preferable.
 Because `neonstore` only writes raw NEON data to the store, there is no risk of overwriting work of other users, while the shared storage can reduce consumption of redundant download and storage.
@@
 However, using the environmental variable instead results in more portable code, by allowing other users to set their own storage locations independently instead of hard-wiring it into the R code.
+
+The local `neonstore` database is written in the same user and OS specific location by default, but is configured using a separate environmental variable, `NEONSTORE_DB`, or optional argument to appropriate functions,  `db_dir`.
+Because simultaneous write-access is not supported by the default high-performance database engine, `duckdb`, in multi-user environments it may be advantageous to share a location for raw data files by setting `NEONSTORE_HOME` while allowing users to create individual local databases in their home directories by leaving `NEONSTORE_DB` unset.
+Alternately, users can share a database as well by setting `NEONSTORE_DB` to a shared location, and adopting workflows which do ensure multiple users do not simultaneously call `neon_store()` to update the database.
 
@@
 Then for each site, for each month for which that site collected data, the function will make an API call to the `/data` endpoint (using the `neonstore` function, `neon_data()`), which returns a table of files found for that product at the requested site and month range.
-The  table from `neon_data()` indicates 4 values: the file name, the MD5 hash of the data file (confusingly labeled as the crc32 sum, though CRC32 hashes are used only for AOP data products.
-Note that MD5 hashes are length-32 character strings, while the CRC32 hashes are 8 characters long), the file size (in bytes -- encoded as a character because some file sizes are too long to be represented as an integer), and a download URL.
+The  table from `neon_data()` indicates 4 values: the file name, the MD5 or CRC32 hash of the data file (CRC32 hashes are used only for AOP data products), the file size (in bytes -- encoded as a character because some file sizes are too long to be represented as an integer), and a download URL.
 The download URL embeds a temporary access token which expires after an hour, so users must complete all the downloads before this time expires.
@@
 For the most part, this issue is invisible to `neonstore` users, as `neon_download()` calls automatically pass all download URLs to the download step.
+
+`neon_download()` handles the following tasks:
+
+- `neonstore` aims to provide persistent storage, writing raw data files to
+  the appropriate app directory for your operating system (see `rappdirs`,
+  [Ratnakumar et al 2016](https://CRAN.R-project.org/package=rappdirs)), now provided
+  by the base R function `tools::R_user_dir()`.  Files are organized into appropriate 
+  subdirectories by product, site, and year-month of sampling.  
+- `neon_download()` provides  clean and concise progress bars for the two key
+  processes involved: querying the API to obtain download URLs (which involves no 
+  large data transfer but counts against API rate limiting, see below), and the
+  actual file downloads.
+- `neon_download()` will verify the integrity of file downloads against the MD5 or CRC32
+  hashes provided. 
+- `neon_download()` will omit downloads of any existing data files in the local 
+  store (even if the timestamp portion of the filename has changed without updating the file)
+- You can request multiple products at once using vector notation, though API
+  rate limiting may interfere with large requests.
+- `neon_download()` uses `curl::curl_download()` instead of `downloadr` package
+  used in `neonUtilities`, which can be finicky on Windows and older versions of R. 
+- Automatic handling NEON's rate limiting policy.
+To do, `neon_download()` must  pause 100 seconds every 150 requests if an API token is not supplied or every 950 requests if a token is available.
+Should a rate-limit error be detected, (possible if the same token is used in parallel requests), `neon_download()` will also pause for 100 seconds before retrying the failed request.
+Requests and pauses are communicated through the user messaging interface.
+- Expanding compressed HDF5 files (`*h5.gz`).
+(Note: unusually, NEON chooses to compress binary h5 files with gzip, even though gzip compression is built internally into the H5 specification at configurable compression levels.
+NEON does not choose to compress text-based formats such as .csv, where compression has larger gains.
+Compression has implications for version management, where `neonstore` must track the hashes of compressed files to avoid duplicate downloads)
+- Updating the local manifest in LMDB with the individual checksum and current release tag for each downloaded object (see Provenance)
 
@@
                site = "YELL")
-
- ## Advanced use: filter for a particular table in the product
- neon_download(product = "DP1.10003.001",
-               start_date = "2018-01-01",
-               end_date = "2019-01-01",
-               site = "YELL",
-               file_regex = ".*brd_countdata.*\\.csv")
 ```
@@
 Even adding a handful of products can generate thousands or hundreds of thousands of files in a local store.
-Surprisingly, these can even exceed the capacity of common bash commands like `ls *.csv`.
-While a directory structure could mitigate this issue, `neonstore` sidesteps this issue through a convenient R-based interface.
+Surprisingly, these can even exceed the capacity of common `bash` shell utilities like `ls`.
+`neonstore` sidesteps this issue through a convenient R-based interface.
 The function `neon_index()` will list all recognized files found in the local store as a `data.frame`.
@@
 Because most observational and instrument-based products are divided by site and month, this is still not a convenient way to browse the data.
-The helper function, `neon_store()`, summarizes products that all share the same file description and can thus be easily stacked.
-Like `neon_index()`, optional arguments can filter the results to specific products, dates, and sites.
-
-```{r}
-neon_store()
-```
-
-In this example, we recognize several tables corresponding to the NEON bird survey, product `DP01.10003.001` (all prefixed with `brd_`).  We can read in all the individual files
 

### 2021-12-21 paper/paper.Rmd
:pen:
support lazy-table

@@
 
-The National Ecological Observatory Network (NEON) represents an important resource and a major investment in data collection infrastructure needed to address many of the grand challenges facing envionmental science [@NAS_NEON; @NAS_GRAND].
+The National Ecological Observatory Network (NEON) represents an important resource and a major investment in data collection infrastructure needed to address many of the grand challenges facing environmental science [@NAS_NEON; @NAS_GRAND].
 
@@
 A focus on provenance means that analyses can always be traced back to individual NEON data files, and supports both workflows that require access to the most recent or updated data and workflows that require repeated access to persistent, stable data files.
-A focus on performance reflects the potentially very large size of its aggregated data products, which can easily exceed available RAM on most machines (e.g.
-10 GB - 10 TB range, depending on products and sites considered).
+A focus on performance reflects the potentially very large size of its aggregated data products, which can easily exceed available working memory (RAM) on most machines: as many unified tables of even single products over all available sites and dates can range from 10 GBs into the terrabyte range.
+Because enabling analyses that could span the full spatial and temporal range of NEON's 81 sites over a 30 year horizon is a foundational motivation for NEON research, it essential that researchers be able to easily work with data across this full span.
+However, doing so requires techniques that go beyond workflows which expect data to be re-downloaded (provenance) or stored in memory for analysis (performance).
+
+To address these challenges, `neonstore` implements a high-performance, provenance aware workflow by drawing on a host of key technologies, including cryptographic hashes, a Lightning Memory Mapped Database (LMDB) for metadata management, a columnar-oriented integrated relational database (duckdb) for data management, and optional support for remote access through parquet files cloud-native object stores to facilitate sharing large databases generated by `neonstore`.
+These technologies are seamlessly integrated behind a user-friendly R interface which requires no expertise in the underlying backend technology beyond a familiarity with R and the "tidy data" model used in the popular R packages, `dplyr` and `tidyr`.
 
@@
 When new or revised data significantly alter results, it is also important to be able to pinpoint precisely what has changed and why.
-These elements creates five distinct tasks for managing NEON data provenance: (1) The ability to determine which of each of the individual raw data files were used in an analysis (2) The ability to append new data as it is collected and released (3) the ability to transparently replace existing data when corrections are published, (4) the ability to avoid re-downloading or re-stacking data that has been previously processed, (5) the ability to freeze the data used in a given analysis, so that results can reliably be repeated without accidentally including revised or newer data products.
+These elements creates five distinct tasks for managing NEON data provenance:
+(1) The ability to determine which of each of the individual raw data files were used in an analysis 
+(2) The ability to append new data as it is collected and released 
+(3) the ability to transparently replace existing data when corrections are published, 
+(4) the ability to avoid re-downloading or re-stacking data that has been previously processed, 
+(5) the ability to freeze the data used in a given analysis, so that results can reliably be repeated without accidentally including revised or newer data products.
 
@@
 
-RStudio users can bring up an interactive list of available tables in the RStudio "Connections" pane using the function `neon_pane()` [Fig #].
+RStudio users can bring up an interactive list of available tables in the RStudio "Connections" pane using the function `neon_pane()` [Fig #1].
 
@@
 library(neonstore)
+library(dplyr)
 ```
@@
 
-The workhorse function of `neonstore` is `neon_download()`, which downloads all files identified by a given product code, optionally constrained to specific site, date range, or file name pattern.
-For instance, we can download all of the NEON landbird survey data, identified as NEON data product `DP1.10003.001`: (The three-digit code identifies this as a Level 1 data product (DP1), product number 10003, revision 001, indicating this is the first and so far the only protocol used for bird survey sampling).
+One of the most important functions of `neonstore` is `neon_download()`, which downloads all files identified by a given product code, optionally constrained to specific site, date range, or file name pattern.
+For instance, we can download all of the NEON landbird survey data, identified as NEON data product `DP1.10003.001`: 
+(The three-digit code identifies this as a Level 1 data product (DP1), product number 10003, revision 001, indicating this is the first and so far the only protocol used for bird survey sampling).
 
@@
 
+Note that if we immediately attempt to re-download a subset of this data, `neonstore` detects that we have precisely matching files already downloaded to our local store:
+
 ```{r}
@@
 Even adding a handful of products can generate thousands or hundreds of thousands of files in a local store.
-Surprisingly, these can even exceed the capacity of common `bash` shell utilities like `ls`.
-`neonstore` sidesteps this issue through a convenient R-based interface.
+Surprisingly, these can even exceed the capacity of common `bash` shell utilities like `ls`.  `neonstore` sidesteps this issue through a convenient R-based interface.
 The function `neon_index()` will list all recognized files found in the local store as a `data.frame`.
@@
 
+Note that the index also contains important provenance information, including the checksum (MD5 or CRC32), timestamp, and release tag indicated by NEON.
+Internally, `neonstore` uses checksums to track the provenance of all files downloaded, which allows it to handle both the cases when NEON changes the name of a file but not the contents, and when NEON changes the contents of a file, even if the name remains unchanged.
 `neon_index()` takes optional arguments such as `product`, `start_date`, `end_date`, and `site` to allow for convenient sub-setting.
-Because most observational and instrument-based products are divided by site and month, this is still not a convenient way to browse the data.
+
+**Stacking datasets**
+
+The highly atomized file structure breaks tables down by site and year/month as individual csv files (and other formats, including h5).  
+Typical workflows will treat a collection of such files as a single data table, with columns indicating the site, year, and month of sampling.
+When combining the collection of files into a single table (a process sometimes known as stacking), `neonstore` must handle several special cases:
+
+1) Not all files have identical columns in identical order across all sites and months. This can break simple, high-performance file readers like `vroom` and `arrow`.
+2) The data types of columns (string, numeric, date, timestamp, etc) must be inferred correctly. 
+   This can be difficult when certain columns within individual files have no non-missing (`NA`) values.
+3) Additional columns must be added to capture information (metadata) that is only available in the file name, such as siteID and position (for sensor-based products).
+4) Most importantly, reading must be provenance-aware -- if there are two files with matching product, table, site, year-month and position (for sensor-based products) must resolve which file should be read in based on the timestamp provided, or the requested version.
+
+`neonstore`'s stacking functions handle all of this for us.
+When a version of the same data is detected with a revised timestamp compared to an earlier download, `neonstore` will notify us of the detection, but default to reading the most recent available version.
+
+`neonstore` provides two alternative mechanisms for stacking data.
+The simplest approach is `neon_read()`, which stacks all matching files into a single, in-memory `data.frame`.
+Note that `neon_read()` can be given arguments to filter by site, date range, etc to read in a particular selection of data more quickly.
+`neon_read()` always reads directly from the original, raw data files downloaded to the local neonstore.
 
 ```{r}
-neon_read("brd_countdata-expanded")
+brd_countdata <- neon_read("brd_countdata-expanded")
 ```
 
-Two other functions access additional API endpoints 
-that may also be of interest. `neon_sites()` returns a `data.frame`
-of site information, including site descriptions and 
-the ecological domain that each site falls into:
+For very large tables such as sensor data collections spanning many sites and years, users will require the increased performance provided by the high-performance, columnar-oriented relational database, `duckdb` [@duckdb].
+To take advantage of this approach, we must first store the contents of the individual data files in the relational database using the function `neon_store()`:
 
 ```{r}
-neon_sites()
+neon_store("brd_countdata")
 ```
 
-Lastly, `neon_products()` returns a table with a list of all neon products,
-which may be useful for data discovery or additional metadata about any
-given product:
+This task needs to be done only once for each table or product after any new files have been downloaded.
+Thanks to provenance tracking, any repeated calls to this function will not result in additional imports.
+If a previously imported file has been deprecated by the publication of a matching file with newer timestamp,
+`neon_store()` will remove the deprecated data and import the new data, with a warning saying so.
+
+Once data has been imported into the local database, we can quickly load the table into R with `neon_table()`
 
 ```{r}
-neon_products()
+brd_countdata <- neon_table("brd_countdata")
 ```
 
-Note that at this time, `neonstore` does not provide a complete
+This will be somewhat faster than `neon_read()`, since there is no need to parse the original data.
+By default, this will still read all the data into a normal in-memory `data.frame` in R.
+When working with very large data tables, it may be preferable or necessary to work with data
+directly in duckdb, which can perform most `dplyr` and `tidyr` operations, such as `group_by()`, `summarise`, 
+`filter`, and so forth, within the database.
 
-## Design Details / comparison to `neonUtilities`
-
-`neonstore` is not meant as a replacement to the `neonUtilities` package
-developed by NEON staff.  `neonUtilities` performs a range of product-specific
-data querying, parsing, and data manipulation beyond what is provided by NEON's
-API or web interface. `neonUtilities` also provides other utilities for working 
-with NEON data beyond the scope of the NEON API or the data download/ingest 
-process. While this processing is undoubtedly useful, it may make
-it difficult to compare results or analyses based on data downloaded and accessed
-using `neonUtilities` R package with analyses based on data accessed directly
-from the web interface, the API, or another tool (or even a different release
-of the `neonUtilities`).
-
-By contrast, `neonstore` aims to do far less.  `neonstore` merely automates the
-download of individual NEON data files.  In contrast to `neonUtilities` which 
-by default "stacks" these raw files into single tables and discards the raw 
-data, `neonstore` preserves only the raw files in the store, stacking the 
-individual tables "on demand" using `neon_read()`. `neon_read()` is a thin
-wrapper around the `vroom` package, 
-[Hester & Wickham, 2020](https://vroom.r-lib.org), which uses the `altrep` 
-mechanism in R to provide very fast reads of rectangular text data into R,
-and trivially handles the case of a single table being broken across many files.
-Some NEON tables are not entirely consistent in their use of columns across the
-individual site-month files, so `neon_read()` transparently checks for this,
-reading in groups of files sharing all matching columns with `vroom` before
-binding the groups together. This makes it easier to always trace an analysis
-back to the original input data, makes it easier to update input data files 
-without facing the challenge of either downloading & stacking the whole 
-data product from scratch again or having to keep track of some previously
-downloaded data file.
-
-A few other differences are also worth noting.
-
-- `neonstore` aims to provide persistent storage, writing raw data files to
-  the appropriate app directory for your operating system (see `rappdirs`,
-  [Ratnakumar et al 2016](https://CRAN.R-project.org/package=rappdirs)). 
-  More details about this can be found in [Provenance](/#Provenance), below.
-- `neon_download()` provides  clean and concise progress bars for the two key
-  processes involved: querying the API to obtain download URLs (which involves no 
-  large data transfer but counts against API rate limiting, see below), and the
-  actual file downloads.
-- `neon_download()` will verify the integrity of file downloads against the MD5
-  hashes provided. 
-- `neon_download()` will omit downloads of any existing data files in the local 
-  store.  
-- You can request multiple products at once using vector notation, though API
-  rate limiting may interfere with large requests.
-- `neon_download()` uses `curl::curl_download()` instead of `downloadr` package
-  used in `neonUtilities`, which can be finicky on Windows and older versions of R. 
-- `neonstore` has slightly lighter dependencies: only `vroom` and `httr`, and
-  packages already used by one of those two (`curl`, `openssl`).
-
-Like `neonUtilities`, You can optionally include site and date filters,
-e.g. to request only records more  recent than a certain date.  Doing so will
-preserve API quota and improve speed (see API limits, below). `neonUtilities`
-is also far more widely tested and has extensive error handling tailored to
-individual data products.
-
-## Provenance
-
-Because `neonstore` only stores raw data products as returned from the NEON API,
-it can easily determine which files have already been downloaded, and only
-download new files without requiring the user to specify specific dates. 
-(It must still query the API for all the metadata in the requested date range).
-This same modular approach also makes it easy to track _data provenance_, an 
-essential element of reproduciblity in comparing results across other analyses
-of the NEON data.
-
-We can list precisely which component files are being read in by `neon_read()`
-by consulting `neon_index()`:
+**Working with very large data products**
 
 ```{r}
-raw_files <- neon_index(table = "brd_countdata-expanded", hash="md5")
-raw_files
+brd_countdata <- neon_table("brd_countdata", lazy = TRUE)
+brd_countdata
 ```
 
-`neon_read()` is a relatively trivial function that simply passes this file 
-list to `vroom::vroom()`, a fast, vectorized parser that can easily read in 
-a single table that is broken into many separate files.
+Note that R displays what appears to be the same data frame as before, but with two essential differences.
+R indicates that the displayed object is a `duckdb_connection`, and lists the number of rows only as `??`
+(or within RStudio's rmarkdown interface, only the first 1000 rows will be displayed in the interactive table).
+The number of rows is not displayed because this query is executed "lazily." Since the print method needs only
+display the first 6 rows, the database does not execute the potentially expensive operation of counting all the rows
+in the table, it merely displays the first 6. 
+This approach is particularly pwoerful if we wish to first subset or summarise the data in a way that will result
+in an overall smaller table.
 
-Imagine instead that we use the common pattern of downloading
-these raw files, stacks and possibly cleans the data, saving only this derived
-product while discarding the individual files.  Now imagine a second researcher,
-at some later date, queries the API over the same reported range of dates and 
-sites, uses the same software package to stack the tables, only to discover the
-resulting table is somehow different from ours (e.g. by comparing file hashes).
-Pinpointing the source of the discrepancy would be challenging and 
-labor-intensive.
-
-In contrast, the same detective-work would be easy with the `neonstore` file
-list.  We can confirm if the API had returned the same number of 
-raw files with the same names; and better, can verify integrity of the contents
-by comparing hashes of files now being returned to those recorded
-by `neon_index()`.  In this way, we could determine if any additional files
-had been included or pinpoint any files that may have changed.
-
-As such, users might want to store the `neon_index()` `data.frame` for the
-table(s) they have used as part of their analysis, including the individual
-file hashes.  One can also generate a zip of all the data files for 
-archival purposes. (Note that NEON is an Open Data provider, see
-[LICENCE](https://www.neonscience.org/data/about-data/data-policies).)
-
-```{r message=FALSE}
-write.csv(raw_files, "index.csv")
-neon_export("brd_countdata.zip")
-```
-
-## Data citation
-
-Always remember to cite your data sources!
-`neonstore` knows how to generate the appropriate citation for the data
-in your local store (or any specific product).
+Consider the example of a much larger table collected by sensor data; say, single-aspirated temperature collected every 30 mintues across all sites:
+Note that we exploit pattern matching to provide a short-hand to the table name.
+We download and import the 30m sampled table
 
 ```{r}
-## Note: NEON's recommendation may soon change
-neon_citation()
-```
-
-## Note on API limits
-
-[The NEON API now rate-limits requests.](https://data.neonscience.org/data-api/rate-limiting/#api-tokens).
-Using a personal token will increase the number of requests you can make.
-See that link for directions on registering for a token.
-Then pass this token in `.token` argument of `neon_download()`,
-or for frequent use, add this token as  an environmental variable, `NEON_DATA`
-to your local `.Renviron` file in your user's home directory.
-
-`neon_download()` must first query each the API of eacn NEON site which collects
-that product, for each month the product is collected.
-(It would be much more efficient on the NEON server if the API could take
-queries of the from `/data/<product>/<site>`, and pool the results, rather than
-require each month of sampling separately!)
-
-```{r include=FALSE}
-unlink("brd_countdata.zip")
-unlink("index.csv")
-Sys.unsetenv("NEONSTORE_HOME")
-```
-
-## Non-stacking files and low-level interface
-
-At it's core, `neonstore` is simply a mechanism to download files from the NEON API.
-While the `.csv` files from the Observation Systems (OS, e.g. bird count surveys),
-and Instrument Systems (e.g. aquatic sensors) are typically stacked into large
-tables, other products, such as the `.laz` and `.tif` images produced by the
-airborne observation platform LIDAR and cameras may require a different approach.
-
-```{r}
-# Read in a large file list for illustration purposes
-cper_data <- readr::read_csv("https://minio.thelio.carlboettiger.info/shared-data/neon_data_catalog.csv.gz")
-
-## Typically one would read all files in local store, e.g. list.file(neon_dir())
-df <- neon_filename_parser(cper_data$name)
+neon_download(product = "DP1.00002.001", table = "30") #Temp single aspirated
+neon_store(product = "DP1.00002.001", table = "30") #Temp single aspirated
 ```
@@
 ```{r}
-library(dplyr)
-df %>% count(EXT, sort=TRUE)
+temp <- neon_table(product = "DP1.00002.001",  table = "30", lazy=TRUE) #Temp single aspirated
 ```
 
-We can take a look at all `laz` LIDAR files:
+```{r}
+temp %>% 
+  group_by(month) %>%
+  summarise(mean = mean(),
+            max = max(),
+            min = min()) %>% 
+  collect()
+```
 
 ```{r}
-df %>% 
-  filter(EXT == "laz")
+index <- neon_index(product = "DP1.00002.001", table = "30") 
+info <- index$path %>% fs::file_info()
+n_files <- length(index$path)
+n_bytes <- sum(info$size)
 ```
 
-Note that many of the airborne observation platform (AOP) products, such as
-these LIDAR files, do not include the PRNUM or REV components that make up part
-of the `productCode`s used in the NEON `product` tables.
+## Dependencies and optional dependencies
+
+`neonstore`'s core dependencies are quite light, avoiding hard requirements on other packages which themselves bring in a large recursive list of dependencies.
+`DBI` [@DBI], `duckdb` [@duckdb], and `thor` [@thor] provide the core database interfaces; all have no recursive dependencies.
+`httr` [@httr] provides functionality required to interface with the NEON REST API, and `R.utils`, `zip`, and `vroom` [@vroom] provide parsers required to unpack the data.
+`progress` provides user-friendly progress bars and estimated run-times for slow functions.
+
+Optional dependencies are required for specific tasks. `dplyr` is needed for lazy read in `neon_table()`.
+`neonUtilities` [@neonUtilities] and `rhdf5` are required at this time for reading in eddy-covariance products.
+`arrow` [@arrow] is required for remote database connections using `neon_remote()`.
+`rstudioapi` is required to utilize `neon_pane()` to view the connections pane from within RStudio.
+The remaining suggested dependencies are utilized within testing only.
+
+## Appendix
+
+Discuss the NEON filename convention parser, link to `neon_filename_parser.R` source for regex patterns.  
+Discuss cases of NEON changing filename conventions.

### 2021-12-21 R/neon_table.R

@@ -6,17 +6,35 @@
 #' We cannot filter on start_date or end_date since these
 #' come only from the filename metadata and are only added
 #' to instrument tables, not observation tables etc.
-#' 
+#' @param lazy logical, default FALSE. Should we return a remote dplyr 
+#' connection to the table in duckdb? This can substantially improve 
+#' performance and avoid out-of-memory errors when working with very large
+#' tables. However, not all R operations can be performed on a remote table,
+#' only (most) functions from `dplyr` and `tidyr`, as these can be 
+#' translated automatically to SQL language used by the remote database.
+#' Use `dplyr` functions like [dplyr::filter()], [dplyr::group_by()], and 
+#' [dplyr::summarise()] to subset
+#' the data appropriately within the remote table before calling
+#'  `[dplyr::collect()]` to import the data fully into R.
 #' @export
 #' @importFrom DBI dbGetQuery
 #' 
 neon_table <- function(table,
                        site = NA,
-                       db = neon_db()){
+                       db = neon_db(),
+                       lazy = FALSE){
 
   con <- db
   tables <- DBI::dbListTables(con)
   table <- check_tablename(table, tables)
+  
+  
+  if(lazy){
+    if(!requireNamespace("dplyr", quietly = TRUE)){
+      stop("dplyr is required for lazy evaluation")
+    }
+    return(dplyr::tbl(db, table))
+  }
   
   where <- NULL
   query <- paste0("SELECT * FROM \"", table, "\"")

### 2021-12-21 inst/examples/thelio_mirror.R

@@ -40,9 +40,9 @@ neon_download("DP1.20217.001", site =  aq_sites) # Groundwater temperature
 neon_download("DP1.20033.001", site =  aq_sites) # Nitrate in surface water
 
 ## Terrestrial meteorology
-neon_download(product = "DP1.00002.001", site = ter_sites, type = "basic", table = "30") #
+neon_download(product = "DP1.00002.001", site = ter_sites, type = "basic", table = "30") #Temp single aspirated
 neon_download(product = "DP1.00006.001", site = ter_sites, type = "basic", table = "30") #
-neon_download(product = "DP1.00007.001", site = ter_sites, type = "basic") #
+neon_download(product = "DP1.00007.001", site = ter_sites, type = "basic") # Precip, thoughfall
 neon_download(product = "DP1.00023.001", site = ter_sites, type = "basic") #
 neon_download(product = "DP2.00024.001", site = ter_sites, type = "basic") #
 neon_download(product = "DP1.00100.001", site = ter_sites, type = "basic") #

### 2021-12-22 paper/paper.Rmd
:pen:

@@
 library(neonstore)
-Sys.setenv("NEONSTORE_HOME" = tempfile())
+#Sys.setenv("NEONSTORE_HOME" = tempfile())
 ```
@@
 The National Ecological Observatory Network (NEON) represents an important resource and a major investment in data collection infrastructure needed to address many of the grand challenges facing environmental science [@NAS_NEON; @NAS_GRAND].
-
 NEON provides access to 181 (active and planned) ecological data products at 81 (permanent and relocatable) sites throughout the United States.
@@
 Here, we describe the design, use, and rationale for the package.
+We illustrate a few example use cases, including downloading the 30 minute single-aspirated temperature data across all available NEON sites and dates, a collection of over 13 thousand individual files totalling nearly 2 gigabytes of data.
+We demonstrate how `neonstore` facilitates summary calculations such as computing monthly means across all sites in a fraction of a second with minimal memory requirements.
+The approaches taken in `neonstore` maximize reproducibility, facilitate collaboration including across computer languages, and help making analyses which leverage the full spatial and temporal scale of NEON easily available to researchers with modest computational resources and expertise.
 
@@
 Note that we exploit pattern matching to provide a short-hand to the table name.
-We download and import the 30m sampled table
+We download and import the 30m sampled table. 
+The initial download is not a fast operation - in our example, rate-limiting rules by NEON make the command take over 20 minutes to complete the API requests that would otherwise take only 3 minutes.
+The download itself is determined more by network speed. 
+In our example this takes about 1 hour to download over 13 thousand individual files.
+Because NEON uses download links that expire after 1 hour, it may be necessary to repeat the download command after more than an hour expires,
+(alternatively, divide the download into multiple batches by years or sites).
+The `neon_download()` utility can pick up from where it left off.
+
+`neon_store()` import is also a slow, one-time operation to read in all of these files into the database, depending on disk and precessor speeds. 
+In our example, this takes about 20 minutes.
+Note that it is not necessary to repeat these commands later except to import newly released data, which will be added to the persistent store.
 
 ```{r}
-neon_download(product = "DP1.00002.001", table = "30") #Temp single aspirated
-neon_store(product = "DP1.00002.001", table = "30") #Temp single aspirated
+neon_download(product = "DP1.00002.001", table = "30min") #Temp single aspirated
+neon_store(product = "DP1.00002.001", table = "30min") #Temp single aspirated
 ```
 
+With the data imported into the database, `neon_table()` can quickly connect to the remote database.
+
 ```{r}
-temp <- neon_table(product = "DP1.00002.001",  table = "30", lazy=TRUE) #Temp single aspirated
+temp <- neon_table(product = "DP1.00002.001",  table = "30min", lazy=TRUE) #Temp single aspirated
 ```
 
+This is a lot of data! Over 19 million rows:
+
 ```{r}
-temp %>% 
-  group_by(month) %>%
-  summarise(mean = mean(),
-            max = max(),
-            min = min()) %>% 
-  collect()
+temp %>% count()
 ```
 
+However, working with the remote duckdb connection, we can compute summary statistics, such as monthly averages for each site, takes only milliseconds:
+
 ```{r}
-index <- neon_index(product = "DP1.00002.001", table = "30") 
+bench::mark({
+temp_monthly <- temp %>%
+  mutate(month = month(startDateTime)) %>%
+  group_by(siteID, month) %>%
+  summarise(mean_temp = mean(tempSingleMean, na.rm=TRUE),
+            max = max(tempSingleMean, na.rm=TRUE),
+            min = min(tempSingleMean, na.rm=TRUE),
+            .groups = "drop") %>% 
+  collect()
+})
+```
+Not only is this about 10x faster than if we had read the table entirely into R first, but this also executes with minimal memory footprint (317Kb) instead of the over 2.6 GB required to perform the same operation in R.
+Having summarized the data using our desired statistics, the `collect()` command has returned the result as a regular in-memory `data.frame` which we can use with other R functions, such as data visualization tools:
+
+```{r}
+library(ggplot2)
+library(lubridate)
+temp_monthly %>% mutate(month = lubridate::month(month, label=TRUE)) %>%
+  filter(siteID %in% c("HARV", "ORNL")) %>%
+  ggplot(aes(month, mean_temp, col=siteID, ymin = min, ymax = max, fill = siteID)) +
+  geom_linerange(lwd=4, alpha=.5) +
+  geom_point(size=8, shape="_")
+```
+
+Similarly, it is straight forward to use most other `dplyr` operations, such as the joins, to combine data across multiple tables.
+The promise of NEON is precisely this ability to work across a large spatial and temporal range of data collected in standardized fashion.
+
+```{r include=FALSE}
+temp %>% select(verticalPosition, horizontalPosition) %>% distinct()
+
+index <- neon_index(product = "DP1.00002.001", table = "30min") 
 info <- index$path %>% fs::file_info()
@@
 ```
+
+**Network-based access**.
+
+In R, we refer to the `duckdb` database as a "remote" connection because the operations are handled outside of R, by `duckdb`.
+Some database connections, like MySQL or Postgres, require the user to install a separate piece of software, often on an external server,
+which we then access over the network connection (with appropriate security credentials), though it can also be run on the same local machine as the 'client' software.
+This client-server approach can be very powerful, but requires more knowledge and resources to deploy.
+In contrast, `duckdb` installs automatically as a stand-alone R package.
+
+As illustrated above, the slowest part of the process by far are the one-time steps of initial download of over 10,000 individual files, and the importing of those files into the database.
+Further, the storage requirements for users working with NEON data spanning many products, sites, and dates can be restrictive. 
+As such, it may be desirable for a research team working on such an analysis to establish a single shared `neonstore` storage center rather than expect that all users will download, import, and maintain their own local stores.
+To facilitate this process, `neonstore` supports access to remote network-based storage over cloud-native storage protocols with the help of  Apache `arrow` [@arrow].
+First, using `neonstore`, we download and import the data as above.  
+Our next step is export this database to a parquet snapshot using `neon_export_db()` and upload that data to cloud native storage system such as Amazon Web Services (AWS) S3 bucket, or, in our case, a self-hosted MINIO [@minio] instance.
+This is much cheaper and raises less security concerns than maintaining a server, as the remote service is providing only networked storage, not general-purpose computing power.
+
+We pass the connection details to `neon_remote_db`.  
+In this case, we use a public (read-only) bucket, though other users may prefer to configure access that requires a secret key.
+
+```{r}
+con <- neon_remote_db(host = "minio.thelio.carlboettiger.info",
+                      bucket = "shared-data",
+                      path = "neonstoredb")
+```
+
+Once a connection to the remote database is established, we can request individual tables using `neon_remote` to access the data using dplyr functions.  
+Like `neon_table()`, we must use `collect()` to draw the data into R if we wish to perform advanced R-specific operations that are not supported by the backend engine.
+
+```{r}
+remote_temp <- neon_remote("30min", product = "DP1.00002.001", db = con)
+remote_temp %>% count(siteID) %>% collect()
+```
+
+Because this approach relies on operations over a network connection, it will generally be bandwidth limited unless the bucket is located in the same region as compute operations (e.g.
+in a regional cloud server).
+Users may also download the `parquet` exported files directly and import them into a local `neonstore` instance (see `neon_import_db()`) to share neonstore snapshots without requiring independent downloads.
+Using `neon_db_export()` to generate snapshots is also a reliable way to preserve and version your analyses for future that does not depend on access to the NEON API or even to `neonstore`. 
+This approach should work equally well for data exchange with collaborators not using R.
+`parquet` is a widely used data interchange format with support in most major languages, including python, Java and C++.
+Similarly, `duckdb` is a cross-platform tool that can be accessed not only from R, but from python, Java, and many other languages, including web assembly (for running in the browser) or run using a stand-alone CLI [@duckdb].
 

### 2021-12-22 R/neon_remote.R

@@ -52,7 +52,7 @@ neon_remote_db <- function(host = "minio.thelio.carlboettiger.info",
 #' neon_remote 
 #'  
 #' select a table from the remote connection
-#' @param tbl table name (pattern match regex)
+#' @param table table name (pattern match regex)
 #' @param product product code
 #' @param type basic or extended (if necessary to distinguish)
 #' @param db a [neon_remote_db] connection.  If not provided, one will be created,
@@ -65,9 +65,9 @@ neon_remote_db <- function(host = "minio.thelio.carlboettiger.info",
 #' grouped and summarised, all without ever downloading the whole table.
 #' Use [dplyr::collect()] to download the (possibly filtered) table into
 #' and pull into memory.  
-neon_remote <- function(tbl = "", product = "", type = "", db = neon_remote_db()){
+neon_remote <- function(table = "", product = "", type = "", db = neon_remote_db()){
   labels <- names(db)
-  i <- grepl(tbl, labels) & grepl(product, labels)  & grepl(type, labels)
+  i <- grepl(table, labels) & grepl(product, labels)  & grepl(type, labels)
   if(sum(i) > 1){
     message(paste0("multiple matches found\n", 
                   "returning list with tables:\n",

### 2021-12-22 R/neon_table.R

@@ -20,13 +20,18 @@
 #' @importFrom DBI dbGetQuery
 #' 
 neon_table <- function(table,
+                       product = NA,
+                       type = NA,
                        site = NA,
                        db = neon_db(),
                        lazy = FALSE){
 
   con <- db
-  tables <- DBI::dbListTables(con)
-  table <- check_tablename(table, tables)
+
+  table <- check_tablename(table, 
+                           product = product,
+                           type = type,
+                           tables = DBI::dbListTables(con))
   
   
   if(lazy){
@@ -58,8 +63,19 @@ neon_table <- function(table,
 
 
 ## Sanitize table names, particularly extended/basic matching
-check_tablename <- function(x, tables){
+check_tablename <- function(x, product = NA, type = NA, tables){
  out <- tables[grepl(x, tables)]
+ 
+ ## Filter on product & type if requested
+ if (!is.na(product)) {
+   out <- out[grepl(product, out)]
+ }
+
+ if (!is.na(type)) {
+   out <- out[grepl(type, out)]
+ }
+ 
+  
  if(length(out) > 1){
    stop(paste("multiple matches for table", 
               x, ":", out), call. = FALSE)

### 2022-01-07 R/neon_stack.R
hide show_col_types

@@ -9,7 +9,9 @@ neon_stack <- function(files,
                        progress = TRUE,
                        vroom_progress = FALSE,
                        ...){
-  
+
+
+    
   ## Stack H5 Files (eddy-covariance only)
   if(any(grepl("[.]h5$", files))){
     stack_eddy(files, progress = progress, ...)
@@ -121,7 +123,11 @@ vroom_many <- function(files,
 
 
 ## Apply vroom over files that share a common schema.
-vroom_ragged <- function(files, altrep = FALSE, vroom_progress = FALSE, ...){
+vroom_ragged <- function(files, 
+                         altrep = FALSE,
+                         vroom_progress = FALSE, 
+                         show_col_types = FALSE,
+                         ...){
   
   ## We read the 1st line of every file to determine schema  
     schema <- lapply(files, 

### 2022-01-07 paper/paper.Rmd

@@
 date: "`r Sys.Date()`"
-bibliography: refs.bib
+#bibliography: refs.bib
 layout: 3p
@@
 
+```{r include=FALSE}
+Sys.unsetenv("AWS_DEFAULT_REGION")
+Sys.unsetenv("AWS_S3_ENDPOINT")
+Sys.unsetenv("AWS_ACCESS_KEY_ID")
+Sys.unsetenv("AWS_SECRET_ACCESS_KEY")
+Sys.setenv("AWS_EC2_METADATA_DISABLED"="TRUE")
+```
+
 ```{r}

### 2022-01-10 paper/paper.Rmd
:pen:

@@
     affiliation: ucb
-  - name: "Quinn Thomas"
+  - name: "R. Quinn Thomas"
     affiliation: vt
@@
 date: "`r Sys.Date()`"
-#bibliography: refs.bib
+bibliography: refs.bib
 layout: 3p
@@
   comment = "#>",
-  fig.path = "man/figures/README-",
   out.width = "100%",
@@
 library(neonstore)
-#Sys.setenv("NEONSTORE_HOME" = tempfile())
+Sys.setenv("NEONSTORE_HOME" = "store")
 ```
 
-The National Ecological Observatory Network (NEON) represents an important resource and a major investment in data collection infrastructure needed to address many of the grand challenges facing environmental science [@NAS_NEON; @NAS_GRAND].
+The National Ecological Observatory Network (NEON) represents an important resource and a major investment in data collection infrastructure needed to address many of the grand challenges facing environmental science [@NAS_NEON; @Balch2020; @Nagy2021].
 NEON provides access to 181 (active and planned) ecological data products at 81 (permanent and relocatable) sites throughout the United States.
@@
 This makes it possible to both eliminate the requirement of separately installing and running a database server while also opening the door to substantial performance improvements beyond the reach of traditional relational database clients.
-`duckdb` is a leading example of such a columnar database [@duckdb], with seamless support for R integration and superior performance benchmarks to both traditional relational databases and in-memory operations in R using `dplyr`.
+`duckdb` is a leading example of such a columnar database [@Raasveldt_2019; @duckdb], with seamless support for R integration and superior performance benchmarks to both traditional relational databases and in-memory operations in R using `dplyr`.
 
@@
 `brd_countdata-basic-DP1.10003.001`, rather than the description component alone, `brd_countdata`.
+RStudio users can bring up an interactive list of available tables in the RStudio "Connections" pane using the function `neon_pane()` [Fig #1].
 
-RStudio users can bring up an interactive list of available tables in the RStudio "Connections" pane using the function `neon_pane()` [Fig #1].
+![RStudio Connections pane showing interactive panel listing all tables imported into the local duckdb database.](rstudio-pane.png)
+
+<!--
 
@@
 
+-->
+
 `neonstore` seeks to help users download and manage NEON files locally in a way that lets analyses be quickly traced back to the raw data files.
-
 `neonstore` is distinguished as much by what it does not do as by what it does.
@@
   the appropriate app directory for your operating system (see `rappdirs`,
-  [Ratnakumar et al 2016](https://CRAN.R-project.org/package=rappdirs)), now provided
-  by the base R function `tools::R_user_dir()`.  Files are organized into appropriate 
-  subdirectories by product, site, and year-month of sampling.  
+  [Ratnakumar et al 2016](https://CRAN.R-project.org/package=rappdirs)), now provided by the base R function `tools::R_user_dir()`.
+  Files are organized into appropriate subdirectories by product, site, and year-month of sampling.
 - `neon_download()` provides  clean and concise progress bars for the two key
@@
 - Automatic handling NEON's rate limiting policy.
-To do, `neon_download()` must  pause 100 seconds every 150 requests if an API token is not supplied or every 950 requests if a token is available.
+  To do this, `neon_download()` must  pause 100 seconds every 150 requests if an API token is not supplied or every 950 requests if a token is available.
   Should a rate-limit error be detected, (possible if the same token is used in parallel requests), `neon_download()` will also pause for 100 seconds before retrying the failed request.
@@
 Even adding a handful of products can generate thousands or hundreds of thousands of files in a local store.
-Surprisingly, these can even exceed the capacity of common `bash` shell utilities like `ls`.  `neonstore` sidesteps this issue through a convenient R-based interface.
+Surprisingly, these can even exceed the capacity of common `bash` shell utilities like `ls`.
+`neonstore` sidesteps this issue through a convenient R-based interface.
 The function `neon_index()` will list all recognized files found in the local store as a `data.frame`.
@@
 
-For very large tables such as sensor data collections spanning many sites and years, users will require the increased performance provided by the high-performance, columnar-oriented relational database, `duckdb` [@duckdb].
+For very large tables such as sensor data collections spanning many sites and years, users will require the increased performance provided by the high-performance, columnar-oriented relational database, `duckdb` [@Raasveldt_2019; @duckdb].
 To take advantage of this approach, we must first store the contents of the individual data files in the relational database using the function `neon_store()`:
@@
 
-```{r}
+```{r download_temp}
 neon_download(product = "DP1.00002.001", table = "30min") #Temp single aspirated
@@
 `neonstore`'s core dependencies are quite light, avoiding hard requirements on other packages which themselves bring in a large recursive list of dependencies.
-`DBI` [@DBI], `duckdb` [@duckdb], and `thor` [@thor] provide the core database interfaces; all have no recursive dependencies.
+`DBI` [@dbi], `duckdb` [@duckdb], and `thor` [@thor] provide the core database interfaces; all have no recursive dependencies.
 `httr` [@httr] provides functionality required to interface with the NEON REST API, and `R.utils`, `zip`, and `vroom` [@vroom] provide parsers required to unpack the data.
 `progress` provides user-friendly progress bars and estimated run-times for slow functions.
-
 Optional dependencies are required for specific tasks. `dplyr` is needed for lazy read in `neon_table()`.
@@
 
-## Appendix
+## Conclusions
 
-Discuss the NEON filename convention parser, link to `neon_filename_parser.R` source for regex patterns.  
-Discuss cases of NEON changing filename conventions.
+`neonstore` provides a high-performance and provenance-aware mechanism for accessing, maintaining, and working with NEON data products, designed to scale seamlessly to analyses which stretch across the full spatial and temporal range which makes this nation-wide dataset so uniquely valuable.
+The `neonstore` package follows simple and robust best-practices for data management that can work across most popular computing languages; such as preserving a local library of the original raw data files downloaded from NEON, and importing data into a local relational database (`duckdb`) that can be accessed from a wide range of libraries.
+Additional utilities enable serverless network-based sharing of a large `neonstore` database across a team through the use of `parquet` and `arrow`'s remote filesystems (which likewise supports most major computing languages.)
+`neonstore` has been co-developed by NEON staff (CL & CL) and members of the Ecological Forecasting Initiative (CB & RQT), tested in supporting the EFI NEON forecasting challenge [@efi_challenge], among other use cases.
+
+# References

### 2022-01-11 README.Rmd
WIP

@@
 [![R build status](https://github.com/cboettig/neonstore/workflows/R-CMD-check/badge.svg)](https://github.com/cboettig/neonstore/actions)
-[![Codecov test coverage](https://codecov.io/gh/cboettig/neonstore/branch/master/graph/badge.svg)](https://codecov.io/gh/cboettig/neonstore?branch=master)
+[![Codecov test coverage](https://codecov.io/gh/cboettig/neonstore/branch/master/graph/badge.svg)](https://app.codecov.io/gh/cboettig/neonstore?branch=master)
 [![CRAN status](https://www.r-pkg.org/badges/version/neonstore)](https://CRAN.R-project.org/package=neonstore)

### 2022-01-11 paper/paper.Rmd
WIP

@@
 
-```{r, setup, include = FALSE}
+```{r setup, include = FALSE}
 knitr::opts_chunk$set(
@@
   out.width = "100%",
-  cache = FALSE
+  cache = TRUE
 )
 library(neonstore)
+library(dplyr)
+library(ggplot2)
+library(magick)
 Sys.setenv("NEONSTORE_HOME" = "store")
+
+## embed images
+ggimage <- function(path, transparent = FALSE){
+  img <- magick::image_read(path)
+  if(transparent)  img <- magick::image_transparent(img, "white")
+  ggplot2::ggplot() + 
+  ggplot2::annotation_raster(img, 
+                    xmin=-Inf, xmax=Inf, ymin=-Inf, ymax=Inf) +
+    theme(axis.line = element_blank(), 
+          panel.background = element_blank(),
+          axis.ticks = element_blank(), 
+          axis.text = element_blank(), 
+          axis.title = element_blank())
+
+}
 ```
 
-The National Ecological Observatory Network (NEON) represents an important resource and a major investment in data collection infrastructure needed to address many of the grand challenges facing environmental science [@NAS_NEON; @Balch2020; @Nagy2021].
+The National Ecological Observatory Network (NEON) represents an important resource and a major investment in data collection infrastructure needed to address many of the grand challenges facing environmental science [@nap_neon; @Balch2020; @Nagy2021].
 NEON provides access to 181 (active and planned) ecological data products at 81 (permanent and relocatable) sites throughout the United States.
@@
 
-![RStudio Connections pane showing interactive panel listing all tables imported into the local duckdb database.](rstudio-pane.png)
-
+```{r fig1, echo=FALSE, fig.cap="RStudio Connections pane showing interactive panel listing all tables imported into the local duckdb database."}
+ggimage("rstudio-pane.png") 
+```
 <!--
@@
 
-```{r}
+```{r cache=FALSE}
 con <- neon_remote_db(host = "minio.thelio.carlboettiger.info",
@@
 
-```{r}
+```{r  cache=FALSE}
 remote_temp <- neon_remote("30min", product = "DP1.00002.001", db = con)

### 2022-01-11 inst/examples/thelio_mirror.R

@@ -3,12 +3,12 @@ bench::bench_time({
 library(neonstore)
 
   
-  mem_limit <- function(mem_limit = 16, units = "GB"){
+  mem_limit <- function(db = neon_db(), mem_limit = 16, units = "GB"){
     DBI::dbExecute(db, paste0("PRAGMA memory_limit='", mem_limit, " ", units,"'"))
   }
   
-  enable_parallel <- function(duckdb_cores = arrow::cpu_count()){
-    DBI::dbExecute(neon_db(), paste0("PRAGMA threads=", arrow::cpu_count()))
+  enable_parallel <- function(db = neon_db(), duckdb_cores = arrow::cpu_count()){
+    DBI::dbExecute(db, paste0("PRAGMA threads=", arrow::cpu_count()))
   }
   mem_limit()
   enable_parallel()

### 2022-03-28 paper/paper.Rmd
WIP

@@
     affiliation: vt
-  - name: "Christine Laney"
+  - name: "Christine M. Laney"
     affiliation: neon
@@
   out.width = "100%",
-  cache = TRUE
+  cache = TRUE,
+  message = FALSE,
+  warning = FALSE,
+  output = "hide"
 )
@@
 
-# Rationale and package design
-
 Provenance and performance are two central aspects to the design of `neonstore`.
@@
 
-## Provenance
+# Provenance
 
@@
 
-## Performance
+# Performance
 
@@
 `brd_countdata-basic-DP1.10003.001`, rather than the description component alone, `brd_countdata`.
-RStudio users can bring up an interactive list of available tables in the RStudio "Connections" pane using the function `neon_pane()` [Fig #1].
+RStudio users can bring up an interactive list of available tables in the RStudio "Connections" pane using the function `neon_pane()` \ref{fig1}.
 
@@
 
-## `neonstore` Functionality
+# `neonstore` Functionality
 
-`neonstore` functionality can be divided into two halves: functions that interface with the NEON Application Programming Interface (API), and functions for working with a local store of previously-downloaded NEON files.
+The `neonstore` package can be divided into two parts.
+One part works directly with the NEON API to handle tasks of _data discovery_ and _data download_ into a persistent local store, while the other provides three mechanisms to work with data.
 
@@
 
-### A local store for neon files
-
-The location of the local store can be configured depending on the context.
-The default behavior is to rely on a user-specific location appropriate for the operating system, e.g.
-`~/.local/share/R/neonstore` on Linux, `C:\Users\<username>\AppData\Local\R\neonstore` on Windows, or `~/Library/Application Support/R/neonstore` on Mac OS X, as determined by `tools::R_user_dir()` which provides appropriate modern defaults [see @rappdirs].
-This approach works well for a single user setup such as a laptop, where a user can share the store across multiple projects in different directories.
-On a server system that is accessed by multiple users, a shared location that permits all users to read and write files to the store may be preferable.
-Because `neonstore` only writes raw NEON data to the store, there is no risk of overwriting work of other users, while the shared storage can reduce consumption of redundant download and storage.
-In yet other cases, a user may prefer a path that is specific to a single project.
-
-It is of course possible to use a temporary directory such as `tempdir()` for the storage location.
-Such an approach is the default download location of the alternative package, `neonUtilities`, although this can be configured there as well.
-However, doing so defeats the purpose of having a persistent store, which allows us to both preserve the original raw data files between R sessions.
-Because new NEON data files are released regularly, this approach allows us to easily download only the new files.
-This contrasts to the typical workflow in `neonUtilities`, in which a user discards the individually downloaded files and is left to their own devices on how to keep store the resulting stacked data in order to avoid re-downloading each time.
-
-The best way to set the file directory of the store is to declare the path using the environmental variable, `NEONSTORE_HOME`, for example, using a `.Renviron` file in the user or system home directory.
-For one-off use, all functions that access the store take an optional argument `dir` to specify the directory location, (which defaults to `neon_dir()`, a function that either uses the value set by `NEONSTORE_HOME` or the `rappdirs` location if unset).
-However, using the environmental variable instead results in more portable code, by allowing other users to set their own storage locations independently instead of hard-wiring it into the R code.
-
-The local `neonstore` database is written in the same user and OS specific location by default, but is configured using a separate environmental variable, `NEONSTORE_DB`, or optional argument to appropriate functions,  `db_dir`.
-Because simultaneous write-access is not supported by the default high-performance database engine, `duckdb`, in multi-user environments it may be advantageous to share a location for raw data files by setting `NEONSTORE_HOME` while allowing users to create individual local databases in their home directories by leaving `NEONSTORE_DB` unset.
-Alternately, users can share a database as well by setting `NEONSTORE_DB` to a shared location, and adopting workflows which do ensure multiple users do not simultaneously call `neon_store()` to update the database.
-
-### NEON API functions
-
-`neon_products()` and `neon_sites()` bind the NEON REpresentational State Transfer (REST) API endpoints for `/products` and `/sites`, providing information about each as a data frame.
-This can be useful for product discovery and occassionally for referencing other useful metadata about products and sites.
-Each of these functions requires only a single API call (see discussion of rate limiting below).
+**Data discovery** is facilitated by functions which query the NEON API for information about data contained in it's catalogue.
+The function `neon_products()` returns a data frame with a description and important metdata about each product, including the variables measured and the sites and dates which it covers.
 
 ```{r}
-products <- neon_products()
-products
+neon_products()
 ```
 
-```{r}
-sites <- neon_sites()
-sites
-```
+Similarly, metadata about all sites is avialable from `neon_sites()`.
 
-One of the most important functions of `neonstore` is `neon_download()`, which downloads all files identified by a given product code, optionally constrained to specific site, date range, or file name pattern.
-For instance, we can download all of the NEON landbird survey data, identified as NEON data product `DP1.10003.001`: 
-(The three-digit code identifies this as a Level 1 data product (DP1), product number 10003, revision 001, indicating this is the first and so far the only protocol used for bird survey sampling).
+**Data downloading** is handled by the `neon_download()` function, which can take a vector of sites or products and date ranges, and can download specific tables or all tables and metadata included in a given product.
+The function handles rate limiting and the use of NEON tokens, see appendix or package README for more details including comparison to functionality in `neonUtilities`. 
+Downloaded data is stored in a persistent, configurable location, and `neonstore`'s provenance tracking system avoids re-downloading files that were previously downloaded, and ensures that updated or corrected data is always downloaded without overwriting earlier versions.
 
@@
 ```
-
-To download this data, `neon_download()` must make many API calls.
-It first uses `neon_sites()` to reference all sites containing this data product.
-Then for each site, for each month for which that site collected data, the function will make an API call to the `/data` endpoint (using the `neonstore` function, `neon_data()`), which returns a table of files found for that product at the requested site and month range.
-The  table from `neon_data()` indicates 4 values: the file name, the MD5 or CRC32 hash of the data file (CRC32 hashes are used only for AOP data products), the file size (in bytes -- encoded as a character because some file sizes are too long to be represented as an integer), and a download URL.
-The download URL embeds a temporary access token which expires after an hour, so users must complete all the downloads before this time expires.
-The access string for each file is unique, thus it is (at the time of writing) impossible to download any given data file without making a corresponding API request, even if the precise file name and address is already known.
-For the most part, this issue is invisible to `neonstore` users, as `neon_download()` calls automatically pass all download URLs to the download step.
-
-`neon_download()` handles the following tasks:
-
-- `neonstore` aims to provide persistent storage, writing raw data files to
-  the appropriate app directory for your operating system (see `rappdirs`,
-  [Ratnakumar et al 2016](https://CRAN.R-project.org/package=rappdirs)), now provided by the base R function `tools::R_user_dir()`.
-  Files are organized into appropriate subdirectories by product, site, and year-month of sampling.
-- `neon_download()` provides  clean and concise progress bars for the two key
-  processes involved: querying the API to obtain download URLs (which involves no 
-  large data transfer but counts against API rate limiting, see below), and the
-  actual file downloads.
-- `neon_download()` will verify the integrity of file downloads against the MD5 or CRC32
-  hashes provided. 
-- `neon_download()` will omit downloads of any existing data files in the local 
-  store (even if the timestamp portion of the filename has changed without updating the file)
-- You can request multiple products at once using vector notation, though API
-  rate limiting may interfere with large requests.
-- `neon_download()` uses `curl::curl_download()` instead of `downloadr` package
-  used in `neonUtilities`, which can be finicky on Windows and older versions of R. 
-- Automatic handling NEON's rate limiting policy.
-  To do this, `neon_download()` must  pause 100 seconds every 150 requests if an API token is not supplied or every 950 requests if a token is available.
-  Should a rate-limit error be detected, (possible if the same token is used in parallel requests), `neon_download()` will also pause for 100 seconds before retrying the failed request.
-  Requests and pauses are communicated through the user messaging interface.
-- Expanding compressed HDF5 files (`*h5.gz`).
-  (Note: unusually, NEON chooses to compress binary h5 files with gzip, even though gzip compression is built internally into the H5 specification at configurable compression levels.
-  NEON does not choose to compress text-based formats such as .csv, where compression has larger gains.
-  Compression has implications for version management, where `neonstore` must track the hashes of compressed files to avoid duplicate downloads)
-- Updating the local manifest in LMDB with the individual checksum and current release tag for each downloaded object (see Provenance)
 
@@
 
-## Interacting with downloaded files
+# Interacting with downloaded files
+
+`neonstore` provides three mechanisms for working with downloaded data, each of which builds on the former: a file interface, a relational database, and remote database.
+The relational database helps `neonstore` scale to big data analyses, while the remote database mechanism extends that further to scale to research teams.
+
+## File Interface: `neon_index()` & `neon_read()`
 
@@
 `neon_index()` takes optional arguments such as `product`, `start_date`, `end_date`, and `site` to allow for convenient sub-setting.
+
+`neon_index()` also allows users to access metadata files, such as README and EML files associated with each data product and/or NEON site.
 
@@
 
+## Relational Database Interface
+
 For very large tables such as sensor data collections spanning many sites and years, users will require the increased performance provided by the high-performance, columnar-oriented relational database, `duckdb` [@Raasveldt_2019; @duckdb].
@@
 
-**Working with very large data products**
-
 ```{r}
@@
 Note that we exploit pattern matching to provide a short-hand to the table name.
-We download and import the 30m sampled table. 
+We download and import the 30 minute sampled table. 
 The initial download is not a fast operation - in our example, rate-limiting rules by NEON make the command take over 20 minutes to complete the API requests that would otherwise take only 3 minutes.
@@
 
-**Network-based access**.
+**Cross-language support**: The `duckdb` database created by `neonstore` is not tied to R, but can easily be accessed using other languages supported by `duckdb`, including python, javascript, web assembly, and a stand-alone CLI interface [@duckdb].
 
-In R, we refer to the `duckdb` database as a "remote" connection because the operations are handled outside of R, by `duckdb`.
-Some database connections, like MySQL or Postgres, require the user to install a separate piece of software, often on an external server,
-which we then access over the network connection (with appropriate security credentials), though it can also be run on the same local machine as the 'client' software.
-This client-server approach can be very powerful, but requires more knowledge and resources to deploy.
-In contrast, `duckdb` installs automatically as a stand-alone R package.
+## Network Database Interface
 
-As illustrated above, the slowest part of the process by far are the one-time steps of initial download of over 10,000 individual files, and the importing of those files into the database.
-Further, the storage requirements for users working with NEON data spanning many products, sites, and dates can be restrictive. 
-As such, it may be desirable for a research team working on such an analysis to establish a single shared `neonstore` storage center rather than expect that all users will download, import, and maintain their own local stores.
-To facilitate this process, `neonstore` supports access to remote network-based storage over cloud-native storage protocols with the help of  Apache `arrow` [@arrow].
-First, using `neonstore`, we download and import the data as above.
-Our next step is export this database to a parquet snapshot using `neon_export_db()` and upload that data to cloud native storage system such as Amazon Web Services (AWS) S3 bucket, or, in our case, a self-hosted MINIO [@minio] instance.
-This is much cheaper and raises less security concerns than maintaining a server, as the remote service is providing only networked storage, not general-purpose computing power.
+While most use cases will be covered by the interfaces discussed above, the networked database interface may be particularly useful for those wishing to scale analyses within or across teams.
+The relational database interface still assumes that each user of the data individually downloads the NEON products they are using and imports that data into the local `duckdb` database on their machine.
+While it is possible for multiple users to access the same database on a single shared server, it is common today for each researcher runs analyses on their own laptop, and also common for an individual researcher to work on their analyses on multiple machines (for instance, a laptop and a server).
+Despite the streamlined tools available in `neonstore`, re-downloading and storing a large library of NEON files on many machines or users can be time-consuming and inefficient.
+For example, downloading and importing the NEON data products required for the NEON Ecological Forecasting Challenge [@efi_neon] takes several hours and occupy over 100 GB of storage.
+In such instances, network-based access provide a much more efficient mechanism for a team to interact with a shared `neonstore` database.
+`neonstore` can take advantage of "cloud-native" interface mechanisms implemented in the Apache `arrow` [@arrow] library and the capabilities of the modern Apache `parquet` [@parquet] data serialization to provide shared access that does not require re-downloading or transferring the entire library to each machine, without the use of a central server.
+In this approach, one user takes the responsibility of running the `neon_download()` and `neon_store()` commands discussed above to create a database of containing the desired data products.
+That user can then make this database available to other users through an S3 bucket [@aws.s3] or compatible system, such as the open-source software MINIO [@minio], using the function `neon_export_db()`.
+Note that by only requiring a 'static' storage mechanism such as an S3 bucket, this approach is much cheaper and raises fewer security concerns than maintaining a central server running a database server such as MySQL.
+Future development may enable access from other static storage services, including Azure cloud or GitHub (already possible in Python, see @fsspec).
 
-We pass the connection details to `neon_remote_db`.
-In this case, we use a public (read-only) bucket, though other users may prefer to configure access that requires a secret key.
+Other team members can then access the remote database using the `neon_remote_db()` and `neon_remote()` functions, without ever having to download the data:
+We pass the connection details to `neon_remote_db` (which can include authentication tokens if desired, see documentation for details.)
+In this case, we will use a public (read-only) bucket:
 
@@
 
-Because this approach relies on operations over a network connection, it will generally be bandwidth limited unless the bucket is located in the same region as compute operations (e.g.
-in a regional cloud server).
-Users may also download the `parquet` exported files directly and import them into a local `neonstore` instance (see `neon_import_db()`) to share neonstore snapshots without requiring independent downloads.
+Because this approach relies on operations over a network connection instead of a local disk, it will be most effective in contexts where high-speed networks are available, including local area networks and cloud computing platforms.
+Alternately, Users may also download the `parquet` exported files directly and import them into a local `neonstore` instance (see `neon_import_db()`) to share neonstore snapshots without requiring independent downloads over the NEON API.
 Using `neon_db_export()` to generate snapshots is also a reliable way to preserve and version your analyses for future that does not depend on access to the NEON API or even to `neonstore`.
-This approach should work equally well for data exchange with collaborators not using R.
-`parquet` is a widely used data interchange format with support in most major languages, including python, Java and C++.
-Similarly, `duckdb` is a cross-platform tool that can be accessed not only from R, but from python, Java, and many other languages, including web assembly (for running in the browser) or run using a stand-alone CLI [@duckdb].
 
-## Dependencies and optional dependencies
+**Cross-language access**: Once again, this approach should work equally well for data exchange with collaborators not using R.
+Apache `arrow` can be used to directly access these remote parquet files from Python, Julia, Matlab, C++, Javascript, Rust, Go, and other popular languages in much the same way [@arrow]. 
+This approach thus allows the `neonstore` user to facilitate efficient and scalable access to their NEON library not only across machines but also across languages.
 
-`neonstore`'s core dependencies are quite light, avoiding hard requirements on other packages which themselves bring in a large recursive list of dependencies.
-`DBI` [@dbi], `duckdb` [@duckdb], and `thor` [@thor] provide the core database interfaces; all have no recursive dependencies.
-`httr` [@httr] provides functionality required to interface with the NEON REST API, and `R.utils`, `zip`, and `vroom` [@vroom] provide parsers required to unpack the data.
-`progress` provides user-friendly progress bars and estimated run-times for slow functions.
-Optional dependencies are required for specific tasks. `dplyr` is needed for lazy read in `neon_table()`.
-`neonUtilities` [@neonUtilities] and `rhdf5` are required at this time for reading in eddy-covariance products.
-`arrow` [@arrow] is required for remote database connections using `neon_remote()`.
-`rstudioapi` is required to utilize `neon_pane()` to view the connections pane from within RStudio.
-The remaining suggested dependencies are utilized within testing only.
+<!--
+In R, we refer to the `duckdb` database as a "remote" connection because the operations are handled outside of R, by `duckdb`.
+Some database connections, like MySQL or Postgres, require the user to install a separate piece of software, often on an external server,
+which we then access over the network connection (with appropriate security credentials), though it can also be run on the same local machine as the 'client' software.
+This client-server approach can be very powerful, but requires more knowledge and resources to deploy.
+In contrast, `duckdb` installs automatically as a stand-alone R package.
+-->
 
@@
 Additional utilities enable serverless network-based sharing of a large `neonstore` database across a team through the use of `parquet` and `arrow`'s remote filesystems (which likewise supports most major computing languages.)
-`neonstore` has been co-developed by NEON staff (CL & CL) and members of the Ecological Forecasting Initiative (CB & RQT), tested in supporting the EFI NEON forecasting challenge [@efi_challenge], among other use cases.
+`neonstore` has been co-developed by NEON staff (CML & CL) and members of the Ecological Forecasting Initiative (CB & RQT), tested in supporting the EFI NEON forecasting challenge [@efi_challenge], among other use cases.
 

### 2022-03-28 R/neon_store.R

@@ -155,7 +155,7 @@ db_chunks <- function(con,
 
 duckdb_memory_manager <- function(con){
   if(!inherits(con, "duckdb_connection")) return(con)
-  if(Sys.getenv("duckdb_restart", TRUE)){
+  if(Sys.getenv("duckdb_restart", FALSE)){
     message("finalizing duckdb import to disk...")
     ## power cycle to force import
     ## shouldn't be necessary when memory management improves in duckdb...

### 2022-03-28 inst/examples/thelio_mirror.R

@@ -1,7 +1,8 @@
 bench::bench_time({
 
 library(neonstore)
-
+  message(neon_dir())
+  
   
   mem_limit <- function(db = neon_db(), mem_limit = 16, units = "GB"){
     DBI::dbExecute(db, paste0("PRAGMA memory_limit='", mem_limit, " ", units,"'"))
@@ -57,7 +58,7 @@ neon_download("DP1.00098.001", table = "30") # Humidity (includes temp)
 
 ## SQL Import
 neon_store(product = "DP1.10003.001") # Birds
-neon_store(product = "DP1.10022.001") # Beetles
+neon_store(product = "DP1.10022.001", type = "expanded") # Beetles
 neon_store(product = "DP1.10093.001") # Ticks
 
 ## Aquatics SQL Import

### 2022-08-29 R/neon_export_db.R
standardize_names

@@ -103,11 +103,20 @@ neon_sync_db <- function(s3, dir = file.path(neon_dir(), "parquet")) {
   
 }
 
-
-
-
-standardize_export_names <- function(s3, 
-                                     dir = file.path(neon_dir(), 
+#' standardize export names
+#' 
+#' @details
+#' DUCKDB clobbers database filenames to avoid potentially incompatible characters.
+#' This is pretty unnecessary, so we can restore the original table names for
+#' use with S3-based remote access which assumes parquet files map to the 
+#' desired table names (i.e. including product numbers.)
+#' 
+#' However, note that `[neon_import_db()]` uses native duckdb functions 
+#' that assume the original mangled names.
+#' 
+#' @inheritParams neon_export_db
+#' @export
+standardize_export_names <- function(dir = file.path(neon_dir(), 
                                                      "parquet")
                                      ) {
   table_names <- parquet_labels(dir)

### 2022-08-29 R/neon_export_db.R
file.rename

@@ -123,7 +123,7 @@ standardize_export_names <- function(dir = file.path(neon_dir(),
   file_paths <- names(table_names)
   status <- lapply(seq_along(table_names), 
                    function(i) {
-                     fs::file_move(file_paths[[i]], table_names[[i]])
+                     file.rename(file_paths[[i]], table_names[[i]])
                    })
 
 }

### 2022-08-29 README.Rmd
update matrix

--- 2021-03-12 README.Rmd
@@
 [![R build status](https://github.com/cboettig/neonstore/workflows/R-CMD-check/badge.svg)](https://github.com/cboettig/neonstore/actions)
-[![Codecov test coverage](https://codecov.io/gh/cboettig/neonstore/branch/master/graph/badge.svg)](https://codecov.io/gh/cboettig/neonstore?branch=master)
+[![Codecov test coverage](https://codecov.io/gh/cboettig/neonstore/branch/master/graph/badge.svg)](https://app.codecov.io/gh/cboettig/neonstore?branch=master)
 [![CRAN status](https://www.r-pkg.org/badges/version/neonstore)](https://CRAN.R-project.org/package=neonstore)
+[![R-CMD-check](https://github.com/cboettig/neonstore/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/cboettig/neonstore/actions/workflows/R-CMD-check.yaml)
 <!-- badges: end -->

### 2024-01-02 R/neon_cloud.R
higher-level memoise, instrument data parsing

@@ -21,36 +21,45 @@ neon_cloud <-function(table,
                       unify_schemas = FALSE,
                       .token = Sys.getenv("NEON_TOKEN")){
   
-  df <- neon_data(product, start_date, end_date, site, type, release,
-                  quiet, api, .token = .token)
+  urls <- neon_urls(table, product, start_date, end_date, site, type, release,
+                    quiet, api, .token = .token)
   
-  # df <- neon_data(product = "DP1.10003.001", type="basic")
-  # table = "brd_countdata"
+  format <- gsub(".*(.\\w+)$", "\\1", urls)
+  if(all(format == "csv")) {
+    df <- cloud_csv(urls)
+  }
   
-  urls <- df$url[ grepl(table, df$name) ]
-  
+  df
+}
+ 
+cloud_csv <- function(urls) {   
   # Parse most recent first. Reduces the chance of int/char coercion failures
   # when hitting an all-empty column
-  meta <- neon_filename_parser(urls)
-  chrono <- order(meta$GENTIME, decreasing = TRUE)
-  urls <- urls[chrono]
+  #meta <- neon_filename_parser(urls)
+  #chrono <- order(meta$GENTIME, decreasing = TRUE)
+  #urls <- urls[chrono]
   ## discover or enforce extension
-  
   ## Detect product type from `meta` and dispatch appropriately.
   
   # https://duckdb.org/docs/data/csv/overview.html
-  # consider igonore_errors=1 as possible fallback? (drops those rows entirely, ick)
-  # consider all_varchar=1 as possible fallback (possibly avoided by coercion after parsing in chrono order?)
+  # consider all_varchar=1 as possible fallback 
+  # (possibly avoided by coercion after parsing in chrono order?)
   df <- duckdbfs::open_dataset(urls, format = "csv", filename = TRUE,
                                unify_schemas = unify_schemas)
+  cols <- colnames(df)
   
-  
-  
-  
+  # sensor metdata
+  if(! "siteID" %in% cols) {
+    df <- df |> dplyr::mutate(
+      file = split_part(filename, "/", 9L), 
+      siteID = regexp_extract(file, IS_DATA, 3L),
+      domainID = regexp_extract(file, IS_DATA, 2L),
+      horizontalPosition= regexp_extract(file, IS_DATA, 7L),
+      verticalPosition = regexp_extract(file, IS_DATA, 8L)
+      )
+  }
+  df
 }
-
-  
-
 
 
 neon_urls <- function(table,
[…]

### 2024-01-02 R/neon_data.R

@@ -23,10 +23,6 @@ neon_data <- function(product,
                       .token = Sys.getenv("NEON_TOKEN")){
   
   
-  cache <- cachem::cache_disk(tempdir())
-  mGET <- memoise::memoise(httr::GET, cache = cache)
-  
-  
   data_api <- data_api_queries(product = product, 
                                start_date = start_date, 
                                end_date = end_date, 
@@ -53,7 +49,7 @@ neon_data <- function(product,
   resp <- vector("list", length = length(data_api))
   for(i in seq_along(data_api)){
     if(!quiet){ pb$tick() }
-    resp[[i]] <- mGET(data_api[[i]],
+    resp[[i]] <- GET(data_api[[i]],
                       httr::add_headers("X-API-Token" = .token))
     status <- neon_warn_http_errors(resp[[i]])
     

### 2024-01-02 tests/testthat/test-cloud.R

@@ -5,6 +5,23 @@ test_that("Direct cloud access", {
   
   skip_if_offline()
   skip_on_cran()
+  # skip_on_os("windows")
+  
+  urls <-  neonstore:::neon_urls(table = "waq_instantaneous",
+                                product = "DP1.20288.001",
+                                start_date = "2023-06-01",
+                                end_date = "2023-08-01",
+                                type="basic"
+  )
+  
+  format <- gsub(".*\\.(\\w+)$", "\\1", urls)
+})
+
+
+test_that("Direct cloud access", {
+  
+  skip_if_offline()
+  skip_on_cran()
  # skip_on_os("windows")
   
   df <-  neonstore:::neon_cloud(table = "waq_instantaneous",
@@ -15,6 +32,17 @@ test_that("Direct cloud access", {
   )
   
   expect_s3_class(df, "tbl_lazy")
+  cols <- colnames(df)
+  expect_true("siteID" %in% cols)
+  
+  test <- df |> 
+    dplyr::select(siteID, domainID, horizontalPosition, verticalPosition) |>
+    dplyr::collect()
+
+  df_local <- dplyr::collect(df)
+  expect_s3_class(df, "tbl")
+  
+  
   
   
   df <-  neonstore:::neon_cloud(table = "bet_sorting",
@@ -26,6 +54,30 @@ test_that("Direct cloud access", {
   
   
   expect_s3_class(df, "tbl_lazy")
+  cols <- colnames(df)
+  expect_true("siteID" %in% cols)
+  expect_s3_class(df, "tbl")
+  df_local <- dplyr::collect(df)
+  
+})  
+  
+  
+test_that("Big (rate-limited) tests of direct cloud access", {
+    
+    skip_if_offline()
+    skip_on_cran()
+    # skip_on_os("windows")
+    
+  
+  df <-  neonstore:::neon_cloud(table = "waq_instantaneous",
+                                product = "DP1.20288.001",
+                                start_date = "2018-01-01",
+                                end_date = "2023-08-01",
+                                type="basic"
+  )
+  
+  expect_s3_class(df, "tbl_lazy")
+  df_local <- dplyr::collect(df)
   
   
 })

### 2024-01-02 R/neon_data.R
fix typo

@@ -54,7 +54,7 @@ neon_data <- function(product,
     status <- neon_warn_http_errors(resp[[i]])
     
     if(status == 429){ # retry once
-      resp[[i]] <- mGET(data_api[[i]],
+      resp[[i]] <- GET(data_api[[i]],
                         httr::add_headers("X-API-Token" = .token))
     }
     

### 2024-01-02 R/neon_cloud.R
punt on h5

@@ -26,13 +26,16 @@ neon_cloud <-function(table,
   
   format <- gsub(".*(.\\w+)$", "\\1", urls)
   if(all(format == "csv")) {
-    df <- cloud_csv(urls)
+    df <- cloud_csv(urls, unify_schemas = unify_schemas)
+  } else {
+    stop("Only csv formatted data are accessible without download at this time.")
   }
   
   df
 }
  
-cloud_csv <- function(urls) {   
+
+cloud_csv <- function(urls, unify_schemas = FALSE) {   
   # Parse most recent first. Reduces the chance of int/char coercion failures
   # when hitting an all-empty column
   #meta <- neon_filename_parser(urls)
@@ -88,3 +91,9 @@ neon_urls <- function(table,
   urls <- df$url[ grepl(table, df$name) ]
   
 }
+
+
+
+
+globalVariables(c("regexp_extract", "split_part", "filename"),
+                package = "neonstore")

### 2024-01-02 R/read_eddy_cloud.R

@@ -1,0 +1,20 @@
+
+
+# Not operational until non-gz H5 versions are made available
+read_eddy_cloud <- function() {
+
+  public_S3_url <- "https://storage.googleapis.com/neon-int-sae-tmp-files/ods/dataproducts/DP4/2020-04-26/GRSM/NEON.D07.GRSM.DP4.00200.001.nsae.2020-04-26.expanded.20230921T200400Z.h5"
+  #listGrp <- rhdf5::h5ls(file = public_S3_url, s3 = TRUE)
+  nee <- rhdf5::h5read(file = public_S3_url,
+                       name = glue::glue("{site}/dp04/data/fluxCo2/turb",
+                                         site="GRSM"), 
+                       s3 = TRUE) 
+  le <- rhdf5::h5read(file = public_S3_url,
+                      name = glue::glue("{site}/dp04/data/fluxH2o/turb",
+                                        site="GRSM"), 
+                      s3 = TRUE) 
+  
+
+  
+}
+

### 2024-01-02 paper/paper.Rmd
wip

@@
   - name: "Christine M. Laney"
+    affiliation: neon
+  - name: "Eric Sokol"
     affiliation: neon    
@@
 Here, we describe the design, use, and rationale for the package.
-We illustrate a few example use cases, including downloading the 30 minute single-aspirated temperature data across all available NEON sites and dates, a collection of over 13 thousand individual files totalling nearly 2 gigabytes of data.
-We demonstrate how `neonstore` facilitates summary calculations such as computing monthly means across all sites in a fraction of a second with minimal memory requirements.
+We illustrate a few example use cases, including working with larger-than-RAM data and cloud-native access patterns that avoid direct downloads.
 The approaches taken in `neonstore` maximize reproducibility, facilitate collaboration including across computer languages, and help making analyses which leverage the full spatial and temporal scale of NEON easily available to researchers with modest computational resources and expertise.
@@
 A focus on performance reflects the potentially very large size of its aggregated data products, which can easily exceed available working memory (RAM) on most machines: as many unified tables of even single products over all available sites and dates can range from 10 GBs into the terrabyte range.
-Because enabling analyses that could span the full spatial and temporal range of NEON's 81 sites over a 30 year horizon is a foundational motivation for NEON research, it essential that researchers be able to easily work with data across this full span.
+Because enabling analyses that could span the full spatial and temporal range of NEON's 81 sites over a 30 year horizon is a fundamental motivation for NEON research, it essential that researchers be able to easily work with data across this full span.
 However, doing so requires techniques that go beyond workflows which expect data to be re-downloaded (provenance) or stored in memory for analysis (performance).
 
-To address these challenges, `neonstore` implements a high-performance, provenance aware workflow by drawing on a host of key technologies, including cryptographic hashes, a Lightning Memory Mapped Database (LMDB) for metadata management, a columnar-oriented integrated relational database (duckdb) for data management, and optional support for remote access through parquet files cloud-native object stores to facilitate sharing large databases generated by `neonstore`.
+To address these challenges, `neonstore` implements a high-performance, provenance aware workflow by drawing on a host of key technologies, including cryptographic hashes, a Lightning Memory Mapped Database (LMDB) for metadata management, a columnar-oriented integrated relational database (duckdb) for data management, and  support for remote access through cloud-native object stores.
 These technologies are seamlessly integrated behind a user-friendly R interface which requires no expertise in the underlying backend technology beyond a familiarity with R and the "tidy data" model used in the popular R packages, `dplyr` and `tidyr`.
-
-# Provenance
-
-`neonstore` focus on provenance emphasizes being able to trace results back precisely to individual raw data files provided by the NEON API.
-NEON data products are distributed using a highly atomized file design: with each data product from either both observation systems (OS) and instrument systems (IS) divided into separate files by site and month.
-Products generated by the airborne observation platform (AOP) are typically divided by type, flight date, flight line, and grid cell.
-This atomized approach facilitates sub-setting and data transfer, because users can identify ahead of time which products, locations, and time ranges they need.
-As some NEON data products can be quite large (NEON's complete data catalog to date is already in excess of 1 petabyte, primarily due to the size of many AOP products), being able to download or read only the files of interest is already important to save bandwidth or computer memory, or disk space.
-This approach also facilitates data provenance and reproducibility, because it means that an update or correction to data collected at a particular site and time results only in a change to a single file and not the whole data product.
-However, the highly atomized structure of NEON's data can also become a barrier to entry.
-Most analyses will inevitably require multiple files to be combined or stacked, e.g.
-to create a unified record for a product across many months or years of sampling or across many sites.
-`neonstore` provides functions to stack data products that have been atomized into separate files by month, site or sensor position into single `data.frame` objects, store stacked tables in a local, high-performance database, and track individual data rows back to their raw data files.
-
-Managing provenance of NEON data means dealing with the addition of new data, corrections published to old data, and changes to file names that do not represent changes in underlying data.
-Any analysis is only as good as as the data supporting it.
-Because NEON data is subject to both revisions (where previously released data are corrected) and new releases (where more recently collected data is made public), researchers will seek to update previous analyses with corrected data and new observations.
-`neonstore` enables scripted workflows through which such updates can occur automatically by re-running the same script.
-More confusingly, NEON frequently updates the timestamp associated with a file even when the file has not changed.
-Typically these events are associated with the update of a separate but related file.
-In such cases, `neonstore` can avoid downloading and importing these unchanged data, ensuring that automated workflows do not run unnecessarily.
-When new or revised data significantly alter results, it is also important to be able to pinpoint precisely what has changed and why.
-These elements creates five distinct tasks for managing NEON data provenance:
-(1) The ability to determine which of each of the individual raw data files were used in an analysis 
-(2) The ability to append new data as it is collected and released 
-(3) the ability to transparently replace existing data when corrections are published, 
-(4) the ability to avoid re-downloading or re-stacking data that has been previously processed, 
-(5) the ability to freeze the data used in a given analysis, so that results can reliably be repeated without accidentally including revised or newer data products.
-
-To provide these five abilities for data provenance, `neonstore` maintains a local file store containing all raw data files downloaded from NEON.
-Additionally, each file is entered into a local registry, along with the file's MD5 checksum (CRC32 for AOP products) and currently associated NEON release (if any).
-This local registry is kept in a Lightning Memory Mapped Database, LMDB, an extremely fast, open source key-value store indexed on both file name and checksum.
-MD5 checksums effectively indicate unique content.
-It is theoretically possible for a malicious actor to create a file with different contents but the same checksum, but vanishingly unlikely to occur by chance.
-On requesting new files to download, `neonstore` compares file names and hashes reported by the API to those in the local store, allowing it to omit any data that has already been downloaded, even if the timestamp in a file name has changed but the contents remain the same.
-
-NEON's filenaming convention provided essential metadata about each file, such as the data product (product level, product number, revision number), site, year and month of the release, file description, a timestamp.
-Sensor data (IS) also include additional information about the horizontal and vertical position and meaasurement interval (frequency) of the sensor product.
-AOP products use a somewhat different convention with flight information.
-`neonstore` recognizes all published NEON file naming conventions and automatically parses NEON filenames into their component parts.
-Data are organized into subdirectories based on product, site, and year-month respectively.
-Data or metadata files shared across all months or all sites of a product are stored in the corresponding parent directory.
-The function `neon_index()` returns a `data.frame` listing every recognized individual file in the local store, along with the corresponding metadata fields parsed from the filename.
-Optional arguments can be given to restrict this table to files matching specific products, sites, date ranges, etc.
 
@@
 Functions which in read in the data should faithfully reproduce the raw data so that analyses can easily be compared to those using other software tools.
-When functions perform tasks like guessing the data type (numeric, string, Date, etc) of a column, or adding additional columns to stack two otherwise matching tables in which one has recorded an additional variable, they should do so in a manner that is transparent, controllable, and consistent with the behaviour of common functions used in the language.
+When functions perform tasks like guessing the data type (numeric, string, Date, etc) of a column, or adding additional columns to stack two otherwise matching tables in which one has recorded an additional variable, they should do so in a manner that is transparent, controllable, and consistent with the behavior of common functions used in the language.
 
@@
 
-## File Interface: `neon_index()` & `neon_read()`
+## File-based Interface: `neon_index()` & `neon_read()`
 
@@
 
-Consider the example of a much larger table collected by sensor data; say, single-aspirated temperature collected every 30 mintues across all sites:
-Note that we exploit pattern matching to provide a short-hand to the table name.
-We download and import the 30 minute sampled table. 
+Consider the example of a much larger table collected by sensor data; say, single-aspirated temperature collected every 30 minutes across all sites.
+We download and import the 30 minute sampled table
+(ote that we exploit pattern matching to provide a short-hand to the table name).
 The initial download is not a fast operation - in our example, rate-limiting rules by NEON make the command take over 20 minutes to complete the API requests that would otherwise take only 3 minutes.
@@
 In our example this takes about 1 hour to download over 13 thousand individual files.
-Because NEON uses download links that expire after 1 hour, it may be necessary to repeat the download command after more than an hour expires,
-(alternatively, divide the download into multiple batches by years or sites).
-The `neon_download()` utility can pick up from where it left off.
 
@@
 temp %>% select(verticalPosition, horizontalPosition) %>% distinct()
-
 index <- neon_index(product = "DP1.00002.001", table = "30min") 
@@
 
-**Cross-language support**: The `duckdb` database created by `neonstore` is not tied to R, but can easily be accessed using other languages supported by `duckdb`, including python, javascript, web assembly, and a stand-alone CLI interface [@duckdb].
+# Cloud-based access
 
-## Network Database Interface
+Cloud-based access offers a simpler and sometimes faster alternative to the standard pattern of first downloading the data (with `neon_download()`), importing to the local database (with `neon_store()`) and then accessing from the database (with `neon_table()`).
+In cloud-based access, we simply access data directly from the Google Cloud Storage platform without downloading:
 
-While most use cases will be covered by the interfaces discussed above, the networked database interface may be particularly useful for those wishing to scale analyses within or across teams.
-The relational database interface still assumes that each user of the data individually downloads the NEON products they are using and imports that data into the local `duckdb` database on their machine.
-While it is possible for multiple users to access the same database on a single shared server, it is common today for each researcher runs analyses on their own laptop, and also common for an individual researcher to work on their analyses on multiple machines (for instance, a laptop and a server).
-Despite the streamlined tools available in `neonstore`, re-downloading and storing a large library of NEON files on many machines or users can be time-consuming and inefficient.
-For example, downloading and importing the NEON data products required for the NEON Ecological Forecasting Challenge [@efi_neon] takes several hours and occupy over 100 GB of storage.
-In such instances, network-based access provide a much more efficient mechanism for a team to interact with a shared `neonstore` database.
-`neonstore` can take advantage of "cloud-native" interface mechanisms implemented in the Apache `arrow` [@arrow] library and the capabilities of the modern Apache `parquet` [@parquet] data serialization to provide shared access that does not require re-downloading or transferring the entire library to each machine, without the use of a central server.
-In this approach, one user takes the responsibility of running the `neon_download()` and `neon_store()` commands discussed above to create a database of containing the desired data products.
-That user can then make this database available to other users through an S3 bucket [@aws.s3] or compatible system, such as the open-source software MINIO [@minio], using the function `neon_export_db()`.
-Note that by only requiring a 'static' storage mechanism such as an S3 bucket, this approach is much cheaper and raises fewer security concerns than maintaining a central server running a database server such as MySQL.
-Future development may enable access from other static storage services, including Azure cloud or GitHub (already possible in Python, see @fsspec).
-
-Other team members can then access the remote database using the `neon_remote_db()` and `neon_remote()` functions, without ever having to download the data:
-We pass the connection details to `neon_remote_db` (which can include authentication tokens if desired, see documentation for details.)
-In this case, we will use a public (read-only) bucket:
-
-```{r include=FALSE}
-Sys.unsetenv("AWS_DEFAULT_REGION")
-Sys.unsetenv("AWS_S3_ENDPOINT")
-Sys.unsetenv("AWS_ACCESS_KEY_ID")
-Sys.unsetenv("AWS_SECRET_ACCESS_KEY")
-Sys.setenv("AWS_EC2_METADATA_DISABLED"="TRUE")
+```{r}
+df <- neon_cloud(product = "DP1.00002.001", table = "30min") # Temp single aspirated
 ```
 
-```{r cache=FALSE}
-con <- neon_remote_db(host = "minio.thelio.carlboettiger.info",
-                      bucket = "shared-data",
-                      path = "neonstoredb")
+Unfortunately, at this time NEON still requires us to access the API to obtain a list of individual file URLs.
+The API is quite slow, because that list of files requires many separate API requests (for each site and for each year-month in the request).
+In future this may become a single request and will thus be almost instantaneous.
+
+`neon_cloud()` does not immediately download any of the actual data fields.
+Instead, it merely reads the metadata (the table schema, i.e.
+column names) of the first CSV in the collection.
+This approach is often referred to as "lazy evaluation" in software engineering design -- the function executes as quickly as possible by deferring any computation until it is absolutely necessary.
+For instance, if we are going to work with only a few columns of the data, we can save computational effort by not downloading and parsing all those extra columns.
+Despite this strategy, our `df` object "feels" like a normal data.frame in most ways -- we can perform most of the usual operations as if the data were already both downloaded and read into memory.
+Here we immediately perform the same calculation as before:
+
+```{r}
+
+temp_monthly <- df %>%
+  mutate(month = month(startDateTime)) %>%
+  group_by(siteID, month) %>%
+  summarise(mean_temp = mean(tempSingleMean, na.rm=TRUE),
+            max = max(tempSingleMean, na.rm=TRUE),
+            min = min(tempSingleMean, na.rm=TRUE),
+            .groups = "drop") %>% 
+  collect()
+
 ```
 
-Once a connection to the remote database is established, we can request individual tables using `neon_remote` to access the data using dplyr functions.
-Like `neon_table()`, we must use `collect()` to draw the data into R if we wish to perform advanced R-specific operations that are not supported by the backend engine.
+# Provenance
 
-```{r  cache=FALSE}
-remote_temp <- neon_remote("30min", product = "DP1.00002.001", db = con)
-remote_temp %>% count(siteID) %>% collect()
-```
+`neonstore` focus on provenance emphasizes being able to trace results back precisely to individual raw data files provided by the NEON API.
+NEON data products are distributed using a highly atomized file design: with each data product from either both observation systems (OS) and instrument systems (IS) divided into separate files by site and month.
+Products generated by the airborne observation platform (AOP) are typically divided by type, flight date, flight line, and grid cell.
+This atomized approach facilitates sub-setting and data transfer, because users can identify ahead of time which products, locations, and time ranges they need.
+As some NEON data products can be quite large (NEON's complete data catalog to date is already in excess of 1 petabyte, primarily due to the size of many AOP products), being able to download or read only the files of interest is already important to save bandwidth or computer memory, or disk space.
+This approach also facilitates data provenance and reproducibility, because it means that an update or correction to data collected at a particular site and time results only in a change to a single file and not the whole data product.
+However, the highly atomized structure of NEON's data can also become a barrier to entry.
+Most analyses will inevitably require multiple files to be combined or stacked, e.g.
+to create a unified record for a product across many months or years of sampling or across many sites.
+`neonstore` provides functions to stack data products that have been atomized into separate files by month, site or sensor position into single `data.frame` objects, store stacked tables in a local, high-performance database, and track individual data rows back to their raw data files.
 
-Because this approach relies on operations over a network connection instead of a local disk, it will be most effective in contexts where high-speed networks are available, including local area networks and cloud computing platforms.
-Alternately, Users may also download the `parquet` exported files directly and import them into a local `neonstore` instance (see `neon_import_db()`) to share neonstore snapshots without requiring independent downloads over the NEON API.
-Using `neon_db_export()` to generate snapshots is also a reliable way to preserve and version your analyses for future that does not depend on access to the NEON API or even to `neonstore`.
+Managing provenance of NEON data means dealing with the addition of new data, corrections published to old data, and changes to file names that do not represent changes in underlying data.
+Any analysis is only as good as as the data supporting it.
+Because NEON data is subject to both revisions (where previously released data are corrected) and new releases (where more recently collected data is made public), researchers will seek to update previous analyses with corrected data and new observations.
+`neonstore` enables scripted workflows through which such updates can occur automatically by re-running the same script.
+More confusingly, NEON frequently updates the timestamp associated with a file even when the file has not changed.
+Typically these events are associated with the update of a separate but related file.
+In such cases, `neonstore` can avoid downloading and importing these unchanged data, ensuring that automated workflows do not run unnecessarily.
+When new or revised data significantly alter results, it is also important to be able to pinpoint precisely what has changed and why.
+These elements creates five distinct tasks for managing NEON data provenance:
+(1) The ability to determine which of each of the individual raw data files were used in an analysis 
+(2) The ability to append new data as it is collected and released 
+(3) the ability to transparently replace existing data when corrections are published, 
+(4) the ability to avoid re-downloading or re-stacking data that has been previously processed, 
+(5) the ability to freeze the data used in a given analysis, so that results can reliably be repeated without accidentally including revised or newer data products.
 
-**Cross-language access**: Once again, this approach should work equally well for data exchange with collaborators not using R.
-Apache `arrow` can be used to directly access these remote parquet files from Python, Julia, Matlab, C++, Javascript, Rust, Go, and other popular languages in much the same way [@arrow]. 
-This approach thus allows the `neonstore` user to facilitate efficient and scalable access to their NEON library not only across machines but also across languages.
+To provide these five abilities for data provenance, `neonstore` maintains a local file store containing all raw data files downloaded from NEON.
+Additionally, each file is entered into a local registry, along with the file's MD5 checksum (CRC32 for AOP products) and currently associated NEON release (if any).
+This local registry is kept in a Lightning Memory Mapped Database, LMDB, an extremely fast, open source key-value store indexed on both file name and checksum.
+MD5 checksums effectively indicate unique content.
+It is theoretically possible for a malicious actor to create a file with different contents but the same checksum, but vanishingly unlikely to occur by chance.
+On requesting new files to download, `neonstore` compares file names and hashes reported by the API to those in the local store, allowing it to omit any data that has already been downloaded, even if the timestamp in a file name has changed but the contents remain the same.
 
-<!--
-In R, we refer to the `duckdb` database as a "remote" connection because the operations are handled outside of R, by `duckdb`.
-Some database connections, like MySQL or Postgres, require the user to install a separate piece of software, often on an external server,
-which we then access over the network connection (with appropriate security credentials), though it can also be run on the same local machine as the 'client' software.
-This client-server approach can be very powerful, but requires more knowledge and resources to deploy.
-In contrast, `duckdb` installs automatically as a stand-alone R package.
--->
+NEON's filenaming convention provided essential metadata about each file, such as the data product (product level, product number, revision number), site, year and month of the release, file description, a timestamp.
+Sensor data (IS) also include additional information about the horizontal and vertical position and meaasurement interval (frequency) of the sensor product.
+AOP products use a somewhat different convention with flight information.
+`neonstore` recognizes all published NEON file naming conventions and automatically parses NEON filenames into their component parts.
+Data are organized into subdirectories based on product, site, and year-month respectively.
+Data or metadata files shared across all months or all sites of a product are stored in the corresponding parent directory.
+The function `neon_index()` returns a `data.frame` listing every recognized individual file in the local store, along with the corresponding metadata fields parsed from the filename.
+Optional arguments can be given to restrict this table to files matching specific products, sites, date ranges, etc.
 

### 2024-01-02 R/neon_cloud.R

@@ -24,7 +24,7 @@ neon_cloud <-function(table,
   urls <- neon_urls(table, product, start_date, end_date, site, type, release,
                     quiet, api, .token = .token)
   
-  format <- gsub(".*(.\\w+)$", "\\1", urls)
+  format <- gsub(".*\\.(\\w+)$", "\\1", urls)
   if(all(format == "csv")) {
     df <- cloud_csv(urls, unify_schemas = unify_schemas)
   } else {
@@ -38,14 +38,16 @@ neon_cloud <-function(table,
 cloud_csv <- function(urls, unify_schemas = FALSE) {   
   # Parse most recent first. Reduces the chance of int/char coercion failures
   # when hitting an all-empty column
-  #meta <- neon_filename_parser(urls)
-  #chrono <- order(meta$GENTIME, decreasing = TRUE)
-  #urls <- urls[chrono]
+  #timestamp <- neon_filename_parser(urls)$GENTIME
+  timestamp <- gsub(paste0(".*", GENTIME, "\\..*"), "\\1", urls)
+  chrono <- order(timestamp, decreasing = TRUE)
+  urls <- urls[chrono]
   ## discover or enforce extension
   ## Detect product type from `meta` and dispatch appropriately.
   
   # https://duckdb.org/docs/data/csv/overview.html
-  # consider all_varchar=1 as possible fallback 
+  # consider all_varchar=1 as possible fallback
+  
   # (possibly avoided by coercion after parsing in chrono order?)
   df <- duckdbfs::open_dataset(urls, format = "csv", filename = TRUE,
                                unify_schemas = unify_schemas)

### 2024-01-02 R/neon_db.R

@@ -73,6 +73,9 @@ neon_db <- function (dir = neon_db_dir(),
     assign("neon_db", db, envir = neonstore_cache)
   }
   
+  e <- globalenv()
+  reg.finalizer(e, function(e) neon_disconnect(db),TRUE)
+  
   db
 }
 

### 2024-01-02 tests/testthat/test-cloud.R

@@ -15,6 +15,13 @@ test_that("Direct cloud access", {
   )
   
   format <- gsub(".*\\.(\\w+)$", "\\1", urls)
+  expect_true(all(format == "csv")) 
+  
+  
+  timestamp <- gsub(paste0(".*", GENTIME, "\\..*"), "\\1", urls)
+  chrono <- order(timestamp, decreasing = TRUE)
+  urls <- urls[chrono]
+  
 })
 
 
@@ -66,6 +73,7 @@ test_that("Big (rate-limited) tests of direct cloud access", {
     
     skip_if_offline()
     skip_on_cran()
+    skip("slow")
     # skip_on_os("windows")
     
   
@@ -73,7 +81,8 @@ test_that("Big (rate-limited) tests of direct cloud access", {
                                 product = "DP1.20288.001",
                                 start_date = "2018-01-01",
                                 end_date = "2023-08-01",
-                                type="basic"
+                                type="basic",
+                                unify_schemas = TRUE
   )
   
   expect_s3_class(df, "tbl_lazy")

### 2024-01-02 R/neon_db.R
testing

@@ -74,7 +74,7 @@ neon_db <- function (dir = neon_db_dir(),
   }
   
   e <- globalenv()
-  reg.finalizer(e, function(e) neon_disconnect(db),TRUE)
+  reg.finalizer(e, function(e) neon_disconnect(),TRUE)
   
   db
 }
@@ -86,7 +86,9 @@ neon_db <- function (dir = neon_db_dir(),
 #' @importFrom DBI dbDisconnect
 neon_disconnect <- function (db = neon_db()) {
   if (inherits(db, "DBIConnection")) {
+    if(DBI::dbIsValid(db)) {
       DBI::dbDisconnect(db, shutdown = TRUE)
+    }
   }
   if (exists("neon_db", envir = neonstore_cache)) {
     suppressWarnings(

### 2024-01-02 R/neon_download.R

@@ -77,8 +77,7 @@
 #' @export
 #' @importFrom R.utils gunzip
 #' @importFrom tools file_path_sans_ext
-#' @examples 
-#' \donttest{
+#' @examplesIf interactive()
 #'  
 #'  ## Omit dir=tempfile() to use persistent storage
 #'  neon_download("DP1.10003.001", 
@@ -95,7 +94,7 @@
 #'                table = "countdata",
 #'                dir = tempfile())
 #' 
-#' }
+#'
 neon_download <- function(product, 
                           table =  NA,
                           site = NA,

### 2024-01-02 R/neon_db.R
finalizer fail

@@ -74,7 +74,7 @@ neon_db <- function (dir = neon_db_dir(),
   }
   
   e <- globalenv()
-  reg.finalizer(e, function(e) neon_disconnect(),TRUE)
+  #reg.finalizer(e, function(e) neon_disconnect(),TRUE)
   
   db
 }

### 2024-01-02 R/neon_db.R
no warning on disconnect

@@ -41,6 +41,7 @@ neon_db <- function (dir = neon_db_dir(),
 
   ## Cannot open read-only on a database that does not exist
   if (!file.exists(dbname) && read_only) {
+    message("initializing database")
     db <- DBI::dbConnect(duckdb::duckdb(), 
                          dbdir = dbname, read_only = FALSE)
     DBI::dbWriteTable(db, "init", data.frame(NEON="NEON"))
@@ -73,8 +74,8 @@ neon_db <- function (dir = neon_db_dir(),
     assign("neon_db", db, envir = neonstore_cache)
   }
   
-  e <- globalenv()
-  #reg.finalizer(e, function(e) neon_disconnect(),TRUE)
+  #e <- globalenv()
+  #reg.finalizer(e, function(e) neon_disconnect(db),TRUE)
   
   db
 }
@@ -85,16 +86,23 @@ neon_db <- function (dir = neon_db_dir(),
 #' @export
 #' @importFrom DBI dbDisconnect
 neon_disconnect <- function (db = neon_db()) {
-  if (inherits(db, "DBIConnection")) {
-    if(DBI::dbIsValid(db)) {
-      DBI::dbDisconnect(db, shutdown = TRUE)
-    }
+  default = getOption("warn")
+  
+  options(warn=-1)
+  
+  if(DBI::dbIsValid(db)) {
+    DBI::dbDisconnect(db, shutdown = TRUE)
   }
+
   if (exists("neon_db", envir = neonstore_cache)) {
     suppressWarnings(
     rm("neon_db", envir = neonstore_cache)
     )
   }
+  
+  gc(FALSE)
+  options(warn=default)
+  invisible(TRUE)
 }
 
 neonstore_cache <- new.env()

### 2024-01-02 R/neon_store.R

@@ -71,7 +71,7 @@ neon_store <- function(table = NA,
     DBI::dbWriteTable(db, "provenance", as.data.frame(index), append = TRUE)
   }
   
-  db <- duckdb_memory_manager(db)
+  #db <- duckdb_memory_manager(db)
   neon_disconnect(db = db)
   invisible(index)
 }

### 2024-01-02 tests/testthat/test-db.R

@@ -6,7 +6,6 @@ context("db")
 
 test_that("neon_db", {
   
-  
   db <- neon_db()
   expect_is(db, "DBIConnection")
 
@@ -14,11 +13,11 @@ test_that("neon_db", {
   
   ## Confirm cached connection
   expect_identical(db, db2)
-  
   neon_disconnect(db)
+  gc()
+
   
   neon_delete_db(ask = FALSE)
-  
 })
 
 

### 2024-01-04 R/neon_cloud.R
warning is confusing

@@ -64,7 +64,6 @@ neon_cloud_ <-function(table,
                     release, quiet, api, .token = .token)
   
   if(length(urls) < 1) {
-    warning(paste("no data matching query"))
     return(NULL)
   }
   

### 2024-01-04 README.Rmd
tidy

@@
 library(neonstore)
+library(dplyr)
+library(stringr)
 Sys.setenv("NEONSTORE_HOME" = tempfile())
@@
 
-```{r}
+```{r, message=FALSE}
 library(neonstore)
+library(tidyverse)
 ```
@@
 products <- neon_products()
-
-i <- grepl("Populations", products$themes)
-products[i, c("productCode", "productName")]
-
-i <- grepl("bird", products$keywords)
-products[i, c("productCode", "productName")]
+products |>
+  filter(str_detect(keywords, "bird")) |> 
+  select(productName, productCode)
 ```
 
-Download all data files in the bird survey data products.
+You may also prefer to explore the [NEON Data Portal](https://data.neonscience.org/) website interactively.
+
+## Download-based workflow
+
+Once we have identified a data product code, we can download all associated data files, e.g.
+in the bird survey data.
+Optionally, we can restrict this download to a set of sites or date ranges of interest, (see function documentation for details).
 
@@
 neon_download("DP1.10003.001")
-
 ```
@@
 
-Once you determine the table of interest, you can
-read in all the component tables into a single `data.frame`
-
-```{r}
-neon_read("brd_countdata-expanded")
-```
-
 ## Database backend
@@
 the raw downloaded files using `neon_store()`:
-
-```{r}
-neon_store(table = "brd_countdata-expanded")
-```
-
-Alternately, we could import all data tables associated with a given product:
 
@@
 
-Access the remote database using `neon_db()`.  This is a `DBIConnection` that
-can easily be used with `dplyr` functions like `tbl()` or `filter()`.  
-Remember that `dplyr` translates these into SQL queries that run directly
-on the database.
+Note that we need to include the product name in the table name when accessing the database, as table names alone may not be unique.
+RStudio users can also list and explore all tables interactively in the Connections pane in RStudio using the function `neon_pane()`.
+
+## Larger-than-RAM data
+
+When working across data from many sites or years simultaneously, it is easy for data to be too big for R to fit into working memory.
+This is especially true when working with sensor data.
+`neonstore` makes it easy to work with such data using dplyr-operations though.
+Just include the option `lazy = TRUE`, and [most dplyr](https://dbplyr.tidyverse.org/reference/index.html) operations will execute quickly on disk instead (by leveraging the `dbplyr` backend and the power of the `duckdb` database).
 
 ```{r}
-library(dplyr)
-con <- neon_db()
-brd <- tbl(con, "brd_countdata-basic-DP1.10003.001")
-brd %>% filter(siteID == "ORNL")
-
+brd <- neon_table("brd_countdata", lazy=TRUE)
+# unique species per site?
+brd |> 
+  distinct(siteID, scientificName) |> 
+  count(siteID, sort=TRUE) |> 
+  collect()
 ```
 
-Note that we need to include the product name in the table name when accessing the database, as table names alone may not be unique.
-RStudio users can also list and explore all tables interactively in the Connections pane in RStudio using `neon_pane()`.
+Use the function `collect()` at the end of a chain of dplyr functions to bring the resulting data into R.
+
+## NEW: Cloud-based workflow
+
+It is now possible to access data directly from NEON's cloud storage system without downloading.
+(Note: this still must ping the NEON API to obtain the most recent list of files, and this list is subject to rate limits).
+Like the local database approach, this strategy works for larger-than-RAM data, and can be substantially faster than downloading.
+However, if you work frequently with the same data products and have ample disk space available, you will find the one-time wait for downloading to be faster.
+
+```{r}
+brd <- neon_cloud("brd_countdata", product="DP1.10003.001")
+
+brd |> 
+  distinct(siteID, scientificName) |> 
+  count(siteID, sort=TRUE) |> 
+  collect()
+```
 
@@
 tables, other products, such as the `.laz` and `.tif` images produced by the
-airborne observation platform LIDAR and cameras may require a different approach.
-
-```{r}
-# Read in a large file list for illustration purposes
-cper_data <- readr::read_csv("https://minio.thelio.carlboettiger.info/shared-data/neon_data_catalog.csv.gz")
-
-## Typically one would read all files in local store, e.g. list.file(neon_dir())
-df <- neon_filename_parser(cper_data$name)
-```
-
-```{r}
-library(dplyr)
-df %>% count(EXT, sort=TRUE)
-```
-
-We can take a look at all `laz` LIDAR files:
-
-```{r}
-df %>% 
-  filter(EXT == "laz")
-```
-
-Note that many of the airborne observation platform (AOP) products, such as
-these LIDAR files, do not include the PRNUM or REV components that make up part
-of the `productCode`s used in the NEON `product` tables.
+airborne observation platform (AOP) sensors such as LIDAR and cameras still
+require the user to work directly with the downloaded files returned
+by `neon_index()`.  Note that the local database can process Eddy Covariance 
+data (h5 files), but at present this does not work with `neon_cloud()`.

### 2024-01-04 R/neon_db.R

@@ -87,7 +87,6 @@ neon_db <- function (dir = neon_db_dir(),
 #' @importFrom DBI dbDisconnect
 neon_disconnect <- function (db = neon_db()) {
   default = getOption("warn")
-  
   options(warn=-1)
   
   if(DBI::dbIsValid(db)) {
@@ -100,6 +99,15 @@ neon_disconnect <- function (db = neon_db()) {
     )
   }
   
+  gc(FALSE)
+  options(warn=default)
+  invisible(TRUE)
+}
+
+
+silent_gc <- function() {
+  default = getOption("warn")
+  options(warn=-1)
   gc(FALSE)
   options(warn=default)
   invisible(TRUE)
